Four Individually Identified Paired Dopamine Neurons Signal Taste Punishment in Larval Drosophila

Version 1

Abstract

Abstract

Dopaminergic neurons (DANs) carry out multiple tasks in the brain, including the transmission of information related to rewards and punishments across various animal species. They are responsible for evaluating sensory input, storing resultant associations as memory, and continuously updating them based on their relevance and reliability. Accurate comprehension of the dopaminergic system’s operation necessitates an understanding of the specific functions mediated by individual DANs. To this end, our research employs Drosophila larvae, which possess approximately 12,000 neurons in their brains, of which only around 1% (approximately 120) are DANs.

The presynaptic projections to the mushroom body (MB) - a brain region pivotal for associative olfactory learning in insects - are limited to only eight larval dopaminergic neurons. These DANs are further subdivided into two clusters: the primary protocerebral anterior medial cluster (pPAM) comprises four cells, and the dorsolateral 1 cluster (DL1) comprises the remaining four cells. Our findings confirm previous research that demonstrates that the pPAM DANs innervating the MB’s medial lobe encode for a gustatory sugar reward signal. Furthermore, we have identified four DANs in the DL1 cluster - DAN-c1, DAN-d1, DAN-f1, and DAN-g1 - each of which innervates distinct compartments of the MB peduncle, lateral appendix, and vertical lobe. Optogenetic activation of DAN-f1 and DAN-g1 alone suffices to substitute for salt punishment. Furthermore, optogenetic inhibition, calcium imaging results and electron microscopy-based reconstruction of all sensory input circuits to the four DL1 DANs demonstrate that each DAN encodes a different aspect of salt punishment, with DAN-g1 being of central importance.

To summarize, our investigation has revealed the existence of a cellular division of labor among larval DANs concerning the transmission of dopaminergic reward (pPAM cluster) and punishment signals (DL1 cluster). Individual DANs in each cluster encode for distinct but partially overlapping aspects of the teaching signal. The striking resemblance in the organizing principle of larval DANs with that of its adult counterpart and the mammalian basal ganglion suggests that there may be a limited number of efficient neural circuit solutions available to address more complex cognitive challenges in nature.

Introduction

To cope with a constantly changing environment, animals derive benefit from learning behavioral rules that need constant adaptation and updating. To test this ability, researchers often use Pavlovian or classical conditioning experiments (Pavlov 1927), which examine an organism’s capacity to form associations between sensory cues, called conditioned stimuli (CS), and rewards or punishments, referred to as aversive and appetitive unconditioned stimuli (US) or teaching signals (Gerber et al 2009, Glanzman 1995, Heisenberg 2003, Menzel 2022, Waddell 2013, Weber et al 2023b, Widmann et al 2018). Dopaminergic neurons (DANs) and other modulatory neurons mediate teaching signals that allow animals to classify a given CS into positive or negative valence based on past experience (Cognigni et al 2018, Menzel 2001, Schultz 2015, Thum & Gerber 2019, Watabe-Uchida et al 2017). For example, when a CS is paired with simultaneous activation of a specific set of DANs in the PAM cluster, larval or adult Drosophila subsequently perceive the odor as attractive (Burke et al 2012, Eschbach et al 2020, Liu et al 2012, Rohwedder et al 2016, Schroll et al 2006) – a similar effect was seen after optogenetic activation of DANs located in the ventral tegmental area of the mouse brain to induce conditioned place preference (Tsai et al 2009). In contrast, activation of larval DL1 DANs (Eschbach et al 2020, Schroll et al 2006, Weiglein et al 2021), adult PPL1 DANs (Aso et al 2012, Aso & Rubin 2016, Aso et al 2010, Claridge-Chang et al 2009), or DANs in the posterior tail of the stratum of mice were shown to reinforce avoidance learning (Menegas et al 2018). These examples illustrate that the dopaminergic system provides a highly conserved and fundamental neural principle of the brain, which has been retained throughout the process of evolution and executes comparable functions across diverse animal species, including even humans (Frick et al 2022).

Our research focuses on the Drosophila larva, an organism endowed with a relatively simple central nervous system, composing approximately 12,000 neurons (Dumstrei et al 2003, Winding et al 2023). Classical conditioning in larvae can be analyzed by stable behavioral assays (Almeida-Carvalho et al 2017, Apostolopoulou et al 2013, Scherer et al 2003, Weber et al 2023a), in conjunction with a wealth of genetic methods for targeting single molecules and neurons (Dietzl et al 2007, Li et al 2014, Perkins et al 2015, Port et al 2020). In addition, the connectome of the brain was recently reconstructed, which enables the extraction and analysis of related wiring motifs, as well as entire learning and memory circuits (Winding et al 2023). Based on these advances, it was possible to establish a more fundamental cellular and synaptic understanding of the larval memory circuit, the mushroom body (MB) network. The MB is a higher-order parallel fiber system present in numerous invertebrate brains, including hemimetabolous as well as holometabolous insects and their larval stages (Strausfeld et al 1998, Strausfeld et al 2020). Analogous to the vertebrate cerebrum, the MB in insects is responsible for forming and maintaining associations (Heisenberg 2003, Menzel 2001, Menzel 2022, Thum & Gerber 2019, Tomer et al 2010, Waddell 2013). In the first developmental stage (L1), the larval MB comprises approximately 110 intrinsic Kenyon cells (KC) per hemisphere, which sparsely encode for conditioned stimulus (CS) information, and is dominated by olfactory input (Eichler et al 2017, Masuda-Nakagawa et al 2009).

The transmission of olfactory information from the peripheral sensory organ is achieved via only 21 olfactory receptor neurons (ORNs) per body side, which in turn send signals across 35 uni- and multiglomerular projection neurons, ultimately reaching the MB KCs (Berck et al 2016, Fishilevich et al 2005, Kreher et al 2005, Masuda-Nakagawa et al 2009, Ramaekers et al 2005). KCs receive modulatory input from 17 neurons, consisting of eight dopaminergic neurons, four octopaminergic neurons, and five neurons with unknown signaling molecule identity (Eichler et al 2017, Rohwedder et al 2016, Saumweber et al 2018, Selcho et al 2009). The DANs are divided into two clusters based on the location of their soma and related neuronal lineage, namely the the primary protocerebral anterior medial cluster (pPAM) and the dorsolateral 1 cluster (DL1) clusters. Each DAN innervates a specific and non-overlapping site in the MB, which, together with one to three output neurons, defines a distinct compartment (Eichler et al 2017). The larval MB comprises a total of 11 compartments, eight of which are defined by DAN input (Eichler et al 2017, Saumweber et al 2018). The four DANs of the pPAM cluster (DAN-h1, DAN-i1, DAN-j1, and DAN-k1) innervate the four related compartments (h, i, j, and k) of the medial lobe and provide teaching signals for sugar rewards (Eichler et al 2017, Rohwedder et al 2016, Schleyer et al 2020). The four DANs of the DL1 cluster (DAN-c1, DAN-d1, DAN-f1, and DAN-g1) innervate the lower peduncle (c), the lateral appendix (d), the intermediate (f), and the lower vertical lobe (g) (Eichler et al 2017, Saumweber et al 2018). Previous studies have shown that the DL1 cluster is capable of mediating aversive teaching signals (Eschbach et al 2020, Schroll et al 2006, Selcho et al 2009, Weiglein et al 2021). Moreover, larvae can use a variety of natural stimuli as aversive teaching signals such as heat, electric shock, vibration, bitter substances (like quinine or caffeine), and high salt concentrations (low salt concentrations, however, are rewarding) (Aceves-Pina & Quinn 1979, Apostolopoulou et al 2016, Apostolopoulou et al 2014, Eschbach et al 2011, Gerber & Hendel 2006, Khurana et al 2009, Khurana et al 2012, Niewalda et al 2008, Pauls et al 2010, von Essen et al 2011). However, it is currently unknown if and how these natural stimuli are represented by the DL1 cluster and to what extent individual DL1 DANs are involved.

We only know that optogenetic stimulation of nociceptive sensory neurons and their downstream basin neurons has been observed to activate three out of the four DL1 DANs, namely DAN-d1, DAN-f1, and DAN-g1 (Eschbach et al 2020). Furthermore, the simultaneous presentation of odor with optogenetic activation of these three DANs has been found to elicit aversion in larvae (Eschbach et al 2020, Weiglein et al 2021). These findings suggest that the activation of DL1 DANs may induce a punishing effect during training, based on somatosensory perception. But can these four individual DANs also process other sensory modalities? Utilizing electron microscopy-based reconstruction, all first and second order input neurons of the MB, consisting of a complete set of 102 neuron pairs, were examined (Eschbach et al 2020, Winding et al 2023). The analysis identified numerous afferent sensory inputs via 20 pairs of feedforward neurons (FFNs), which can now be integrated with the published wiring diagram of all chemosensory input neurons (Miroschnikow et al 2018). This enables a more comprehensive understanding of the distinct modes of action of the four DL1 DANs in relation to the processing of different sensory modalities.

Results

The larval dopaminergic system is functionally bipartite to convey positive and negative teaching signals

There are eight DANs in the larval brain that innervate the mushroom body that can be anatomically subdivided in four pPAM cluster and four DL1 cluster neurons (Figure 1A, supplemental Figure 1A). Three of the four pPAM cluster DANs can be labeled with the driver line R58E02-Gal4 (supplemental Figure 1A-D) (Rohwedder et al 2016). Pairing an odor presentation with simultaneous optogenetic blue light activation of R58E02 positive DANs using a mutant microbial-type rhodopsin ChR2XXL, which depolarizes neurons artificially (Dawydow et al 2014), establishes an appetitive olfactory memory (supplemental Figure 1E and F) (Rohwedder et al 2016). In line with this finding, it was shown that ablation or inhibition of R58E02 positive DANs strongly reduces appetitive olfactory memory (Rohwedder et al 2016) but has no functional significance for aversive olfactory memory (supplemental Figure 1G and H) (Rohwedder et al 2016). Therefore, it can be concluded that pPAM cluster DANs are not involved in the encoding of aversive teaching signals. How are aversive teaching signals encoded?

The larval dopaminergic system is subdivided in two functionally distinct clusters.

(A) The larval dopaminergic neurons (DANs) can be anatomically subdivided into the primary protocerebral anterior medial (pPAM) and dorsolateral 1 (DL1) cluster based on their cell body position. (B) The DL1 cluster (cell bodies in green and purple) consists of four DANs providing input to the c, d, g, and f compartment of the vertical lobe, peduncle and lateral appendix of the MB (in grey). (C, D) Four DL1 DANs (DAN-c1, DAN-d1, DAN-g1, and DAN-f1) are included in the expression pattern of the TH-GAL4 driver line. However, expression of a UAS-mCD8::GFP reporter via TH-GAL4 labels many more neurons (in green, anti-GFP) throughout the entire CNS (in red and blue, anti-N-cadherin and anti-discs large), in total about 100 neurons. (E) To test whether optogenetic activation of the DL1 DANs is sufficient to substitute for a punishment, we used the TH-GAL4 driver in combination with UAS-ChRXXL. Experimental and control larvae were trained by simultaneously presenting and odor and blue light and thus artificial activation of DL1 DANs, whereas a second odor was presented in darkness. Only larvae of the experimental genotype (p > 0.05), but not of the two genetic controls (both p < 0.05), retrieve an aversive associative olfactory memory. The same result was seen after one or three training trials (E and F, respectively). (G, H) To test for the acute function of the DL1 DANs in aversive associative olfactory memory, we expressed GtACR2 via the TH-GAL4 driver. Acute optogenetic inhibition of synaptic output from DL1 and other DANs reduced odor high salt memory. Experimental larvae raised on supplemented food (0.5 mM all-trans-retinal, ATR) and trained in blue light performed on a lower level than control animals kept on standard food (p < 0.05). A memory impairment was seen after one and three training trials (G and H, respectively). All behavioral data is shown as box-plots. Differences between groups are highlighted by horizontal lines between them. Performance indices different from random distribution are indicated below each box-plot. The sample size of each group (N=15) is given below each box-plot. n.s. p > 0.05; * p < 0.05. Scale bars: in C 50 µm and in D 25 µm.

The driver line TH-Gal4 can be used to label the four DANs of the DL1 cluster (Figure 1A and B), as well as most other DANs, except for a small subset of DANs, such as pPAM DANs and neurons in the subesophageal zone (Figure 1C and D) (Selcho et al 2009). By coupling odor presentation with simultaneous optogenetic activation of TH-Gal4 positive DANs through ChR2XXL, an aversive olfactory memory can be established, which was observed after both one (Figure 1E) and three (Figure 1F) odor-DAN activation pairings. Consistent with these findings, aversive associative memory formed after one or three training cycles of odor high salt pairings were significantly reduced when DANs activity was inhibited optogenetically via GtACR2 (Figure 1G and H), a light-gated channel from the algae Guillardia theta (Mohammad et al 2017). Please note that in these experiments salt has to be added during the test to initiate the recall of the aversive memory (indicated as red test plates, Figure 1) (Gerber & Hendel 2006, Schleyer et al 2015). Previous studies have shown that the larva only recalls aversive odor-high salt memory when tested in the presence of the teaching signal. The prevailing explanation suggests that the larva engages in memory-based search during the test, but only if the memory content can be used in a beneficial way (Schleyer et al 2015).

From our results, we infer a division of labor at the level of DANs toward the MB. On the one hand, pPAM neurons mediate a dopaminergic appetitive teaching signal, and on the other hand, DL1 neurons convey a dopaminergic aversive teaching signal. Nevertheless, the findings related to DL1 DANs necessitate additional validation as the TH-Gal4 driver line used in the experiment labels approximately 120 cells, thus limiting the ability to draw cell-specific conclusions.

Anatomical single-cell analysis of DL1 DANs

In an earlier investigation, we conducted a screening of several thousand Gal4 lines in order to construct a high-resolution atlas of the mushroom body at the cellular level via light microscopy (Figure 2A) (Saumweber et al 2018). A subsequent reconstruction of the wiring diagram of the entire MB at the electron microscopic level encompassed not only individual neurons but all synaptic connections (Eichler et al 2017). Utilizing this anatomical work, we identified a total of 102 Gal4 lines featuring MB extrinsic neuron expression patterns. These lines were subjected to intersectional strategies in order to restrict expression patterns to single neurons (Eschbach et al 2020, Saumweber et al 2018). For the present study, we employed eight split-Gal4 lines chosen from this set, as they contained only one or two DANs of the DL1 cluster.

Anatomical single cell analysis of DL1 DAN specific split-GAL4 driver lines.

(A) The larval MB is organized into 11 compartments: CX calyx; IP and LP intermediate and lower peduncle; LA lateral appendix; UVL, IVL, and LVL upper, intermediate, and lower vertical lobe; SHA, UT, IT, LT shaft as well as upper, intermediate, and lower toe of the medial lobe. Single-letter synonyms of compartment names are given as “a–k”. These letters are used to indicate compartment innervation by the MB input and output neurons (Saumweber et al 2018). DL1 cluster DANs are DAN-c1, DAN-d1, DAN-g1, and DAN-f1 that innervate the respective four different compartments of the MB. (B-J) Individual split-Gal4 driver lines were crossed with the reporter strain UAS-mCD8::GFP;mb247-lexA,lexAop-mRFP. Third instar larval brains were dissected, fixed and mounted to visualize the endogenous expression of the MB reporter (mb247-lexA,lexAop-mRFP shown in magenta) and the respective DAN pattern (GFP shown in green). (B-E) SS02160 (DAN-c1), MB328B (DAN-d1), SS02180 (DAN-f1), SS01716 (DAN-g1) each specifically label a single DL1 DAN (cell bodies are highlighted by white arrowheads). (F–K) Two neurons can be seen in MB065B, SS01702, and MB054B split-Gal4 that express in DAN-c1/DAN-f1, DAN-c1/MBIN-e1, and DAN-f1/DAN-g1. Please note that MB065B shows strong expression DAN-f1 but weaker staining in DAN-c1. (H–K) MB054B showed reliable strong expression in DAN-f1 and DAN-g1. In some brains a third weak cell body was visible right next to the other two DANs (I; GFP channel inverted and shown in black; J; cell body highlighted with gray arrowhead). Due to the low expression level we were not able to identify this cell given that only the g and f compartment of the MB were innervated (H). (K) Analysis of the entire brain via native fluorescence expression of GFP (green) and n-Syb (magenta) did not reveal additional cells for MB054B split-Gal4. Scale bars: (B-J) 20 µm, (K) 50µm.

DAN-c1 was specifically labeled by SS02160 (Figure 2B), DAN-d1 by MB328B and MB143B (Figure 2C and supplement Figure 2A), DAN-f1 by SS02180 and MB145B (Figure 2D and supplement Figure 2B) and DAN-g1 by SS01716 (Figure 2E). In addition, we used the split-Gal4 lines MB065B (Figure 2F), SS01702 (Figure 2G) and MB054B (Figure H-K) that specifically labeled DAN-c1/DAN-f1, DAN-c1/MBIN-e1 or DAN-f1/DAN-g1, respectively. For our evaluation we crossed each split-Gal4 line with the reporter strain UAS-mCD8::GFP;mb247-LexA,lexAop-mRFP (Burke et al 2012) that allowed us to monitor the expression pattern (green) in the background of a MB specific reference staining (magenta). Our results confirm the reported specificity even despite the use of different reporter lines and lacking antibody staining (Eschbach et al 2020), as we analyzed only endogenous GFP expression using a more cautious and faster whole mount staining protocol (see Material and Methods). Having established anatomical validity, we subsequently employed these lines to explore the physiological response of each DAN within the DL1 cluster to sugar and salt stimuli. For the purpose of analyzing DAN-d1 and DAN-f1, we limited our investigation to the split-Gal4 lines MB328B and SS02180 as these lines displayed more robust expression levels.

Physiological single-cell analysis of DL1 DANs

In order to investigate how high salt teaching signals are represented at the level of individual DL1 DANs, we conducted calcium imaging experiments using a microfluidic chip on intact and immobilized larvae, as previously described (Si et al 2019). We expressed GCaMP6m under the control of four different split-Gal4 lines specific to DL1 DANs (SS02160 (DAN-c1), MB328B (DAN-d1), SS02180 (DAN-f1), and SS01716 (DAN-g1)), as well as the pPAM specific R58E02-Gal4, and exposed the larvae to solutions containing 100 mM salt, 1 M salt, and also to 500 mM fructose. Our results showed that while none of the cells were activated by 100 mM salt, DAN-c1, DAN-d1, and DAN-g1 responded to 1 M salt (Figure 3A, B, and D). Conversely, pPAM DANs showed a reduced signal in response to 100mM and 1 M salt (Figure 3E). Both DAN-c1 and pPAM DANs showed a calcium increase in response to 500 mM fructose (Figure 3A and E). Although DAN-f1 did not respond to either salt concentration, a reduction in the calcium signal was seen for 500 mM fructose (Figure 3C). Overall, these results provide support for the functional division of the larval dopaminergic system at the physiological level, where most DL1 neurons are activated by high salt concentrations, with one exception inhibited by fructose. In contrast, pPAM cluster neurons display the opposite response, being inhibited by high salt concentrations and strongly activated by fructose.

Calcium responses of DANs to gustatory stimulation.

Four different split-Gal4 lines and the R58E02 driver line were crossed with UAS-GCaMP6m to express a calcium reporter in DANs. The responses of each DAN towards 100 mM NaCl, 1 M NaCl, and 500 mM fructose was tested in intact larvae using a microfluidic chip-based setup. (A) 1 M NaCl (red) and 500 mM fructose (green, p < 0.05 for both), but not 100 mM NaCl (orange, p > 0.05), induced a calcium response in DAN-c1. (B) DAN-d1 calcium responses were only seen after 1 M NaCl stimulation (red, p < 0.05), but not after 100 mM NaCl (orange, p > 0.05) and 500 mM fructose stimulation (green, p > 0.05). (C) Stimulation with 100 mM (orange, p > 0.05) and 1 M NaCl (red, p > 0.05) did not induce calcium responses in DAN-f1. However, stimulation with 500 mM fructose reduced the obtained calcium signal (green, p < 0.05). (D) 1 M NaCl (red, p < 0.05), but not 100 mM NaCl (orange, p > 0.05) and 500 mM fructose (green, p > 0.05), induced a calcium response in DAN-g1. (E) pPAM DANs calcium responses were only seen after 500 mM fructose stimulation (green, p < 0.05). Stimulation with low (orange) and high salt concentrations (red) did not increase calcium signals, however both reduced pPAM activity (p < 0.05). Each graph shows the mean calcium signal plotted as the relative response strength ΔF/F and the related standard error of the mean on the y axis. The time in seconds is given below each graph on the x axis. The grey box indicates the duration of the stimulus application. The sample size of each group (N=5-7) is given above each row. n.s. p > 0.05; * p < 0.05.

Behavioral single-cell analysis of DL1 DANs

We employed split-Gal4 lines SS02160 (DAN-c1), MB328B (DAN-d1), SS02180 (DAN-f1), and SS01716 (DAN-g1) to express the apoptosis proteins Hid and Reaper, resulting in the ablation of the respective individual DL1 DANs. The ablated animals were subjected to the aversive odor-high salt memory assay, using one (Figure 4A-D) and three training cycles (Figure 4E-H). We utilized two different training protocols to differentiate between the effects on short-term memory (STM) and anesthesia-resistant memory (ARM), as previous studies have demonstrated that larvae exhibit both after one training cycle, whereas only ARM is present after three training cycles (Widmann et al 2016). However, to our surprise, we found that the ablation of individual DL1 DANs had no effect on aversive olfactory memories in any of the eight experiments conducted (Figure 4A-H). For DAN-d1 and DAN-f1 we verified these results by using the additional split-Gal4 lines MB143B and MB145B (supplemental Figure 2). In addition, ablation of DAN-d1, DAN-f1 and DAN-g1 individually did not have any impact on odor-quinine and odor-fructose memories established after one cycle training (supplemental Figures 3 and 4). Moreover, MB input neurons that are not dopaminergic (OAN-a1, a2, MBIN-b1, b2, and OAN-e1) were also found to have no relevant effect (supplemental Figure 5). These results suggest that although DL1 DANs respond to high salt concentrations (Figure 3), they are not individually necessary for mediating the aversive teaching signal. It appears that the aversive teaching signal is distributed among multiple cells, indicating the presence of some redundancy. Alternatively, it is possible that there are developmental compensatory effects for DL1 cluster DANs, allowing the remaining three neurons (or other cells) to take over the function of the ablated one.

Ablation of individual DANs does not impair aversive olfactory memory.

In all panels aversive associative performance indices are shown for tests immediately after odor-high salt classical conditioning. In the upper panels (A-D) larvae are trained once, whereas in the lower panels (E-H) three training cycles were applied. Schematic overviews for both conditioning protocols are shown on the left. The four different DL1 DAN specific split-Gal4 driver strains SS02160, MB328B, SS02180, and SS01716 were crossed to the effector UAS-hid,rpr to induce apoptosis (A-H). Cell-specific ablation of distinct DL1 DANs did not impair odor-high salt memories after either one or three cycle training. In no case were associative performance indices upon DL1 DAN ablation lower than in both genetic controls (A-H, in all experiments at least one or even both control groups are compared to the experimental group p > 0.05). All behavioral data is shown as box-plots. Differences between groups are highlighted by horizontal lines between them. Performance indices different from random distribution are indicated below each box-plot. The sample size of each group (N=15, 20, or 25) is given below each box-plot. n.s. p > 0.05; * p < 0.05.

The combination of DAN-f1 and DAN-g1 is required to establish aversive odor-high salt memories

In order to investigate a possible redundancy among DANs, we proceeded by using split-Gal4 lines MB054B (DAN-f1/DAN-g1), MB065B (DAN-c1/DAN-f1), and SS01702 (DAN-c1/MBIN-e1) to express apoptosis proteins Hid and Reaper. This approach selectively ablated either two DL1 DANs or one DL1 DAN and another MB input neuron with unknown neurotransmitter identity. It should be noted that no additional split-Gal4 lines are presently available for further DL1 DAN combinations or the complete DL1 cluster. We then subjected the larvae to the aversive odor-high salt memory assay, using one (Figure 5A-C) and three (Figure 5D-F) training cycles. Only larvae with ablated DAN-f1/DAN-g1 exhibited significantly reduced aversive olfactory memory after one cycle training (Figure 5A). No memory reduction was observed when ablating DAN-c1/DAN-f1 or DAN-c1/MBIN-e1 (Figure 5B and C). Ablation of all three sets of neurons, including MB054B, had no effect on aversive olfactory memory after three training cycles (Figure 5D-F). To ascertain the independence of the DAN-f1/DAN-g1 neuron-mediated impairment of aversive olfactory memory after one cycle training from the specific odors used, hexyl acetate and benzaldehyde were employed instead of amyl acetate and benzaldehyde. The results of these experiments confirmed the previous findings, as ablation of DAN-f1/DAN-g1 neurons significantly reduced aversive odor-high salt memory after one cycle training (Figure 5G), but not after three cycles (Figure 5J). Innate olfactory and gustatory behavior was not impaired by DAN-f1/DAN-g1 ablation (supplemental Figure 6).

Ablation of DAN-f1 and DAN-g1 together impairs aversive olfactory memory.

In all panels associative performance indices are shown for tests immediately after classical conditioning. In the upper panels (A-C) larvae are trained once by pairing an olfactory stimulus with high salt punishment, whereas in the lower panels (D-F) three training cycles were applied. Schematic overviews for both conditioning protocols are shown on the left. The three different DL1 DAN specific split-Gal4 driver MB054B, MB065B, and SS01702 that each label two neurons were crossed to the effector UAS-hid,rpr to induce apoptosis (A-F). (A) With MB054B used as driver strain to ablate the DL1 DAN combination DAN-f1/DAN-g1, the aversive associative performance index of the experimental group was decreased compared to both controls (p < 0.05). (B-F) In all other experiments ablation of different DL1 DAN combinations did not reveal a phenotype. (in all experiments at least one or even both control groups are compared to the experimental group p > 0.05). Please note, that this also includes MB054B crossed with UAS-hid,rpr tested after three training trials (D). (G) To verify the memory phenotype of MB054B crossed with UAS-hid,rpr tested after one trail conditioning, we repeated the experiment using the odor pair hexyl acetate (HA) and benzaldehyde (BA). Again, experimental larvae tested after one trial learning showed a robust decrease in odor-high salt memory when compared to both genetic control groups (p < 0.05). The memory phenotype was not seen after three training trials (J, p > 0.05). (H, K) With MB054B used as driver strain to ablate the DL1 DAN combination DAN-f1/DAN-g1, aversive odor-quinine memory was not impaired after one or three cycle conditioning (p > 0.05 when comparing experimental and control groups). Similarly, appetitive odor-fructose learning after one and three cycle conditioning was not impaired when ablating DAN-f1/DAN-g1 (p > 0.05 when comparing experimental and control groups). All behavioral data is shown as box-plots. Differences between groups are highlighted by horizontal lines between them. Performance indices different from random distribution are indicated below each box-plot. The sample size of each group (N=15, 20, or 25) is given below each box-plot. n.s. p > 0.05; * p < 0.05.

Subsequently, we investigated the selectivity of DAN-f1/DAN-g1 neurons concerning the nature of the aversive teaching stimulus. To this end, we utilized quinine as an aversive teaching signal, a stimulus that has been demonstrated to be learnable by the larva, as evidenced by previous studies (Apostolopoulou et al 2014, El-Keredy et al 2012). Removal of the DAN-f1/DAN-g1 neurons using the split-Gal4 line MB054B had no effect on aversive odor-quinine memory independent of the number of training cycles (Figure 5H and K). Furthermore, ablation of DAN-f1/DAN-g1 did not alter appetitive odor-fructose memories after one or three training cycles (Figure 5I and L). Therefore, the DL1 DAN-f1/DAN-g1 combination seems to be specifically required for aversive odor-high salt memory.

Individual DL1 DANs can instruct an aversive high salt memory

In order to determine if the activation of individual DL1 DANs signals the natural high salt punishment, we conducted a one training trial learning experiment using ChR2XXL to artificially activate DAN-c1, DAN-d1, DAN-f1, DAN-g1, and the combination of DAN-f1/DAN-g1. We used split-Gal4 lines SS02160, MB328B, SS02180, SS01716, and MB054B, respectively (Figure 6). Our approach suggests that an aversive olfactory memory induced by optogenetic activation of a single DL1-DAN, when recalled on a salt plate, corresponds to a natural aversive odor-high salt memory.

Optogenetic DL1 DAN activity can substitute for salt punishment.

In all panels associative performance indices are shown for tests immediately after classical conditioning. In panels (A-E) larvae are trained once by pairing an olfactory stimulus with artificial blue light activation, whereas in panel (F) three training cycles were applied. Schematic overviews for both conditioning protocols are shown to the left of (A) and (F). (A-D) To test whether optogenetic activation of the individual DL DANs DAN-c1, DAN-d1, DAN-f1, and DAN-g1 is sufficient to substitute for a punishment, we used the split-Gal4 lines SS02160, MB328B, SS02180, and SS01716 in combination with UAS-ChR2XXL. (E, F) For simultaneous optogenetic activation of DAN-f1/DAN-g1 we used MB054B. Larvae of the experimental genotypes for DAN-c1, DAN-f1, DAN-g1 and the DAN-f1/DAN-g1 combination (for all p < 0.05), but not for DAN-d1 and all genetic controls (for all p > 0.05), showed an aversive associative memory. The results imply that in the tested conditions, the punishment signal can be mediated by the artificial activation of all individual DL1 DANs, with the exception of DAN-d1. All behavioral data is shown as box-plots. Differences between groups are highlighted by horizontal lines between them. Performance indices different from random distribution are indicated below each box-plot. The sample size of each group (N=15) is given below each box-plot. n.s. p > 0.05; * p < 0.05.

Simultaneously activating DAN-f1/DAN-g1 positive DANs via ChR2XXL during an odor presentation establishes an aversive olfactory memory (Figure 6E and F). These results are comparable to those observed for TH-Gal4 (as shown in Figure 1E and F). Additionally, this memory is observed after one (Figure 6E) and three training cycles (Figure 6F). Furthermore, single-cell analysis indicates that individual activation of DAN-f1 and DAN-g1, but not of DAN-d1 and DAN-c1 (please note that experimental animals are different from zero and significantly different to one control), can induce an aversive odor-high salt memory (Figure 6A-D). In summary, these results suggest that DAN-f1 and DAN-g1 encode aspects of the natural aversive high salt teaching signal under the conditions that we tested.

Individual DL1 DANs are acutely necessary to establish aversive odor-high salt memories

Next, we examined the extent to which the individual DL1 DANs are acutely required for the transmission of a functional aversive teaching signal. We have already shown that the combination of DAN-f1/DAN-g1 is required for aversive odor-high salt memory (Figure 5A and G). However, in these experiments DANs were removed during the entire development, which may have led to compensatory mechanisms or side effects.

To overcome this limitation, we left the individual DL1 DANs intact during embryonic and larval development and used GtACR2 to selectively inhibit DAN activity only during training, thus isolating the teaching signals from other DAN functions. Optogenetic inhibition of DAN-f1/DAN-g1 resulted in a reduced aversive odor-high salt memory. This was the case after one training cycle (Figure 7E) and three training cycles (Figure 7F). Thus, the learning deficit in this experiment was more pronounced than after ablation of these two DL1 DANs throughout development (Figure 5D). Optogenetic inhibition of individual DL1 DANs showed no effect on aversive odor-high salt memories (Figure 7A-C), except for DAN-g1 that led to a small but significant reduction (Figure 7D).

Optogenetic inhibition of DL1 DAN activity impairs aversive olfactory memory.

In all panels associative performance indices are shown for tests immediately after classical odor high salt conditioning. In panels (A-E) larvae are trained once by pairing an olfactory stimulus with an aversive high salt stimulation, whereas in panel (F) three training cycles were applied. Schematic overviews for both conditioning protocols are shown to the left of (A) and (F). (A-D) To test whether optogenetic inhibition of the individual DL DANs DAN-c1, DAN-d1, DAN-f1, and DAN-g1 during training impairs punishment signaling, we used the split-Gal4 lines SS02160, MB328B, SS02180, and SS01716 in combination with UAS-GtACR2 and blue light stimulation during the entire training phase. (E, F) For simultaneous optogenetic inhibition of DAN-f1/DAN-g1 we used MB054B. (A-C) Larvae with inhibited DAN-c1, DAN-d1, or DAN-f1 function during training showed no impairment of odor-high salt memory comparable to controls that were kept on standard food without supplemented all-trans-retinal (ATR, 0.5 mM) and received the same protocol (p < 0.05) (for all p > 0.05). (D-F) In contrast inhibition of DAN-g1 alone, or the combination of DAN-f1/DAN-g1 after single trial and three trial conditioning impaired odor-high salt memory compared to controls (for all p < 0.05). This shows that DAN-g1 function is of central importance for signaling a salt punishment teaching signal. All behavioral data is shown as box-plots. Differences between groups are highlighted by horizontal lines between them. Performance indices different from random distribution are indicated below each box-plot. The sample size of each group (N=15) is given below each box-plot. n.s. p > 0.05; * p < 0.05.

A synaptic reconstruction of all chemosensory inputs for each individual DL1 and pPAM DAN.

Given the apparent cellular division of labor of DANs for reward and punishment teaching signals through the pPAM and DL1 cluster (Figure 1 and supplement Figure 1) and even qualitative differences between individual DL1 DANs (Figure 3-7), since not all DANs code equally for the punishment teaching signal, the question arises as to how this may result from the respective cellular and synaptic connectivity. Therefore, we compared the input wiring diagrams of the pPAM and DL1 clusters and their individual DANs by mapping the synaptic connections from sensory neurons - to interneurons - to individual DANs, using the full larval brain EM volume (Figure 8, supplemental Figure 7) (Eschbach et al 2020, Miroschnikow et al 2018, Winding et al 2023). We found that DAN-c1, DAN-d1, DAN-f1, and DAN-g1 (black circle) get input from 1, 6, 6 or 11 sensory-to-DAN interneurons (orange circle), respectively (Figure 8A). The number of sensory neurons (numbers in colored circuits below) varies between 3, 42, 35 and 9. There is no direct input from sensory neurons onto DL1 DANs. DAN-c1 receives only very weak internal and gustatory input, while DAN-d1 and DAN-g1 have an overall stronger and additional input from somatosensory neurons. In addition, DAN-d1, as well as DAN-f1, receive information from olfactory and thermosensory neurons (Figure 8D). DAN-f1 and DAN-g1 get 17% and 11% of their entire input from these sensory-to-DAN interneurons. For DAN-d1 and DAN-c1 these number were lower (5% and 1%, respectively). In general, DL1 DANs get additional input from other upstream neurons (US) that are not directly linked to sensory neurons (grey circles; between 17% and 37%), as well as MB KCs (purple circles; between 36% and 49%) as well as a bit of MBON feedback (dark-red circle; between 1% and 2%). Please note that we restricted our analysis on direct sensory-to-DANs interneurons (single hop analysis) and did not expand the analysis onto two hop or three hop connections to keep the evaluation restricted to the shortest paths from sensory neurons to DANs. For the pPAM DANs we found the same general connectivity principles as for DL1 DANs regarding input from MBONs, KCs, and non-sensory upstream neurons. Please note that the DAN-h1 neuron is not present at the early L1 EM brain volume and thus cannot be included in this evaluation (Figure 8B). In contrast, however, it is striking that pPAM DANs seem to get input from more interneurons connected to the sensory system (up to 101 for DAN-j1) signaling additional internal somatosensory and CO2 information. A closer inspection of the sensory-to-DAN interneurons showed that 27 of the 35 neurons are specifically innervating only DL1 or pPAM DANs (Figure 8C). Only 8 out of 35 interneurons connect to both clusters. Based on a calculated hub score sensory-to-DAN interneurons pair 1 (FB2IN-12) and pair 5 (FFN-21) were likely most significant for instructing DL1 DANs with sensory information. Whereas pairs 10 (FFN-17) and 11 (FFN-12) were – based on connectivity – most important in instructing DL1 and pPAM DANs. For the three pPAM DANs sensory-to-DAN interneuron pairs 16, 17, 18, 19 and 20 resulted in the highest hub scores (FFN-37, FFN-4, FFN-5, FFN-18 and FB-4). The difference in cellular connectivity of DL1 and pPAM DANs is also reflected in different local input brain hubs: axo-dendritic synapses of sensory-to-DAN interneurons with DL1 DANs formed a characteristic hook-shape in the superior protocerebrum; in contrast, pPAM DANs get synaptic input from sensory-to-DAN interneurons in the anterior inferior protocerebrum (Figure 8F).

Interneurons and hub analysis of sensory to DAN pathways.

(A, B) Schematic graph representation of individual DL1 (A) and pPAM (B) DANs. The outer ring at the bottom of each scheme represents the sensory composition of neurons targeting sensory-to-DANs interneurons (orange circle). The type of sensory information is encoded by the respective color (ACa: anterior central sensory compartment, AVa: ventral anterior sensory compartment, ACal: lateral anterior central sensory compartment, ACp: posterior anterior central sensory compartment, ACpl: posterior-lateral anterior central sensory compartment, VM: ventromedical sensory compartment, TD CO2: tracheal dendritic neurons responding to CO2, ORNs: olfactory receptor neurons). DAN input neurons are shown that get no direct sensory input (grey circles). Individual DANs are shown in the middle of the scheme as black circles. They are connected to MB KC (purple circles), which in turn connect to mushroom body output neuron (MBONs; dark red circles). Arrows indicate the direction of the synaptic connection and its strength (coded by arrow thickness). Numbers in circles indicate number of neurons. The percentages indicate the proportion relative to the total input that the cell receives from the specific neuronal partners. For example, DAN-f1 receives 17% of its input from six different sensory-to-DAN interneurons (yellow circle), 17% from 19 other, non-sensory interneurons, 41% from 64 MB KCs, and 2% from two MBONs. (C) Dot plot showing the importance of interneurons acting as sensory to DAN hub. Dot size was calculated using the fraction of total input an interneuron receives from sensory neurons multiplied by the fraction of total input this interneuron gives to an DAN. Colored backgrounds of dots are highlighted in orange for the connections with a hub size of 0.001 or above. (D, E) Schematic of graph representation. The outer ring represents the sensory composition of neurons targeting upstream neurons of DANs. The type of sensory information is encoded by the respective color. Synaptic threshold for upstream interneurons of DANs = 3 and of upstream sensory neurons = 1. Line thickness to interneurons and targets represents the percentage of synaptic input. White or orange circles connected to the outer ring represent the interneuron layer. The inner ring represents individual target neurons, DL1 and pPAM DANs. The identity of each DAN and interneuron is given by the label in its related circle. (F) EM reconstruction of DL1 and pPAM DANs (grey) highlighting their presynaptic sites in red (for DL1 DANs) and green (for pPAM DANs). At the top a horizontal view of the brain is shown. At the bottom a frontal view of the brain is shown.

Discussion

A functional assignment for individual DANs of the DL1 cluster in their ability to encode a salt punishment teaching signal

The discovery of DL1 neurons as aversive teaching signal mediators in larval Drosophila complements previous findings that demonstrate the sufficiency of four DANs from the pPAM cluster as an appetitive reinforcement signal in these animals (Rohwedder et al 2016). This dichotomy demonstrates an organizational principle shared by adult Drosophila and larvae (Aso et al 2012, Aso & Rubin 2016, Aso et al 2010, Burke et al 2012, Claridge-Chang et al 2009, Das et al 2014, Huetteroth et al 2015, Liu et al 2012, Schleyer et al 2020, Schroll et al 2006, Selcho et al 2009, Weiglein et al 2021, Yamagata et al 2015), despite a significant reduction in cell numbers. The principle appears to apply to mammals as well (Groessl et al 2018, Lammel et al 2012, Menegas et al 2018, Schultz 2015). But how do individual neurons contribute to the teaching signal of the complete cluster? Is the signal evenly distributed across multiple cells, supporting the mass action hypothesis, or do individual DANs specifically encode for the salt stimulus, as posited by the labeled line hypothesis? Our results show that the four DL1 DANs perform this function in a mixture of both forms. There is no unique labeled line because both ablation of each individual DAN (Figure 4) and the acute optogenetic inhibition (Figure 7) do not abolish the odor-high salt memory. The teaching signal is combinatorically encoded within the DL1 cluster, yet its distribution among the four DANs is non-uniform and relies heavily on the functionalities of DAN-g1 and DAN-f1, (Table 1, Figure 5-7). Each DAN possesses a distinct identity, which is defined by its specific, albeit partially overlapping, neuronal input circuitry and physiological response to salt (Table 1, Figure 3 and 8).

We speculate that DAN-c1 activity does not encode a taste-specific teaching signal. It rather transmits preprocessed information from the higher brain to build a more integrated aversive memory, such as a general alert memory or a state-dependent memory as this particular cell is only weakly connected to three sensory neurons via a single sensor-to-DAN interneuron and is activated by both positive and negative tastes. DAN-d1 activity appears to have a negligible impact on the formation of odor-high salt memories in the paradigms we tested. Even though the neuron responds to high salt concentrations and receives input from 42 sensory neurons via six sensor-to-DAN interneurons, its activity did not induce an aversive memory and was not found to be essential for odor-high salt memory. Nonetheless, it is possible that DAN-d1 could instruct other salt-dependent memory types, established after non-associative training, second-order conditioning, or reversal learning (Kacsoh et al 2015, Konig et al 2019, Mancini et al 2019, Paranjpe et al 2012, Tabone & de Belle 2011, Yamada et al 2023). Indeed, other studies involving optogenetic training using CsChrimson and one-odor training suggest that DAN-d1 activation can even instruct aversive olfactory memory in larvae (Eschbach et al 2020, Weiglein et al 2021). In combination, DAN-f1 activity and DAN-g1 seem to have a crucial function in high salt punishment. DAN-f1 is inhibited by sugar and obtains input from 35 sensory neurons through six sensor-to-DAN interneurons. Despite persistent and current suppression of activity having no impact on the retention of a memory related to a high salt stimulus, DAN-f1 activity can instruct the formation of such a memory. This discovery is consistent with previous research, although their experimental designs differ from our own in the test situation (Eschbach et al 2020, Weiglein et al 2021). Consequently, it is plausible that DAN-f1 activity encodes for a teaching signal that carries additional information beyond high salt. It is evident that DAN-g1 exerts the most potent effect. Following exposure to high salt, this cell becomes active, is crucial for the formation of an odor-high salt memory and can even establish it. Additionally, it receives extensive sensory input from nine neurons through 11 distinct interneurons. Therefore, DAN-g1 is the only cell of the DL1 cluster that satisfies all prerequisites to independently encode the high salt teaching signal. In summary, the DL1 cluster appears to be arranged in a judicious functional configuration. The instructive signal is encoded by a mere four cells, with the requisite redundancy to preclude loss of function in the event of single-cell failure. Nevertheless, the cells manifest variation, thereby covering different and cumulative characteristics of stimulation.

The functioning of the DL1 cluster DANs as a multimodal node for punishment

Larvae, of course, perceive not only high salt concentrations as punishing, but also other sensory stimuli such as bitter substances (quinines and caffeine), electric shock, temperature, mechanosensory input (vibration), and light (Aceves-Pina & Quinn 1979, Apostolopoulou et al 2016, Apostolopoulou et al 2014, Eschbach et al 2011, Gerber & Hendel 2006, Khurana et al 2012, Pauls et al 2010, von Essen et al 2011). However, the extent to which the DL1 cluster is involved in coding for most of these aversive teaching signals remains unclear. Currently, comparisons at the level of DANs can only be made for sensory inputs that are dependent on high salt or mechanosensation. Interestingly, similar to high salt concentrations, somatosensory information can also result in the establishment of aversive olfactory memory in larvae (Eschbach et al 2020). The transmission of these somatosensory teaching signals is orchestrated by mechanosensory neurons from the chordotonal organs, class IV multidendritic nociceptive neurons distributed along the body, and multisensory basin neurons (Hwang et al 2007, Jovanic et al 2016, Ohyama et al 2015, Tracey et al 2003). The connectome revealed that these mechanosensory neurons also link to the DL1 cluster through three to four interneurons and elicit the activation of DAN-d1, DAN-f1, and DAN-g1 (Eschbach et al 2020). Upon comparing the interneurons that relay high salt and mechanosensory information to the DL1 cluster, six pairs of interneurons are of critical significance, respectively. These pairs include FFN-20, FB2N-12, FB2N-19, FFN-23, FFN-29, and FB2N-15 for mechanosensory information (Eschbach et al 2020), and FB2N-12, FB2N-18, FB2IN-11, FB2IN-6, FFN-21, and FFN-24 for high salt concentrations (Figure 8; supplemental Figure 7, rightmost column). Interestingly, only one pair of interneurons - FB2N12 - is shared between the two types of information and the other five types are specific to the sensory modality. The wiring patterns of DAN input neurons exhibit a notable consistency, with a limited number of neurons dedicated to a specific sensory modality, and single neurons shared across different modalities. This results in a unique combination of cellular codes that are partly redundant and partly specific for the four DL1 DANs. While certain DANs may hold greater significance for a particular information, such as DAN-d1 for somatosensory information, the DAN-g1 cell appears to play a general central role for encoding an aversive teaching signal. DAN-g1 responds to all aversive stimuli tested thus far and its activity has proven informative for all aversive memories analyzed to date. In order to gain a deeper comprehension of the processing of diverse sensory modalities in the larval brain, it is crucial to broaden the analytical framework presented in our work to encompass visual stimuli, various bitter substances, and temperature stimuli.

However, the overall depiction presented above of the information flow and processing of DL1 DANs is an oversimplification. Indeed, DL1 and pPAM DANs are among the most complex and highly recurrent neurons in the brain (Winding et al 2023). This exceptional level of recurrent connectivity enables DANs to provide high-dimensional feedback, which enables them to encode a diverse range of features and engage in parallel computations. Such computations can guide a working memory, contain feedback from neurons that integrate both learned and innate values, and receive long-range feedback from descending neurons that encode motor commands; all in addition to polysynaptic feedforward inputs from the entire set of sensory modalities.

A cellular comparison of DAN function across Drosophila metamorphosis

A recent study conducted by Truman and colleagues explored the integration, or lack thereof, of individual input and output neurons from the larval MB into the adult network following metamorphosis (Truman et al 2023). The study demonstrated that while all four DANs of the pPAM cluster perished, three out of the four cells from the DL1 cluster persisted as part of the adult MB circuitry. Only DAN-f1 altered its innervation in the adult brain, relocating from the larval MB to the adult superior medial protocerebrum. In contrast, DAN-c1, DAN-d1, and DAN-g1 maintained their function as MB DANs in the adult system, identified as PPL1-γ1pedc, PPL1-γ2α’1, and PPL1-γ1, respectively.

Due to the lack of appropriate genetic tools, data on the function of PPL1-γ1 is scarce (Aso et al 2014). Therefore, a functional comparison between larval and adult DANs is only possible for the DAN-c1/PPL1-γ1pedc and DAN-d1/PPL1-γ2α’1 combination. Intriguingly, the adult PPL1 cluster has a bipartite functional organization (Schnitzer et al 2023, Vrontou et al 2021). Embryonic-born DAN-c1/PPL1-γ1pedc, DAN-d1/PPL1-γ2α’1, and DAN-g1/PPL1-γ1 exhibit sustained activity and respond to various sensory stimuli, including electric shock, temperature, or bitter taste, that can instruct short lasting aversive memories (Aso et al 2012, Aso & Rubin 2016, Aso et al 2010, Claridge-Chang et al 2009, Das et al 2014, Galili et al 2014, Kirkhart & Scott 2015, Schnitzer et al 2023, Tomchik 2013, Villar et al 2022, Vrontou et al 2021). Conversely, rewarding stimuli such as sugar water inhibit the spiking of these DANs (Schnitzer et al 2023), possibly with involvement of NPF (Krashes et al 2009). The remaining three PPL1-DANs (PPL1-α′3, PPL1-α′2α2, PPL1-α3), which do not have a larval history, exhibit more sporadic firing and do not respond as strongly to environmental stimuli (Vrontou et al 2021). With repeated conditioning cycles, these slower-acting PPL1-DANs become less inhibited by feedback from the short-term learning units DAN-c1/PPL1-γ1pedc and DAN-d1/PPL1-γ2α’1, enabling them to respond more strongly to sensory input (Schnitzer et al 2023). Thus, the MB circuit in adult flies features a feedback loop involving embryonic-born DAN-c1/PPL1-γ1pedc and DAN-d1/PPL1-γ2α’1, which connect to output neurons that in turn communicate with later-born PPL1-α′3, PPL1-α′2α2, and PPL1-α3 DANs. This loop serves to facilitate short-term memory formation and regulate the induction of long-term memory traces, ensuring that only repeated and reliable associations are retained (Schnitzer et al 2023). Despite the massive reorganization of the mushroom body, the basic function of DAN-c1/PPL1-γ1pedc and DAN-d1/PPL1-γ2α’1, mediating punishment teaching signals that instruct short-term memories, remains the same. Evidence indicates that the underlying molecular signaling pathways maintain their identity and kinetics throughout development, with larval and adult interstimulus interval curves for DAN-d1/PPL1-γ2α’1 being almost identical (Aso & Rubin 2016, Schnitzer et al 2023, Weiglein et al 2021).

During adulthood, PPL1-DANs are responsible for numerous other functions, including innate olfactory responses, suppression of appetite memory in response to nutritional status, pre-exposure learning, memory reconsolidation, and forgetting (Berry et al 2015, Berry et al 2018, Felsenberg et al 2017, Jacob et al 2021, Krashes et al 2009, McCurdy et al 2021, Placais et al 2012, Siju et al 2020, Tian et al 2016, Vrontou et al 2021). Additionally, they are involved in various behaviors like locomotion and sleep (Berry et al 2015). An open question is whether larval DL1 DANs encode these same functions to a similar extent. Our study establishes a fundamental framework that can be leveraged to investigate this issue. Considering the numerical simplicity of the larval nervous system, the available genetic tools, and the complete connectome, it may be possible to attain an understanding of the larval dopaminergic system at the level of single cells and single synapses. It is plausible that nature has a limited number of effective neural circuit solutions for complex cognitive and behavioral problems, such as the remarkable ability of brains to learn from a small number of examples. Therefore, comprehending the larval dopaminergic system may provide insight into the shared circuit motifs that exist across the animal kingdom.

Materials and Methods

Fly Strains

Flies were raised and maintained on Drosophila standard food at 25°C, 60-80% relative humidity and a 14/7-hour light/dark cycle. For anatomical analysis, UAS-mCD8::GFP (Bloomington stock center no. 32194) (Selcho et al 2009), UAS-mCD8::GFP;mb247-LexA,lexAop-mRFP/TM3,Sb (Burke et al 2012) and UAS-mCD8::GFP;nSyb-LexA,lexAop-mRFP/TM6B (Burke et al 2012, Riabinina et al 2015) were used to analyze the morphology of DANs. For behavioral experiments, UAS-hid,rpr (Abbott & Lengyel 1991, White et al 1996), UAS-GtACR2 (Mohammad et al 2017) (Bloomington stock center no. 92984) and UAS-ChR2XXL (Dawydow et al 2014) (Bloomington stock center no. 58374) were used to ablate or optogenetically silence or activate neurons by blue light (470 nm). UAS-GCaMP6m (Chen et al 2013) (Bloomington stock center no. 42748) flies were used for Ca2+-imaging experiments. The Gal4 strains TH-Gal4 (Friggi-Grelin et al 2003) and R58E02-Gal4 (Rohwedder et al 2016) (Bloomington stock center no. 41347) were used to manipulate different sets of DANs. Split-Gal4 lines MB054B, MB065B, MB328B, MB143B, MB145B, SS02160, SS02180, SS01716, SS01958, SS21716, SS24765, SS01702 (Eschbach et al 2020, Pfeiffer et al 2010, Saumweber et al 2018) were used for anatomical, physiological and behavioral experiments. w1118 flies (Bloomington stock center no. 3605) were used to obtain heterozygous driver and effector control groups. See supplementary table 1 for more information.

Anatomical Analysis

Immunostaining

To confirm expression patterns of R58E02-Gal4 and TH-Gal4, flies were crossed to UAS-mCD8::GFP and raised at 25°C for seven days before dissection. Third instar larvae (wandering stage) were put on ice and dissected in PBS (Phosphate Buffered Saline; Sigma Aldrich, cat. no. P4417). Brains were fixed in 4% formaldehyde solution (in PBS, Thermo Scientific, cat. no. 047392.9M) for 20 minutes at room temperature. After eight rinses in PBT (PBS with 3% Triton X-100, Sigma Aldrich, cat. no. X100), brains were blocked with 5% normal goat serum (NGS, Sigma Aldrich, cat. no. G9023) in PBT for 1 hour at room temperature and incubated for 48 hours with primary antibodies at 4°C. Before application of the secondary antibodies for at least 24 hours at 4°C, brains were washed seven times with PBT. After that, larval brains were again washed eight times with PBT and mounted on poly-L-lysin-coated coverslips (Janelia FlyLight DPX mounting receipt: https://www.janelia.org/project-team/flylight/protocols), dehydrated through a series of increasing concentrations of ethanol and cleared three times for 5 minutes in xylene (Sigma Aldrich, cat. no. 247642). At the end, larval brains were mounted in DPX mounting medium (dibutyl phthalate in xylene, Sigma Aldrich, cat. no. 06522) and left to rest in the dark for at least 24 hours before imaging. Primary antibodies were: rat anti-N-Cadherin (1:50; Hybridoma, cat. no. DN-Ex #8-s), rabbit anti-GFP (1:1000; Life Technologies, cat. no. A6455) and mouse 4F3 anti-DLG (1:200; Hybridoma, cat. no. 4F3 anti-discs large). Secondary antibodies were: goat anti-rat Alexa Fluor 647 (1:500, for anti-N-Cadherin, Life Technologies, cat. no. A21247), goat anti-rabbit Alexa Fluor 488 (1:500, for anti-GFP, Life Technologies, cat. no. A11008) and goat anti-mouse Alexa Fluor 568 (1:500, for anti-DLG, Life Technologies, cat. no. A10037).

Native fluorescence

For native fluorescence, we used a previously described protocol for adult flies (Pitman et al 2011) and adapted it to larvae. All split-Gal4 lines were crossed to UAS-mCD8::GFP;mb247-LexA,lexAop-mRFP/TM3,Sb, while split-Gal4 line MB054B was additionally crossed to UAS-mCD8::GFP;nSyb-LexA,lexAop-mRFP/TM6B and raised at 25°C for seven days. Third instar larvae (wandering stage) were put on ice and dissected in PBS. Brains were fixed under vacuum in 4% formaldehyde solution (in PBS) for 25 minutes at room temperature. After five washes in PBT (PBS with 3% Triton X-100), larval brains were rinsed twice directly and once for 10 minutes under vacuum in PBS at room temperature. Samples were mounted on poly-L-lysin-coated coverslips and embedded in Vectashield (Vector Laboratories, cat. no. H-1000-10). Slides were stored in the dark and cold for at least 24 hours before imaging.

Confocal microscopy

Confocal microscopy was conducted on a Zeiss LSM800 confocal laser scanning microscope with ZEN 2.3 software. All images were projected and adjusted with ImageJ (Fiji is just ImageJ, Version 1.53c, Java 1.8.0_172 (64-bit)). Final Figures were arranged and labeled in Affinity Publisher 2.1.1.

Functional Imaging and microfluidics

For larval imaging experiments, adult flies were transferred to larvae collection cages (Genesee Scientific) containing grape juice agar plates and 180 mg of fresh yeast paste per cage. Flies were allowed to lay eggs on the agar plate for 1–2 days before the plate was removed for collection of larvae. Calcium imaging experiments were performed in L1 larvae (2 days AEL). We used a previously described method for microfluidic delivery of chemicals in aqueous form with simultaneous imaging of calcium activity in intact larvae (Si et al 2019). All experiments used an 8-channel microfluidic chip equipped with a vacuum port to stabilize the animal’s head. The same tastants were used as in the behavioral experiments (NaCl: VWR Chemicals, cat. no. 27810.364; D-Fructose: Sigma Aldrich cat. no. 47740). Stimuli consisted of 5-10 s pulses interleaved with 15 s water washout periods. An L1 larva was washed with deionized water and loaded into the microfluidic device using a 1 mL syringe filled with Triton X-100 (0.1% [v/v]) solution. The animal was pushed to the end of the loading channel with its dorsal side facing the objective. GCaMP6m signal was recorded using an inverted Nikon Ti-E spinning disk confocal microscope and a 60X water immersion objective (NA 1.2). A CCD microscope camera (Andor iXon EMCCD) captured frames at 30 Hz. Recordings from at least 5–7 larvae were collected for each condition. For each larva the average response over 5s around the peak value during stimulus presentation was compared to chance level analyzed via normal t-test. Responses during stimulation were normalized based on the baseline response 10s before the stimulation. P-values for each experiment are given in the respective figure panel. Each graph shows the mean calcium signal plotted as the relative response strength ΔF/F and the related standard error of the mean on the y axis. The time in seconds is given below each graph on the x axis. The grey box indicates the duration of the stimulus application. The sample size of each group (N=5-7) is given above each row. n.s. p > 0.05; * p < 0.05.

Behavioral Experiments

Associative olfactory learning

Standard experiments were done as described before (Apostolopoulou et al 2013, Gerber et al 2013, Hendel et al 2005, Michels et al 2017, Widmann et al 2018). Learning experiments were conducted on assay plates (85 mm diameter, Sarstedt, cat. no. 82.1472) filled with a thin layer of either 2.5% (w/v) pure agarose solution (Sigma Aldrich, cat. no. A9539) or 2.5% (w/v) agarose plus either 1.5 M sodium chloride solution (VWR Chemicals, cat. no. 27810.364), 2 M D-Fructose solution (Sigma Aldrich cat. no. 47740) or 10 mM quinine solution (quinine hemisulfate; Sigma Aldrich cat. no. Q1250). Before closing and labeling the lids, solutions were let to cool down at room temperature to avoid condensation. Plates were stored at 18°C and used within five days. As olfactory stimuli, we used two different odor combinations: either amyl acetate (AM, Sigma Aldrich, cat. no. 46022) diluted 1:250 in paraffin oil (Sigma Aldrich, cat. no. 76235) and undiluted benzaldehyde (BA, Sigma Aldrich, cat. no. 12010) or benzaldehyde and hexyl acetate (HA, Sigma Aldrich, cat. no. 108154) both diluted 1:100 in paraffin oil. For both odor combinations, 10 µL odor were loaded into custom-made Teflon containers (4.5 mm diameter) with perforated lids (Scherer et al 2003).

A minimum of 30 early third instar larvae were collected and exposed to a first odor (amyl acetate or hexyl acetate) while crawling on pure agarose for five minutes, followed by five minutes exposition to a second odor (benzaldehyde) on agarose medium with either 1.5 M sodium chloride, 10 mM quinine as negative reinforcer or 2 M D-Fructose as positive reinforcer (Odor1/Odor2+). For three cycle training, two more repetitions of training trials were performed. A second group of larvae received reciprocal training (Odor1+/Odor2). After one or three training cycles, larvae were transferred onto test plates containing either pure agarose (reward learning) or agarose plus salt or quinine (punishment learning) on which odor 1 and odor 2 were represented on opposite sides. After five minutes, larvae were counted on each side of the test plate (#Odor1, #Odor2) or in a 10 mm neutral zone in the middle of the plate (#Neutral). A Preference index (PREF) for each group of larvae was calculated as follows: To measure specifically the effect of associative learning, we then calculated the associative performance index (PI) as the difference in preference between the reciprocally trained larvae: Negative Performance Indices represent aversive associated learning, whereas positive values represent appetitive learning. Division by 2 ensures scores are bound within (-1; 1).

Odor preference test

Pure agarose (2.5% w/v) plates were cast as previously described (Apostolopoulou et al 2013, Gerber et al 2013, Michels et al 2017, Scherer et al 2003, Widmann et al 2018). Furthermore, the odors were prepared according to the dilutions used in the learning experiments. An odor container (either AM 1:250, BA undiluted, HA 1:100, or BA 1:100) was placed on one side of the agarose plate. An empty container was placed on the opposite side to exclude visual or other side effects. Approximately 30 third instar feeding stage larvae were collected and placed on the plate for 5 min. After this time, larvae were counted, subdivided into larvae on the odor side (#Odor), larvae on the empty container side (#Empty) and larvae in a 10 mm neutral zone in the middle of the plate, as well as larvae that were on the lid of the Petri dish (#Neutral). Preferences were calculated as follows: Positive values indicate a preference for the respective odor, while negative values indicate odor avoidance. The preference tests were performed in different orientations (odor container on plate up/down/left/right) to exclude directional bias.

Gustatory preference test

High salt-dependent choice behavior experiments were performed using standard methods (Hendel et al 2005, Huser et al 2017, Huser et al 2012, Niewalda et al 2008, Selcho et al 2009, Widmann et al 2016). A 2.5% (w/v) agarose solution was boiled as previously described and a thick layer was poured into Petri dishes. After cooling, half of the agarose in the Petri dish was removed and filled by 2.5% (w/v) agarose solution with 1.5 M sodium chloride. For the choice experiment, about 30 third instar feeding stage larvae were put in the middle of the Petri dish and let them crawl for five minutes. Larvae were then counted as being located on the sodium chloride side (#NaCl), pure agarose side (#Agarose), or a neutral area of about 10 mm width in the middle of the plate (#Neutral), as well as larvae that were on the lid of the Petri dish (#Neutral). The gustatory preference indices for sodium chloride were calculated as follows: Negatives values for preference indices indicate an aversion to sodium chloride.

Optogenetic substitution experiment

To substitute an actual salt punishment by remotely activating DANs, we used UAS-ChR2XXL. Effector lines were crossed to w1118 to obtain appropriate genetic controls. Larvae were maintained in vials with Drosophila standard food on 25°C and wrapped in aluminum foil to ensure development in constant darkness. A group of 30 feeding-stage third-instar larvae were placed onto plates containing 2.5% agarose and exposed to either AM or BA. During the presentation of the first odor, the larvae were exposed to blue light (470 nm, 2.5 V, 220 lux) for 5 minutes. The second odor BA (undiluted) was presented in darkness for five minutes on a pure agarose plate. As described for odor-high salt learning, training was performed reciprocally and the sequence of training trials was alternated across repetitions of the experiment. Data were then scored as above. Please note, tests were done in the presence of 1.5 M sodium chloride to test for the specific quality of the memory.

Optogenetic inhibition of neuronal activity

To acutely block synaptic output, we used UAS-GtACR2. Flies were maintained on standard food supplemented with 0.5 mM all-trans-retinal (ATR, Sigma Aldrich, cat. no. R2500) at 25°C as described before (Meloni et al 2020). Vials were wrapped in aluminum foil to ensure larval development in constant darkness. Groups of about 30 third instar feeding stage larvae received a reciprocal two-odor training as described before, exposed to blue light (470 nm, 2.7 V, 1100 lux) during the entire training phase. The test was performed for 5 minutes in darkness on an agarose plate mixed with 1.5 M sodium chloride. Control experiments were performed with the same genotype but with standard food lacking 0.5 mM all-trans-retinal. Data were then calculated and score as mentioned above.

Evaluation of DAN input wiring diagrams

Pie charts of sensory profiles were calculated using the percentage of total synaptic input of interneurons as fraction (thereby ignoring other inputs to show distribution of sensory origins). Percentages then give the percentage of total sensory synaptic input to interneurons. The calculation of the hub score was done in the following way: Fraction of total synaptic input from all sensory neurons to defined interneurons (see IDs) was multiplied by the total fraction of input of the DANs from this interneuron. For anatomical reconstructions and visualizations included here, we made use of earlier published data (Eichler et al 2017, Eschbach et al 2020, Miroschnikow et al 2018, Winding et al 2023).

Statistical Analysis

Behavioral results were analyzed in GraphPad Prism 8.4.3. To test for normal distribution, Shapiro-Wilk test was applied. Groups that did not violate the assumption of normal distribution, were analyzed with an unpaired t-test (comparison of two groups) or a one-way ANOVA followed by a Tukey’s post hoc test (comparisons between groups larger than two). For nonparametric statistics, Mann-Whitney test (comparison between two groups) or Kruskal-Wallis followed by Dunn’s multiple comparisons tests were applied. Furthermore, Wilcoxon signed-rank tests were performed to compare medians against chance level. Results are visualized in box-plots, indicating the median as middle line, 25%/75% quantiles as box boundaries and minimum/maximum performance indices as whiskers. Each data point is represented as black dot and sample sizes are noted within graphics. Asterisks and ‘‘n.s.’’ indicate p < 0.05 and p > 0.05, respectively.

Supplemental Information

Supplemental Information includes Supplemental Experimental Procedures and additional Figures and can be found with this article online.

Supporting information

eLife Assessment

This comprehensive study presents valuable results that delineate the involvement of a small subset of (DL1) dopaminergic neurons in the Drosophila larva's aversive learning response to high salt. Systematic loss-of-functional and gain-of-function manipulations coupled with in-vivo calcium imaging offer compelling evidence for the pivotal roles of these neurons, thereby advancing our understanding of the cellular mechanisms underlying associative conditioning. Despite its report of notable similarities between the learning mechanisms of learning in flies and mammals, the work underscores the necessity to further elucidate the interplay between aversive and appetitive pathways in future work.

Reviewer #1 (Public Review):

Summary:

In this paper, Weber et al. investigate the role of 4 dopaminergic neurons of the Drosophila larva in mediating the association between an aversive high-salt stimulus and a neutral odor. The 4 DANs belong to the DL1 cluster and innervate non-overlapping compartments of the mushroom body, distinct from those involved in appetitive associative learning. Using specific driver lines, they show that activation of the DAN-g1 is sufficient to mimic an aversive memory and it is also necessary to form a high-salt memory of full strength, although optogenetic silencing of this neuron only partially affects the performance index. The authors use calcium imaging to show that the DAN-g1 is not the only one that responds to salt. DAN-c1 and d1 also respond to salt, but they seem to play no role in the assays tested. DAN-f1, which does not respond to salt, is able to lead to the formation of memory (if optogenetically activated), but it is not necessary for the salt-odor memory formation in normal conditions. However, silencing of DAN-f1 together with DAN-g1, enhances the memory deficit of DAN-g1.

Strengths:

The paper therefore reveals that also in the Drosophila larva as in the adult, rewards and punishments are processed by exclusive sets of DANs and that a complex interaction between a subset of DANs mediates salt-odor association.

Overall, the manuscript contributes valuable results that are useful for understanding the organization and function of the dopaminergic system. The behavioral role of the specific DANs is accessed using specific driver lines which allow for testing of their function individually and in pairs. Moreover, the authors perform calcium imaging to test whether DANs are activated by salt, a prerequisite for inducing a negative association with it. Proper genetic controls are carried across the manuscript.

Weaknesses:

The authors use two different approaches to silence dopaminergic neurons: optogenetics and induction of apoptosis. The results are not always consistent, and the authors could improve the presentation and interpretation of the data. Specifically, optogenetics seems a better approach than apoptosis, which can affect the overall development of the system, but apoptosis experiments are used to set the grounds of the paper.

The physiological data would suggest the role of a certain subset of DANs in salt-odor association, but a different partially overlapping set seems to be necessary. This should be better discussed and integrated into the author's conclusion. The EM data analysis reveals a non-trivial organization of sensory inputs into DANs and it is hard to extrapolate a link to the functional data presented in the paper.

Reviewer #2 (Public Review):

Summary:

In this work, the authors show that dopaminergic neurons (DANs) from the DL1 cluster in Drosophila larvae are required for the formation of aversive memories. DL1 DANs complement pPAM cluster neurons which are required for the formation of attractive memories. This shows the compartmentalized network organization of how an insect learning center (the mushroom body) encodes memory by integrating olfactory stimuli with aversive or attractive teaching signals. Interestingly, the authors found that the 4 main dopaminergic DL1 neurons act redundantly, and that single-cell ablation did not result in aversive memory defects. However, ablation or silencing of a specific DL1 subset (DAN-f1,g1) resulted in reduced salt aversion learning, which was specific to salt but no other aversive teaching stimuli were tested. Importantly, activation of these DANs using an optogenetic approach was also sufficient to induce aversive learning in the presence of high salt. Together with the functional imaging of salt and fructose responses of the individual DANs and the implemented connectome analysis of sensory (and other) inputs to DL1/pPAM DANs, this represents a very comprehensive study linking the structural, functional, and behavioral role of DL1 DANs. This provides fundamental insight into the function of a simple yet efficiently organized learning center which displays highly conserved features of integrating teaching signals with other sensory cues via dopaminergic signaling.

Strengths:

This is a very careful, precise, and meticulous study identifying the main larval DANs involved in aversive learning using high salt as a teaching signal. This is highly interesting because it allows us to define the cellular substrates and pathways of aversive learning down to the single-cell level in a system without much redundancy. It therefore sets the basis to conduct even more sophisticated experiments and together with the neat connectome analysis opens the possibility of unraveling different sensory processing pathways within the DL1 cluster and integration with the higher-order circuit elements (Kenyon cells and MBONs). The authors' claims are well substantiated by the data and clearly discussed in the appropriate context. The authors also implement neat pathway analyses using the larval connectome data to its full advantage, thus providing network pathways that contribute towards explaining the obtained results.

Weaknesses:

While there is certainly room for further analysis in the future, the study is very complete as it stands. Suggestions for clarification are minor in nature.

Reviewer #3 (Public Review):

The study of Weber et al. provides a thorough investigation of the roles of four individual dopamine neurons for aversive associative learning in the Drosophila larva. They focus on the neurons of the DL-1 cluster which already have been shown to signal aversive teaching signals. However, the authors go far beyond the previous publications and test whether each of these dopamine neurons responds to salt or sugar, is necessary for learning about salt, bitter, or sugar, and is sufficient to induce a memory when optogenetically activated. In addition, previously published connectomic data is used to analyze the synaptic input to each of these dopamine neurons. The authors conclude that the aversive teaching signal induced by salt is distributed across the four DL-1 dopamine neurons, with two of them, DAN-f1 and DAN-g1, being particularly important. Overall, the experiments are well designed and performed, support the authors' conclusions, and deepen our understanding of the dopaminergic punishment system.

Strengths:

1. This study provides, at least to my knowledge, the first in vivo imaging of larval dopamine neurons in response to tastants. Although the selection of tastants is limited, the results close an important gap in our understanding of the function of these neurons.

2. The authors performed a large number of experiments to probe for the necessity of each individual dopamine neuron, as well as combinations of neurons, for associative learning. This includes two different training regimens (1 or 3 trials), three different tastants (salt, quinine, and fructose) and two different effectors, one ablating the neuron, the other one acutely silencing it. This thorough work is highly commendable, and the results prove that it was worth it. The authors find that only one neuron, DAN-g1, is partially necessary for salt learning when acutely silenced, whereas a combination of two neurons, DAN-f1 and DAN-g1, are necessary for salt learning when either being ablated or silenced.

3. In addition, the authors probe whether any of the DL-1 neurons is sufficient for inducing an aversive memory. They found this to be the case for three of the neurons, largely confirming previous results obtained by a different learning paradigm, parameters, and effector.

4. This study also takes into account connectomic data to analyze the sensory input that each of the dopamine neurons receives. This analysis provides a welcome addition to previous studies and helps to gain a more complete understanding. The authors find large differences in inputs that each neuron receives, and little overlap in input that the dopamine neurons of the "aversive" DL-1 cluster and the "appetitive" pPAM cluster seem to receive.

5. Finally, the authors try to link all the gathered information in order to describe an updated working model of how aversive teaching signals are carried by dopamine neurons to the larva's memory center. This includes important comparisons both between two different aversive stimuli (salt and nociception) and between the larval and adult stages.

Weaknesses:

1. The authors repeatedly claim that they found/proved salt-specific memories. I think this is problematic to some extent.

1a. With respect to the necessity of the DL-1 neurons for aversive memories, the authors' notion of salt-specificity relies on a significant reduction in salt memory after ablating DAN-f1 and g1, and the lack of such a reduction in quinine memory. However, Fig. 5K shows a quite suspicious trend of an impaired quinine memory which might have been significant with a higher sample size. I therefore think it is not fully clear yet whether DAN-f1 and DAN-g1 are really specifically necessary for salt learning, and the conclusions should be phrased carefully.

1b. With respect to the results of the optogenetic activation of DL-1 neurons, the authors conclude that specific salt memories were established because the aversive memories were observed in the presence of salt. However, this does not prove that the established memory is specific to salt - it could be an unspecific aversive memory that potentially could be observed in the presence of any other aversive stimuli. In the case of DAN-f1, the authors show that the neuron does not even get activated by salt, but is inhibited by sugar. Why should activation of such a neuron establish a specific salt memory? At the current state, the authors clearly showed that optogenetic activation of the neurons does induce aversive memories - the "content" of those memories, however, remains unknown.

2. In many figures (e.g. figures 4, 5, 6, supplementary figures S2, S3, S5), the same behavioural data of the effector control is plotted in several sub-figures. Were these experiments done in parallel? If not, the data should not be presented together with results not gathered in parallel. If yes, this should be clearly stated in the figure legends.

Author response:

The following is the authors’ response to the original reviews

Public reviews:

Reviewer #1 (Public Review):

Summary:

In this paper, Weber et al. investigate the role of 4 dopaminergic neurons of the Drosophila larva in mediating the association between an aversive high-salt stimulus and a neutral odor. The 4 DANs belong to the DL1 cluster and innervate non-overlapping compartments of the mushroom body, distinct from those involved in appetitive associative learning. Using specific driver lines, they show that activation of the DAN-g1 is sufficient to mimic an aversive memory and it is also necessary to form a high-salt memory of full strength, although optogenetic silencing of this neuron only partially affects the performance index. The authors use calcium imaging to show that the DAN-g1 is not the only one that responds to salt. DAN-c1 and d1 also respond to salt, but they seem to play no role in the assays tested. DAN-f1, which does not respond to salt, is able to lead to the formation of memory (if optogenetically activated), but it is not necessary for the salt-odor memory formation in normal conditions. However, silencing of DAN-f1 together with DAN-g1, enhances the memory deficit of DAN-g1.

Strengths:

The paper therefore reveals that also in the Drosophila larva as in the adult, rewards and punishments are processed by exclusive sets of DANs and that a complex interaction between a subset of DANs mediates salt-odor association.

Overall, the manuscript contributes valuable results that are useful for understanding the organization and function of the dopaminergic system. The behavioral role of the specific DANs is accessed using specific driver lines which allow for testing of their function individually and in pairs. Moreover, the authors perform calcium imaging to test whether DANs are activated by salt, a prerequisite for inducing a negative association with it. Proper genetic controls are carried across the manuscript.

Weaknesses:

The authors use two different approaches to silence dopaminergic neurons: optogenetics and induction of apoptosis. The results are not always consistent, and the authors could improve the presentation and interpretation of the data. Specifically, optogenetics seems a better approach than apoptosis, which can affect the overall development of the system, but apoptosis experiments are used to set the grounds of the paper.

The physiological data would suggest the role of a certain subset of DANs in salt-odor association, but a different partially overlapping set seems to be necessary. This should be better discussed and integrated into the author's conclusion. The EM data analysis reveals a non-trivial organization of sensory inputs into DANs and it is hard to extrapolate a link to the functional data presented in the paper.

We would like to thank reviewer 1 for the positive evaluation of our work and for the critical suggestions for improvement. In the new version of the manuscript, we have centralized the optogenetic results and moved some of the ablation experiments to the Supplement. We also discuss in detail the experimental differences in the results. In addition, we have softened our interpretation of the specificity of memory for salt. As a result, we now emphasize more the general role of DANs for aversive learning in the larva. These changes are now also summarized and explained more simply and clearly in the Discussion, along with a revised discussion of the EM data.

Reviewer #2 (Public Review):

Summary:

In this work, the authors show that dopaminergic neurons (DANs) from the DL1 cluster in Drosophila larvae are required for the formation of aversive memories. DL1 DANs complement pPAM cluster neurons which are required for the formation of attractive memories. This shows the compartmentalized network organization of how an insect learning center (the mushroom body) encodes memory by integrating olfactory stimuli with aversive or attractive teaching signals. Interestingly, the authors found that the 4 main dopaminergic DL1 neurons act redundantly, and that single-cell ablation did not result in aversive memory defects. However, ablation or silencing of a specific DL1 subset (DAN-f1,g1) resulted in reduced salt aversion learning, which was specific to salt but no other aversive teaching stimuli were tested. Importantly, activation of these DANs using an optogenetic approach was also sufficient to induce aversive learning in the presence of high salt. Together with the functional imaging of salt and fructose responses of the individual DANs and the implemented connectome analysis of sensory (and other) inputs to DL1/pPAM DANs, this represents a very comprehensive study linking the structural, functional, and behavioral role of DL1 DANs. This provides fundamental insight into the function of a simple yet efficiently organized learning center which displays highly conserved features of integrating teaching signals with other sensory cues via dopaminergic signaling.

Strengths:

This is a very careful, precise, and meticulous study identifying the main larval DANs involved in aversive learning using high salt as a teaching signal. This is highly interesting because it allows us to define the cellular substrates and pathways of aversive learning down to the single-cell level in a system without much redundancy. It therefore sets the basis to conduct even more sophisticated experiments and together with the neat connectome analysis opens the possibility of unraveling different sensory processing pathways within the DL1 cluster and integration with the higher-order circuit elements (Kenyon cells and MBONs). The authors' claims are well substantiated by the data and clearly discussed in the appropriate context. The authors also implement neat pathway analyses using the larval connectome data to its full advantage, thus providing network pathways that contribute towards explaining the obtained results.

Weaknesses:

While there is certainly room for further analysis in the future, the study is very complete as it stands. Suggestions for clarification are minor in nature.

We would like to thank reviewer 2 for the positive evaluation of our work. In fact, follow-up work is already underway to further analyze the role of the individual DL1 DANs. We have addressed the constructive and detailed suggestions for improvement in our point-by-point responses in the “Recommendations for the authors” section.

Reviewer #3 (Public Review):

The study of Weber et al. provides a thorough investigation of the roles of four individual dopamine neurons for aversive associative learning in the Drosophila larva. They focus on the neurons of the DL-1 cluster which already have been shown to signal aversive teaching signals. However, the authors go far beyond the previous publications and test whether each of these dopamine neurons responds to salt or sugar, is necessary for learning about salt, bitter, or sugar, and is sufficient to induce a memory when optogenetically activated. In addition, previously published connectomic data is used to analyze the synaptic input to each of these dopamine neurons. The authors conclude that the aversive teaching signal induced by salt is distributed across the four DL-1 dopamine neurons, with two of them, DAN-f1 and DAN-g1, being particularly important. Overall, the experiments are well designed and performed, support the authors' conclusions, and deepen our understanding of the dopaminergic punishment system.

Strengths:

(1) This study provides, at least to my knowledge, the first in vivo imaging of larval dopamine neurons in response to tastants. Although the selection of tastants is limited, the results close an important gap in our understanding of the function of these neurons.

(2) The authors performed a large number of experiments to probe for the necessity of each individual dopamine neuron, as well as combinations of neurons, for associative learning. This includes two different training regimens (1 or 3 trials), three different tastants (salt, quinine, and fructose) and two different effectors, one ablating the neuron, the other one acutely silencing it. This thorough work is highly commendable, and the results prove that it was worth it. The authors find that only one neuron, DAN-g1, is partially necessary for salt learning when acutely silenced, whereas a combination of two neurons, DAN-f1 and DAN-g1, are necessary for salt learning when either being ablated or silenced.

(3) In addition, the authors probe whether any of the DL-1 neurons is sufficient for inducing an aversive memory. They found this to be the case for three of the neurons, largely confirming previous results obtained by a different learning paradigm, parameters, and effector.

(4) This study also takes into account connectomic data to analyze the sensory input that each of the dopamine neurons receives. This analysis provides a welcome addition to previous studies and helps to gain a more complete understanding. The authors find large differences in inputs that each neuron receives, and little overlap in input that the dopamine neurons of the "aversive" DL-1 cluster and the "appetitive" pPAM cluster seem to receive.

(5) Finally, the authors try to link all the gathered information in order to describe an updated working model of how aversive teaching signals are carried by dopamine neurons to the larva's memory center. This includes important comparisons both between two different aversive stimuli (salt and nociception) and between the larval and adult stages.

Weaknesses:

(1) The authors repeatedly claim that they found/proved salt-specific memories. I think this is problematic to some extent.

(1a) With respect to the necessity of the DL-1 neurons for aversive memories, the authors' notion of salt-specificity relies on a significant reduction in salt memory after ablating DAN-f1 and g1, and the lack of such a reduction in quinine memory. However, Fig. 5K shows a quite suspicious trend of an impaired quinine memory which might have been significant with a higher sample size. I therefore think it is not fully clear yet whether DAN-f1 and DAN-g1 are really specifically necessary for salt learning, and the conclusions should be phrased carefully.

(1b) With respect to the results of the optogenetic activation of DL-1 neurons, the authors conclude that specific salt memories were established because the aversive memories were observed in the presence of salt. However, this does not prove that the established memory is specific to salt - it could be an unspecific aversive memory that potentially could be observed in the presence of any other aversive stimuli. In the case of DAN-f1, the authors show that the neuron does not even get activated by salt, but is inhibited by sugar. Why should activation of such a neuron establish a specific salt memory? At the current state, the authors clearly showed that optogenetic activation of the neurons does induce aversive memories - the "content" of those memories, however, remains unknown.

(2) In many figures (e.g. figures 4, 5, 6, supplementary figures S2, S3, S5), the same behavioural data of the effector control is plotted in several sub-figures. Were these experiments done in parallel? If not, the data should not be presented together with results not gathered in parallel. If yes, this should be clearly stated in the figure legends.

We would also like to thank reviewer 3 for his positive assessment of our work. As already mentioned by reviewer 1, we understand the criticism that the salt specificity for which the individual DANs are coded is not fully always supported by the results of the work. We have therefore rewritten the relevant passages, which are also cited by the reviewer. We have also included the second point of criticism and incorporated it into our manuscript. As the control groups were always measured in parallel with the experimental animals, we can also present the data together in a sub-figure. We clearly state this now in the revised figure legends.

Summary of recommendations to authors:

Overall, the study is commendable for its systematic approach and solid methodology. Several weaknesses were identified, prompting the need for careful revisions of the manuscript:

We thank the reviewers for the careful revision of our manuscript. In the subsequent sections, we aim to address their concerns as thoroughly as possible. A comprehensive one-to-one listing can be found below.

(1) The authors should reconsider their assertion of uncovering a salt-specific memory, as the evidence does not conclusively demonstrate the exclusive necessity of DAN-f1 and DAN-g1 for salt learning. In particular, the optogenetic activation of DAN-f1 leads to plasticity but this might not be salt-specific. The precise nature of the memory content remains elusive, warranting a nuanced rephrasing of the conclusions.

We only partially agree – optogenetic activation of DANs does not really allow to comment on its salt-specificity, true. However, we used high-salt concentrations during test. Over the years, the Gerber lab nicely demonstrated in several papers that larvae recall an aversive odor-salt memory only if salt is present during test (Gerber and Hendel, 2006; Niewalda et al 2008; Schleyer et al. 2011; Schleyer et al. 2015). The used US has to be present during test. Even at the same concentration other aversive stimuli (e.g. bitter quinine) are not able to allow the larvae to recall this particular type of memory. So, if the optogenetic activation of DAN-f1 establishes a memory that can be recalled on salt, we argue that it has to encode aspects of the salt information. On the other hand, only for DAN-g1 we see the necessity for salt learning. And – although (based on the current literature) very unlikely, we cannot fully exclude that the activation of DAN-f1 establishes a yet unknown type of memory that can be also recalled on a salt plate. Therefore, we partially agree and accordingly have rephrased the entire manuscript to avoid an over-interpretation of our data. Throughout the manuscript we avoid now to use the term salt-specific memory but rather describe the type of memory as aversive memory.

(2) A thorough examination or discussion about the potential influence of blue light aversion on behavioral observations is necessary to ensure a balanced interpretation of the findings.

To address this point every single behavioral experiment that uses optogenetic blue light activation runs with appropriate and mandatory controls. For blue light activation experiments, two genetic controls are used that either get the same blue light treatment (effector control, w1118>UAS-ChR2XXL) or no blue light treatment (dark control, XY-split-Gal4>UAS-ChR2XXL). For blue light inactivation experiments one group is added that has exactly the same genotype but did not receive food containing retinal. These experiments show that blue light exposure itself does not induce an aversive nor positive memory and blue light exposure does not impair the establishment of odor-high salt memory. In addition, we used the latest established transgenes available. ChR2XXL is very sensitive to blue light. Only 220 lux (60 µW/cm²) were necessary to obtain stable results. In our hands – short term exposure for up to 5 minutes with such low intensities does not induce a blue light aversion. Following the advice of the reviewer, we also address this concern by adding several sentences into the related results and methods sections.

(3) The authors should address the limitations associated with the use of rpr/hid for neuronal ablations, such as the effects of potential developmental compensation.

We agree with this concern. It is well possible that the ablation experiments induce compensatory effects during larval development. Such an effect may be the reason for differences in phenotypes when comparing hid,rpr ablation with optogenetic inhibition. This is now part of the discussion. In addition, we evaluated if the ablation worked in our experiments. So far controls were missing that show that the expression of hid,rpr really leads to the ablation of DANs. We now added these experiments and clearly show anatomically that the DANs are ablated (related to figure 4-figure supplement 6).

(4) While the connectome analysis offers valuable insights into the observed functions of specific DANs in relation to their extrinsic (sensory) and intrinsic (state) inputs, integrating this data more cohesively within the manuscript through careful rewriting would enhance the coherence of the study.

We understand this concern. Therefore, the new version of our manuscript is now intensifying the inclusion of the EM data in our interpretation of the results. Throughout the entire manuscript we have now rewritten the related parts. We have also completely revised the corresponding section in the results chapter.

(5) More generally, the authors are encouraged to discuss internal discrepancies in the results of their functional manipulation experiments.

Thank you for this suggestion. We do of course understand that we have not given the different results enough space in the discussion. We have now changed this and have been happy to comprehensively address the concern.

Recommendations for the authors:

Reviewer #1 (Recommendations For The Authors):

Here are some suggestions for clarification and improvement of the manuscript:

(1) The authors should discuss why the silencing experiment with TH-GAL4 (Fig. 1) does not abolish memory formation (I assume that the PI should go to zero). Does it mean that other non-TH neurons are involved in salt-odor memory formation? Are there other lines that completely abolish this type of learning?

Thank you very much for highlighting this crucial point. Indeed, the functional intervention does not completely eliminate the memory. There could be several reasons, or a combination thereof, for this outcome. For instance, it's plausible that the UAS-GtACR2 effector doesn't entirely suppress the activity of dopaminergic neurons. Additionally, the memory may comprise different types, not all of which are linked to dopamine function. It's also noteworthy that TH-Gal4 doesn't encompass all dopaminergic neurons – even a neuron from the DL1 cluster is absent (as previously reported in Selcho et al., 2009). Considering we're utilizing high salt concentrations in this experiment, it's conceivable that non gustatory-driven memories are formed based solely on the systemic effects of salt (e.g., increased osmotic pressure). These possibilities are now acknowledged in the text.

(2) The Rpr experiments in Fig. 4 do not lead to any phenotype and there is a general assumption that the system compensates during development. However, there is no demonstration that Rpr worked or that development compensated for that. What do we learn from these data? Would it make sense to move it to supplement to make the story more compact? In addition: the conclusion at L 236 "DL1.... Are not individually necessary" is later disproved by optogenetic silencing. Similarly, optogenetic silencing of f1+g1 is affecting 1X and 3X learning, but not when using Rpr. Moreover, Rpr wdid not give any phenotype in other data in the supplementary material. I'm not sure how valid these results are.

We acknowledge this concern and have actively deliberated various options for restructuring the presented ablation data. Ultimately, we reached a consensus that relocating Figure 4 to the supplement is warranted. Furthermore, corresponding adjustments have been made in the text. This decision amplifies the significance of the optogenetic results. In addition, we also addressed the other part of the concern. We examined the efficacy of hid and rpr in our experiments. Indeed, we successfully ablated specific DANs, as illustrated in the new anatomical data presented in Figure 4- figure supplement 6, which strengthens the interpretation of the hid,rpr experiments.

(3) In most figures that show data for 1X and 3X training, there is no difference between these two conditions (I would suggest moving one set as a supplement). When a difference appears (Fig.5A-D) the implications are not discussed properly. Is it known that some circuits are necessary for the 1X but not for the 3X protocol? Is that a reasonable finding? I would expect the opposite, but I might lack of knowledge here. However, the optogenetic silencing of the same neurons in Figure 7 shows the same phenotype for 1X and 3X. Again, the validity of the Rpr experiments seems debatable.

Different training protocols lead to different memory phases (STM and STM+ARM). We have shown that in the past in Widmann et al. 2016. Therefore, we are convinced that it makes sense to keep both data sets in the main manuscript. However, we agree that this was not properly introduced and discussed and therefore made the respective changes in the manuscript.

(4) In Figure 3, it is unclear what the responses were tested against. Since they are so small and noisy there would be a need for a control. Moreover, in some cases, it looks like the DF/F is normalized to the wrong value: e.g. in DAN-c1 100mM, the activity in 0-10s is always above zero, and in pPAM with fructose is always below zero. This might not have any consequence on the results but should be adjusted.

Thank you very much for your criticism, which we greatly appreciate. We have carefully re-examined the data and found that there was a mistake for the normalization of the values. We made the necessary adjustments to the evaluation, as per your suggestions. The updated figures, figure legends, and results have been incorporated into the new version of the manuscript. As noted by the reviewer, these corrections have not altered the interpretation of the data or the primary responses of the various DANs.

(5) In the abstract: "Optogenetic activation of DAN-f1 and DAN-g1 alone suffices to substitute for salt punishment... Each DAN encodes a different aspect of salt punishment". These sentences might be misleading and an overstatement: only DAN-g1 shows a clear role, while the function of the other DANs in the context of salt-odor learning remains obscure.

We have refined the respective part of the abstract accordingly. Consequently, we have reworded the related section, aiming to avoid any exaggeration.

(6) The physiology is done in L1 larvae but behavior is tested in L3 larvae. There could be a change in this time that could explain the salt responses in c1 and d1 but no role in salt-odor learning?

While we cannot dismiss the possibility of a developmental change from L1 to L3, a comparison of the anatomical data of the DL1 DANs from electron microscopy (EM) and light microscopy (LM) data indicates that their overall morphology remains consistent. However, it's important to note that this observation does not analyse the physiological aspects of these cells. Consequently, we have incorporated this concern into the discussion of the revised version of the manuscript.

(7) The introduction needs some editing starting at L 129, as it ends with a discussion of a previously published EM data analysis. I would rather suggest stating which questions are addressed in this paper and which methods will be used and perhaps a hint on the results obtained.

We understand the concern. We have added a concise paragraph to the conclusion of the introduction, highlighting the biological question, technical details, and a short hint on the acquired findings.

(8) It is clear to me that the presentation of salt during the test is necessary for recall, however in L 166 I don't understand the explanation: how is the memory used in a beneficial way in the test? The salt is present everywhere and the odor cue is actually useless to escape it.

Extensive research, exemplified by studies such as Schleyer et al. (2015) published in Elife, clearly demonstrates that the recall of odor-high salt memory occurs exclusively when tested on a high salt plate. Even when tested on a bitter quinine plate, the aversive memory is not recalled. This phenomenon is attributed to the triggering of motivation to recall the memory by the omnipresent abundance of the unconditioned stimulus (US) during the test, which in our case is high salt. Furthermore, the concentration of the stimulus plays a crucial role (Schleyer et al. 2011). The odor cue indicates where the situation could potentially be improved; however, if high salt is absent, this motivational drive diminishes as there is no memory present to enhance the already favorable situation. Additionally, the motivation to evade the omnipresent and unpleasant high salt stimulus persists throughout the entire 5-minute test period.

(9) L288: the fact that f1 shows a phenotype in this experiment does not mean that it encodes a salt signal, indeed it does not respond to salt. It perhaps induces a plasticity that can be recalled by salt, but not necessarily linked to salt. The synergy between f1 and g1 in the salt assay was postulated based on exp with Rpr, but the validity of these experiments is dubious. I'm not sure there is sufficient evidence from Figures 6 and 7 to support a synergistic action between f1 and g1.

It is true that DAN-f1 alone is not necessary for mediating a high salt teaching signal based on ablation, optogenetic inhibition and even physiology. However, optogenetic activation alone shows a memory tested on a salt plate. Given the logic explained above that is accepted by several publications, we would like to keep the statement. Especially as the joined activation with DAN-g1 gives rise to significant higher or lower values after joined optogenetic activation or inactivation (Figure 5E and F, Figure 6E and F in the new version). Nevertheless, we have modified the sentence. In the text we describe these effects now as “these results may suggest that DAN-f1 and DAN-g1 encode aspects of the natural aversive high salt teaching signal under the conditions that we tested”. We think that this is an appropriate and three-fold restricted statement. Therefore, we would like to keep it in this restricted version. However, we are happy to reconsider this if the reviewer thinks it is critical.

(10) I find the EM analysis hard to read. First of all, because of the two different graphical representations used in Fig. 8, wouldn't one be sufficient to make the point? Secondly, I could not grasp a take-home-message: what do we learn from the EM data? Do they explain any of the results? It seems to me that they don't provide an explanation of why some DL1 neurons respond to salt and others don't.

We understand that the EM analysis is hard to read and have now carefully rewritten this part of the manuscript. See also general concern 4 above. The main take home message is not to explain why some DL1 neurons respond to salt and other do not. This cannot be resolved due to the missing information on the salt perceiving receptor cells. Unfortunately, we miss the peripheral nervous system in the EM - the first layer of salt information processing. However, our analysis shows clearly that the 4 DANs have their own identity based on their connectivity. None of them is the same – but to a certain extent similarities exist. This nicely reflects the physiological and behavioral results. We have now clarified that in the result to ease the understanding for the readership. In addition, we also clearly state that we don’t address the point why some DL1 neurons respond to salt and why others don’t respond.

(11) Do the manipulations (activation and silencing) affect odor preference in the presence of salt? Did the authors test that the two odors do not drive different behaviors on the salty plate? Or did they only test the odor preference on plain agarose? Can we exclude a role for the DAN in driving multisensory-driven innate behavior?

Innate odor preferences are not changed by the presence of salt or even other tastants (this work but see also Schleyer et al 2015, Figure 3, Elife). Even the naïve choice between two odors is the same if tested in the presence of different tastants (Schleyer et al 2015, Figure 3, Elife). This shows – at least for the tested stimuli and conditions – that are similar to the ones that we use – that there is no multisensory-driven innate odor-taste behavior. Therefore – at least to our knowledge - experiments as the ones suggested by the reviewer were never done in larval odor-taste learning studies. Therefore, we suggest that DAN activation has no effect on innate larval behavior. However, we are happy to reconsider this if the reviewer thinks it is critical.

(12) L 280: the authors generalize the conclusion to all DL1-DANs, but it does not apply to c1 and d1.

Thanks for this comment. We deleted that sentence as suggested and thus do not anymore generalize the conclusion to all DL-DANs.

(13) L345: I do not see the described differences in Fig. 8F, presynaptic sites of both types seem to appear in rather broad regions: could the author try to clarify this?

We understand that the anatomical description of the data is often hard to read. Especially to readers that are not used to these kind of figures. We have therefore modified the text to ease the understanding and clarify the difference in the labeled brain regions for the broad readership.

(14) L373: the conclusion on c1 is unsupported by data: this neuron responds to both salt and fructose (Figure 3 ) while the conclusion is purely based on EM data analysis.

The sentence is not a conclusion but a speculation and we also list the cell's response to positive and negative gustatory stimuli. Therefore, we do not understand exactly what the reviewer means here. However, we have tried to address the criticism and have revised the sentences.

(15) L385: the data on d1 seem to be inconsistent with Eschbach 2020, but the authors do not discuss if this is due to the differential vs absolute training, or perhaps the presence of the US during the test (which does not seem to be there in Eschbach, 2020) - is the training protocol really responsible for this inconsistency? For f1 the data seem to be consistent across these studies. The authors should clarify how the exp in Fig 6 differs from Eschbach, 2020 and how one could interpret the differences.

True. This concern is correct. We now discuss the difference in more detail. Eschbach et al. used Cs-Crimson as a genetic tool, a one odor paradigm with 3 training cycles, and no gustatory cues in their approach. These differences are now discussed in the new version of the manuscript.

(16) L460-475 A long part of this paragraph discusses the similarities between c1 and d1 and corresponding PPL1 neurons in the adult fly. However, c1 and d1 do not really show any phenotype in this paper, I'm not sure what we learn from this discussion and how much this paper can contribute to it. I would have wished for a discussion of how one could possibly reconcile the observed inconsistencies.

Based on the comments of the different reviewers several paragraphs in the discussion were modified. We agree that the part on the larval-adult comparison is quite long. Thus we have shortened it as suggested by the reviewer.

Minor corrections:

L28 "resultant association" maybe resulting instead.

L55 "animals derive benefit": remove derive.

L78 "composing 12,000 neurons": composed of.

L79 what is stable in a "stable behavioral assay"?

L104: 2 times cluste.

L122: "DL1 DANs are involved" in what?

Fig. 1 please check subpanels labels, D repeats.

L 362: "But how do individual neurons contribute to the teaching signal of the complete cluster?" I don't understand the question.

L364 I did not hear before about the "labeled line hypothesis" in this context - could the author clarify?

L368: edit "combinatorically".

L390: "current suppression" maybe acute suppression.

L 400 I'm not sure what is meant by "judicious functional configuration" and "redundancy". The functions of these cells are not redundant, and no straightforward prediction of their function can be done from their physiological response to salt.

Thanks a lot for your in detail review of our manuscript. We welcome your well-taken concerns and have made the requested changes for all points that you have raised.

Reviewer #2 (Recommendations For The Authors):

(1) In Figure 1 the reconstruction of pPAM and DL1 DANs shows the compartmentalized innervation of the larval MB. However, the images are a bit low in color contrast to appreciate the innervation well. In particular in panel B, it is hard to identify the innervated MB body structure. A schematic model of the larval MB and DAN innervation domains like in Fig. 2A would help to clarify the innervation pattern to the non-specialist.

We understand this concern and have changed figure 1 as suggested by the reviewer. A schematic model of the MB and DANs is now presented already in figure 1 as well as the according supplemental figure.

(2) Blue light itself can be aversive for larvae and thus interfere with the aversive learning paradigm. Does the given Illuminance (220 lux) used in these experiments affect the behavior and learning outcome?

Yes, in former times high intensities of blue light were necessary to trigger the first generation optogenetic tools. The high intensity blue light itself was able to establish an aversive memory (e.g. Rohwedder et al. 2016). Usage of the second generation optogenetic tools allowed us to strongly reduce the applied light intensity. Now we use 220 lux (equal to 60 µW/cm2). Please note that all Gal4 and UAS controls in the manuscript are nonsignificant different from zero. The mild blue light stimulation therefore does not serve as a teaching signal and has neither an aversive nor an appetitive effect. Furthermore, we use this mild light intensity for several other behavioral paradigms (locomotion, feeding, naïve preferences) and have never seen an effect on the behavior.

(3) Fig.2: Except for MB054B-Gal4 only the MB expression pattern is shown for other lines. Is there any additional expression in other cells of the brain? In the legend in line 761, the reporter does not show endogenous expression, rather it is a fluorescent reporter signal labeling the mushroom body.

The lines were initially identified by a screen on larval MB neurons done together with Jim Truman, Marta Zlatic and Bertram Gerber. Here full brain scans were always analyzed. These images can be seen in Eschbach et al. 2020, extended figure 1. Neither in their evaluation nor in our anatomical evaluation (using a different protocol) additional expression in brain cells was detectable. We also modified the figure legend as suggested.

(4) Fig.3: Precise n numbers per experiment should be stated in the figure legend.

True, we now present n numbers per experiment whenever necessary.

(5) Fig.4: Have the authors confirmed complete ablation of the targeted neuron using rpr/hid? Ablations can be highly incomplete depending on the onset and strength of Gal4 expression, leaving some functionality intact. While the ablation experiments are largely in line with the acute silencing of single DANs during high salt learning performed later on (Fig.7), there is potentially an interesting aspect of developmental compensation hidden in this data. Not a major point, but potentially interesting to check.

We agree with this criticism. We have not tested if the expression of hid,rpr in DL1 DANs does really ablate them. Therefore we did an additional experiment to show that. The new data is now present as a supplemental figure (Figure 4- figure supplement 6). The result shows that expression of hid,rpr ablates also DL1 DANs similar to earlier experiments where we used the same effectors to ablate serotoniergic neurons (Huser et al., 2012, figure 5).

(6) The performance index in Fig. 4 and 5 sometimes seems lower and the variability is higher than in some of the other experiments shown. Is this due to the high intrinsic variability of these particular experiments, or the background effects of the rpr/hid or splitGal4 lines?

The general variability of these experiments is within the expected and known borders. In these kind of experiments there is always some variation due to several external factors (e.g. experimental time over the year). Therefore it is always important to measure controls and experimental animals at the same time. Of course that’s what we did and we only compare directly results of individual datasets. But not between different datasets. This is further hampered given that the experiments of Figure 4 (now Figure 4- figure supplement 1) and Figure 5 (now Figure 4) differ in several parameters from other learning experiments presented later in the text. Optogenetic activation uses blue light stimulation instead of “real world” high salt. Most often direct activation of specific DANs in the brain is more stable than the external high salt stimulation. Also optogenetic inactivation uses blue light stimulation and also retinal supplemented food. Both factors can affect the measurement. We thus want to argue that it is for each experiment most often the particular parameters that affect the variability of the results rather than background effects of the rpr/hid and split-Gal4 lines.

(7) Fig.7: This is a neat experiment showing the effects of acute silencing of individual DL1 DANs. As silencing DAN-f1/g1 does not result in complete suppression of aversive learning, it would be highly interesting to test (or speculate about) additive or modulatory effects by the other DANs. Dan-c-1/d-1 also responds to high salt but does not show function on its own in these assays. I am aware that this is currently genetically not feasible. It would however be a nice future experiment.

True, we were intensively screening for DL1 cluster specific driver lines that cover all 4 DL1 neurons or other combinations than the ones we tested. Unfortunately, we did not succeed in identifying them. Nevertheless, we will further screen new genetic resources (e.g. Meissner et al., 2024, bioRxiv) to expand our approach in future experiments. Please also see our comment on concern 1 of reviewer 1 for further technical limitations and biological questions that can also potentially explain the absence of complete suppression of high salt learning and memory. Some of these limitations are now also mentioned and discussed in the new version of the manuscript.

(8) The discussion is excellent. I would just amend that it is likely that larval DAN-c1, which has high interconnectivity within the larval CNS, is likely integrating state-dependent network changes, similar to the role of some DANs in innate and state-dependent preference behavior. This might contribute to modulating learned behavior depending on the present (acute) and previous environmental conditions.

Thanks a lot for bringing this up. We rewrote this part and added a discussion on recent work on DAN-c1 function in larvae as well as results on DAN function in innate and state-dependent preference behavior.

(9) Citation in line 1115 missing access information: "Schnitzer M, Huang C, Luo J, Je Woo S, Roitman L, et al. 2023. Dopamine signals integrate innate and learned valences to regulate memory dynamics. Research Square".

Unfortunately this escaped our notice. The paper is now published in Nature: Huang, C., Luo, J., Woo, S.J. et al. Dopamine-mediated interactions between short- and long-term memory dynamics. Nature 634, 1141–1149 (2024). https://doi.org/10.1038/s41586-024-07819-w. We have now changed the citation. The new citation includes the missing access information.

Reviewer #3 (Recommendations For The Authors):

Regarding my issue about salt specificity in the public review, I want to make clear that I do not suggest additional experiments, but to be very careful in phrasing the conclusions, in particular whenever referring to the experiments with optogenetic activation. This includes presenting these experiments as "(salt) substitution" experiments - inferring that the optogenetic activation would substitute for a natural salt punishment. As important and interesting as the experiments are, they simply do not allow such an interpretation at this point.

Results, line 140ff: When presenting the results regarding TH-Gal4 crossed to ChR2-XXL, please cite Schroll et al. 2006 who demonstrated the same results for the first time.

Thanks for mentioning this. We now cite Schroll et al. 2006 here in the text of the manuscript.

Figure 3: The subfigure labels (ABC) are missing.

Unfortunately this escaped our notice. Thanks a lot – we have now corrected this mistake.

Figure 5: For I and L, it reads "salt replaced with fru", but the sketch on the left shows salt in the test. I assume that fructose was not actually present in the test, and therefore the figure can be misleading. I suggest separate sketches. Also, I and L are not mentioned in the figure legend.

True, this is rather confusing. Based on the well taken concern we have changed the figure by adding a new and correct scheme for sugar reward learning that does not symbolize fructose during test.

Figure S1: The experimental sketches for E,F and G,H seem to be mixed up.

We thank the reviewer for bringing this up. In the new version we corrected this mistake.

Figure S5: There are three sub-figures labelled with B. Please correct.

Again, thanks a lot. We made the suggested correction in Figure S5.

Discussion, line 353ff: this and the following sentences can be read as if the authors have discovered the DL-1 neurons as aversive teaching mediators in this study. However, Eschbach et al. 2020 already demonstrated very similar results regarding the optogenetic activation of single DL-1 DANs. I suggest to rephrase and cite Eschbach et al. 2020 at this point.

That is correct. Our focus was on the gustatory pathway. The original discovery was made by Eschbach et al. We have now corrected this in the discussion and clarified our contribution. It was never our intention to hide this work, as the laboratory was also involved. Nevertheless, this is an annoying omission on our side.

Line 385-387: this sentence is only correct with respect to Eschbach et al. 2020. Weiglein et al. 2021 used ChR2-XXL as an effector, but another training regimen.

We understand this criticism. Therefore, we changed the sentence as suggested by the reviewer. See also our response on concern 15 of reviewer 1.

Line 389ff: I do not understand this sentence. What is meant by persistent and current suppression of activity? If this refers to the behavioural experiments, it is misleading as in the hid, reaper experiments neurons are ablated and not suppressed in activity.

We made the requested changes in the text. It is true that the ablation of a neuron throughout larval life is different from constantly blocking the output of a persisting neuron.

Methods, line 615 ff: the performance index is said to be calculated as the difference between the two preferences, but the equation shows the average of the preferences.

Thanks a lot. We are sorry for the confusion. We have carefully rewritten this part of the methods section to avoid any misunderstanding.

When discussing the organization of the DL1 cluster, on several occasions I have the impression the authors use the terms "redundant" and "combinatorial" synonymously. I suggest to be more careful here. Redundancy implies that each DAN in principle can "do the job", whereas combinatorial coding implies that only a combination of DANs together can "do the job". If "the job" is establishing an aversive salt memory, the authors' results point to redundancy: no experimental manipulation totally abolished salt learning, implying that the non-manipulated neurons in each experiment sufficed to establish a memory; and several DANs, when individually activated, can establish an aversive memory, implying that each of them indeed can "do the job".

Based on this concern we have rewritten the discussion as suggested to be more precise when talking about redundancy or combinatorial coding of the aversive teaching signal. Basically, we have removed all the combinatorial terms and replaced them by the term “redundancy”.

The authors mix parametric and non-parametric statistical tests across the experiments dependent on whether the distribution of the data is normal or not. It would help readers if the authors would clearly state for which data which tests were used.

We understand the criticism and now have added an additional supplemental file that includes all the information on the statistical tests applied and the distribution of the data.

Version 2

Abstract

Abstract

Dopaminergic neurons (DANs) carry out multiple tasks in the brain, including the transmission of information related to rewards and punishments across various animal species. They are responsible for evaluating sensory input, storing resulting associations as memory, and continuously updating them based on their relevance and reliability. Accurate comprehension of the dopaminergic system’s operation necessitates an understanding of the specific functions mediated by individual DANs. To this end, our research employs Drosophila larvae, which possess approximately 12,000 neurons in their brains, of which only around 1% (approximately 120) are DANs.

The presynaptic projections to the mushroom body (MB) - a brain region pivotal for associative olfactory learning in insects - are limited to only eight larval dopaminergic neurons. These DANs are further subdivided into two clusters: the primary protocerebral anterior medial cluster (pPAM) comprises four cells, and the dorsolateral 1 cluster (DL1) comprises the remaining four cells. Our findings confirm previous research that demonstrates that the pPAM DANs innervating the MB’s medial lobe encode for a gustatory sugar reward signal. Furthermore, we have identified four DANs in the DL1 cluster - DAN-c1, DAN-d1, DAN-f1, and DAN-g1 - each of which innervates distinct compartments of the MB peduncle, lateral appendix, and vertical lobe. Optogenetic activation of DAN-f1 and DAN-g1 alone suffices to substitute for punishment. Furthermore, optogenetic inhibition, calcium imaging results and electron microscopy-based reconstruction of all sensory input circuits to the four DL1 DANs demonstrate that each DAN encodes a different aspect of punishment, with DAN-g1 being of central importance for the salt dependent teaching signal.

To summarize, our investigation has revealed the existence of a cellular division of labor among larval DANs concerning the transmission of dopaminergic reward (pPAM cluster) and punishment signals (DL1 cluster). Individual DANs in each cluster encode for distinct but partially overlapping aspects of the teaching signal. The striking resemblance in the organizing principle of larval DANs with that of its adult counterpart and the mammalian basal ganglion suggests that there may be a limited number of efficient neural circuit solutions available to address more complex cognitive challenges in nature.

Introduction

To cope with a constantly changing environment, animals benefit from learning behavioral rules that need constant adaptation and updating. To test this ability, researchers often use Pavlovian or classical conditioning experiments (Pavlov 1927), which examine an organism’s capacity to form associations between sensory cues, called conditioned stimuli (CS), and rewards or punishments, referred to as aversive and appetitive unconditioned stimuli (US) or teaching signals (Gerber et al 2009, Glanzman 1995, Heisenberg 2003, Menzel 2022, Waddell 2013, Weber et al 2023b, Widmann et al 2018). Dopaminergic neurons (DANs) and other modulatory neurons mediate teaching signals that allow animals to classify a given CS into positive or negative valence based on past experience (Cognigni et al 2018, Menzel 2001, Schultz 2015, Thum & Gerber 2019, Watabe-Uchida et al 2017). For example, when a CS is paired with simultaneous activation of a specific set of DANs in the PAM cluster, larval or adult Drosophila subsequently perceive the odor as attractive (Burke et al 2012, Eschbach et al 2020, Liu et al 2012, Rohwedder et al 2016, Schroll et al 2006) – a similar effect was seen after optogenetic activation of DANs located in the ventral tegmental area of the mouse brain to induce conditioned place preference (Tsai et al 2009). In contrast, activation of larval DL1 DANs (Eschbach et al 2020, Schroll et al 2006, Weiglein et al 2021), adult PPL1 DANs (Aso et al 2012, Aso & Rubin 2016, Aso et al 2010, Claridge-Chang et al 2009), or DANs in the posterior tail of the stratum of mice were shown to reinforce avoidance learning (Menegas et al 2018). These examples illustrate that the dopaminergic system provides a highly conserved and fundamental neural principle of the brain, which has been retained throughout the process of evolution and executes comparable functions across diverse animal species, including even humans (Frick et al 2022).

Our research focuses on the Drosophila larva, an organism endowed with a relatively simple central nervous system, composed of approximately 12,000 neurons (Dumstrei et al 2003, Winding et al 2023). Classical conditioning in larvae can be analyzed by standardized behavioral assays (Almeida-Carvalho et al 2017, Apostolopoulou et al 2013, Scherer et al 2003, Weber et al 2023a), in conjunction with a wealth of genetic methods for targeting single molecules and neurons (Dietzl et al 2007, Li et al 2014, Perkins et al 2015, Port et al 2020). In addition, the connectome of the brain was recently reconstructed, which enables the extraction and analysis of related wiring motifs, as well as entire learning and memory circuits (Winding et al 2023). Based on these advances, it was possible to establish a more fundamental cellular and synaptic understanding of the larval memory circuit, the mushroom body (MB) network. The MB is a higher-order parallel fiber system present in numerous invertebrate brains, including hemimetabolous as well as holometabolous insects and their larval stages (Strausfeld et al 1998, Strausfeld et al 2020). Analogous to the vertebrate cerebrum, the MB in insects is responsible for forming and maintaining associations (Heisenberg 2003, Menzel 2001, Menzel 2022, Thum & Gerber 2019, Tomer et al 2010, Waddell 2013). In the first developmental stage (L1), the larval MB comprises approximately 110 intrinsic Kenyon cells (KC) per hemisphere, which sparsely encode for conditioned stimulus (CS) information, and is dominated by olfactory input (Eichler et al 2017, Masuda-Nakagawa et al 2009).

The transmission of olfactory information from the peripheral sensory organ is achieved via only 21 olfactory receptor neurons (ORNs) per body side, which in turn send signals across 35 uni- and multiglomerular projection neurons, ultimately reaching the MB KCs (Berck et al 2016, Fishilevich et al 2005, Kreher et al 2005, Masuda-Nakagawa et al 2009, Ramaekers et al 2005). KCs receive modulatory input from 17 neurons, consisting of eight dopaminergic neurons, four octopaminergic neurons, and five neurons with unknown signaling molecule identity (Eichler et al 2017, Rohwedder et al 2016, Saumweber et al 2018, Selcho et al 2009). The DANs are divided into two clusters based on the location of their soma and related neuronal lineage, namely the primary protocerebral anterior medial cluster (pPAM) and the dorsolateral 1 cluster (DL1). Each DAN innervates a specific and non-overlapping site in the MB, which, together with one to three output neurons, defines a distinct compartment (Eichler et al 2017). The larval MB comprises a total of 11 compartments, eight of which are defined by DAN input (Eichler et al 2017, Saumweber et al 2018). The four DANs of the pPAM cluster (DAN-h1, DAN-i1, DAN-j1, and DAN-k1) innervate the four related compartments (h, i, j, and k) of the medial lobe and provide teaching signals for sugar rewards (Eichler et al 2017, Rohwedder et al 2016, Schleyer et al 2020). The four DANs of the DL1 cluster (DAN-c1, DAN-d1, DAN-f1, and DAN-g1) innervate the lower peduncle (c), the lateral appendix (d), the intermediate (f), and the lower vertical lobe (g) (Eichler et al 2017, Saumweber et al 2018). Previous studies have shown that the DL1 cluster is capable of mediating aversive teaching signals (Eschbach et al 2020, Schroll et al 2006, Selcho et al 2009, Weiglein et al 2021). Moreover, larvae can use a variety of natural stimuli as aversive teaching signals such as heat, electric shock, vibration, bitter substances (like quinine or caffeine), and high salt concentrations (low salt concentrations, however, are rewarding) (Aceves-Pina & Quinn 1979, Apostolopoulou et al 2016, Apostolopoulou et al 2014, Eschbach et al 2011, Gerber & Hendel 2006, Khurana et al 2009, Khurana et al 2012, Niewalda et al 2008, Pauls et al 2010, von Essen et al 2011). However, it is currently unknown if and how these natural stimuli are represented by the DL1 cluster.

We only know that optogenetic stimulation of nociceptive sensory neurons and their downstream basin neurons has been observed to activate three out of the four DL1 DANs, namely DAN-d1, DAN-f1, and DAN-g1 (Eschbach et al 2020). Furthermore, the simultaneous presentation of odor with optogenetic activation of these three DANs has been found to elicit aversion in larvae (Eschbach et al 2020, Weiglein et al 2021). These findings suggest that the activation of DL1 DANs may induce a punishing effect during training, based on somatosensory perception. But can these four individual DANs also process other sensory modalities? Utilizing electron microscopy-based reconstruction, all first and second order input neurons of the MB, consisting of a complete set of 102 neuron pairs, were examined (Eschbach et al 2020, Winding et al 2023). The analysis identified numerous afferent sensory inputs via 20 pairs of feedforward neurons (FFNs), which can now be integrated with the published wiring diagram of all chemosensory input neurons (Miroschnikow et al 2018). This enables a more comprehensive understanding of the distinct modes of action of the four DL1 DANs in relation to the processing of different sensory modalities.

Consequently, we have investigated the role of the four DL1 DANs in the larval brain regarding their involvement in encoding a salt-dependent aversive teaching signal at the single-cell level. Utilizing behavioral assays, fluorescence microscopy, calcium imaging experiments, and connectome analyses, we have established a more comprehensive description of the coding space of the DANs at the single-cell level. Our investigation unveils a cellular specialization among the four DL1 DANs, where individual DANs code for different but partially overlapping aspects of the aversive teaching signal

Results

The larval dopaminergic system is functionally bipartite to convey positive and negative teaching signals

There are eight DANs in the larval brain that innervate the mushroom body that can be anatomically subdivided into four pPAM cluster and four DL1 cluster neurons (Figure 1A-C and Figure 1- figure supplement 1A-C). Three of the four pPAM cluster DANs can be labeled with the driver line R58E02-Gal4 (Figure 1- figure supplement 1A-E) (Rohwedder et al 2016). Pairing an odor presentation with simultaneous optogenetic blue light activation of R58E02 positive DANs using a mutant microbial-type rhodopsin ChR2XXL, which depolarizes neurons artificially (Dawydow et al 2014), establishes an appetitive olfactory memory (Figure 1- figure supplement 1F and G) (Rohwedder et al 2016). Please be aware that the activation of blue light itself does not cause any disruptive behavioral side effects, as it does not independently induce memory formation. In line with this finding, it was shown that ablation or inhibition of R58E02 positive DANs strongly reduces appetitive olfactory memory (Rohwedder et al 2016) but has no functional significance for aversive olfactory memory (Figure 1- figure supplement 1H and I) (Rohwedder et al 2016). Therefore, it can be concluded that pPAM cluster DANs are not involved in the encoding of aversive teaching signals. How are aversive teaching signals encoded?

The larval dopaminergic system is subdivided in two functionally distinct clusters.

(A) The larval dopaminergic neurons (DANs) can be anatomically subdivided into the primary protocerebral anterior medial (pPAM) and dorsolateral 1 (DL1) cluster based on their cell body position. (B) The DL1 cluster (cell bodies in green and purple) consists of four DANs providing input to the c, d, g, and f compartment of the vertical lobe, peduncle and lateral appendix of the MB (in grey). (C) Schematic representation of the larval MB using single letter acronyms based on (Saumweber et al 2018) to indicate compartment innervation by MB input and output neurons. DL1 DANs specifically innervate the compartments c, d, f and g in the vertical lobe, lateral appendix and peduncle, whereas pPAM DANs innervate the shaft (h) and medial lobe (I, j, k) of the MB. (D,E) Four DL1 DANs (DAN-c1, DAN-d1, DAN-g1, and DAN-f1) are included in the expression pattern of the TH-GAL4 driver line. However, expression of a UAS-mCD8::GFP reporter via TH-GAL4 labels many more neurons (in green, anti-GFP) throughout the entire CNS (in red and blue, anti-N-cadherin and anti-discs large), in total about 100 neurons. (F) To test whether optogenetic activation of the DL1 DANs is sufficient to substitute for a punishment, we used the TH-GAL4 driver in combination with UAS-ChRXXL. Experimental and control larvae were trained by simultaneously presenting and odor and blue light and thus artificial activation of DL1 DANs, whereas a second odor was presented in darkness. Only larvae of the experimental genotype (p > 0.05, N=15), but not of the two genetic controls (both p < 0.05, for each group N=15), retrieve an aversive associative olfactory memory. The same result was seen after one or three training trials (F and G, respectively). (H, I) To test for the acute function of the DL1 DANs in aversive associative olfactory memory, we expressed GtACR2 via the TH-GAL4 driver. Acute optogenetic inhibition of synaptic output from DL1 and other DANs reduced odor high salt memory. Experimental larvae raised on supplemented food (0.5 mM all-trans-retinal, ATR) and trained in blue light performed on a lower level than control animals kept on standard food (p < 0.05, for each group N=15). A memory impairment was seen after one and three training trials (G and H, respectively). All behavioral data is shown as box-plots. Differences between groups are highlighted by horizontal lines between them. Performance indices different from random distribution are indicated below each box-plot. The precise sample size is given below each box-plot. n.s. p > 0.05; * p < 0.05. Scale bars: in C 50 µm and in D 25 µm. The source data and results of all statistical tests are documented in the Supplemental Information file.

The dopaminergic pPAM cluster encodes for a teaching signal in larvae.

(A, B) The pPAM cluster (cell bodies in blue) consists of four DANs providing input to the h, i, j, and k compartments of the medial lobe of the mushroom body (MB; in grey) – the larval memory center. (C) Schematic representation of the larval MB using single letter acronyms based on (Saumweber et al 2018) to indicate compartment innervation by MB input and output neurons. DL1 DANs specifically innervate the compartments c, d, f and g in the vertical lobe, lateral appendix and peduncle, whereas pPAM DANs innervate the shaft (h) and medial lobe (I, j, k) of the MB. (D, E) Three pPAM DANs (DAN-h1, DAN-i1, and DAN-j1) are specifically labeled via the R58E02 driver line. Expression of a UAS-GFP reporter via R58E02 labels only these three DANs (in green, anti-GFP) throughout the entire CNS (in red and blue, anti-N-cadherin and anti-discs large) besides one additional cell that does not innervate the MB (asterisk in E). (F) To test whether optogenetic activation of the pPAM DANs is sufficient to substitute for a reward, we used the R58E02 driver in combination with UAS-ChRXXL to express a modified form of Channelrhodopsin2. Experimental and effector control larvae were trained via presenting and odor stimulus together with blue light stimulation and thus artificial activation of pPAM DANs, whereas a second odor was presented in darkness. Furthermore, a genetic control group received the same training protocol without blue light stimulation. Only larvae of the experimental genotype (p > 0.05), but not of the two genetic controls (both p < 0.05, for each group N=15), retrieve an appetitive associative olfactory memory. This shows that activation of the pPAM DANs is sufficient to mediate a reward teaching signal. The same effect was seen applying one or three training trials (F and G, respectively). (H, I) To test for the acute function of the pPAM DANs in aversive associative olfactory memory, we expressed GtACR2, an anion-conducting, light-gated channel via the R58E02 driver. Acute optogenetic inhibition of synaptic output from pPAM DANs during training did not change odor high-salt memory. Experimental larvae kept on supplemented food with 0.5 mM all-trans-retinal (ATR) or genetic controls, maintained on standard food performed at the same level (p > 0.05, for each group N=15). Intact odor high-salt memory was seen after one or three training trials (H and I, respectively). All behavioral data is shown as box-plots. Differences between groups are highlighted by horizontal lines between them. Performance indices different from random distribution are indicated below each box-plot. The precise sample size of each group is given below each box-plot. n.s. p > 0.05; * p < 0.05. Scale bars: (D) 50 µm, (E) 25 µm. The source data and results of all statistical tests are documented in the Supplemental Information file.

The driver line TH-Gal4 can be used to label the four DANs of the DL1 cluster (Figure 1A-E), as well as most other DANs, except for a small subset of DANs, such as pPAM DANs and neurons in the subesophageal zone (Figure 1D and E) (Selcho et al 2009). By coupling odor presentation with simultaneous optogenetic activation of TH-Gal4 positive DANs through ChR2XXL, an aversive olfactory memory can be established (Schroll et al 2006), which was observed after both one (Figure 1F) and three (Figure 1G) odor-DAN activation pairings. Consistent with these findings, aversive associative memory formed after one or three training cycles of odor high salt pairings were significantly reduced when DANs activity was inhibited optogenetically via GtACR2 (Figure 1H and I), a light-gated channel from the algae Guillardia theta (Mohammad et al 2017) (Please refer to the discussion regarding the partial reduction of the memory). In these experiments salt has to be added during the test to initiate the recall of the aversive memory (indicated as red test plates, Figure 1F-I) (Gerber & Hendel 2006, Schleyer et al 2015, Schleyer et al 2011). Previous studies have shown that the larva only recalls aversive odor-high salt memory when tested in the presence of the teaching signal. The prevailing explanation suggests that the larva engages in memory-based search during the test, but only if the memory content can be used in a beneficial way (Gerber & Hendel 2006, Schleyer et al 2015, Schleyer et al 2011).

From our results, we infer a division of labor at the level of DANs toward the MB. On the one hand, pPAM neurons mediate a dopaminergic appetitive teaching signal, and on the other hand, DL1 neurons convey a dopaminergic aversive teaching signal. Nevertheless, the findings related to DL1 DANs necessitate additional validation as the TH-Gal4 driver line used in the experiment labels approximately 120 cells, thus limiting the ability to draw cell-specific conclusions.

Anatomical single-cell analysis of DL1 DANs

In an earlier investigation, we conducted a screening of several thousand Gal4 lines in order to construct a high-resolution atlas of the mushroom body at the cellular level via light microscopy (Saumweber et al 2018). A subsequent reconstruction of the wiring diagram of the entire MB at the electron microscopic level encompassed not only individual neurons but all synaptic connections (Eichler et al 2017). Utilizing this anatomical work, we identified a total of 102 Gal4 lines featuring MB extrinsic neuron expression patterns. These lines were subjected to intersectional strategies in order to restrict expression patterns to single neurons (Eschbach et al 2020, Saumweber et al 2018). For the present study, we employed eight split-Gal4 lines chosen from this set, as they contained only one or two DANs of the DL1 cluster (Figure 2).

Anatomical single cell analysis of DL1 DAN specific split-GAL4 driver lines.

(A) The larval MB is organized into 11 compartments: CX calyx; IP and LP intermediate and lower peduncle; LA lateral appendix; UVL, IVL, and LVL upper, intermediate, and lower vertical lobe; SHA, UT, IT, LT shaft as well as upper, intermediate, and lower toe of the medial lobe. Single-letter synonyms of compartment names are given as “a–k”. These letters are used to indicate compartment innervation by the MB input and output neurons (Saumweber et al 2018). DL1 cluster DANs are DAN-c1, DAN-d1, DAN-g1, and DAN-f1 that innervate the respective four different compartments of the MB. (B-J) Individual split-Gal4 driver lines were crossed with the reporter strain UAS-mCD8::GFP;mb247-lexA,lexAop-mRFP. Third instar larval brains were dissected, fixed and mounted to visualize the fluorescent reporter signal labeling the MB (mb247-lexA,lexAop-mRFP shown in magenta) and the respective DAN pattern (GFP shown in green). (B-E) SS02160 (DAN-c1), MB328B (DAN-d1), SS02180 (DAN-f1), SS01716 (DAN-g1) each specifically label a single DL1 DAN (cell bodies are highlighted by white arrowheads). (F–K) Two neurons can be seen in MB065B, SS01702, and MB054B split-Gal4 that express in DAN-c1/DAN-f1, DAN-c1/MBIN-e1, and DAN-f1/DAN-g1. Please note that MB065B shows strong expression DAN-f1 but weaker staining in DAN-c1. (H–K) MB054B showed reliable strong expression in DAN-f1 and DAN-g1. In some brains a third weak cell body was visible right next to the other two DANs (I; GFP channel inverted and shown in black; J; cell body highlighted with gray arrowhead). Due to the low expression level we were not able to identify this cell given that only the g and f compartment of the MB were innervated (H). (K) Analysis of the entire brain via native fluorescence expression of GFP (green) and n-Syb (magenta) did not reveal additional cells for MB054B split-Gal4. Scale bars: (B-J) 20 µm, (K) 50µm.

Ablation of individual DL DANs does not impair aversive olfactory memory reinforced by high salt.

(A) The larval MB is organized into 11 compartments. Letters from a-k are used to indicate compartment innervation by the MB input and output neurons. (B, C) Individual split-Gal4 driver lines were crossed with the reporter strain UAS-mCD8::GFP;mb247-lexA,lexAop-mRFP. Third instar larval brains were dissected, fixed and mounted to visualize the fluorescent reporter signal labeling the MB (mb247-lexA,lexAop-mRFP shown in magenta) and the respective DAN pattern (GFP shown in green). (B) MB143B labels specifically DAN-d1. (C) MB145B labels DAN-f1 and very weakly several additional cells. (D, E) Aversive associative performance indices are shown for tests immediately after classical odor-high salt conditioning using one training trial. A schematic overview of the conditioning protocol is shown at the left. Neither DAN-d1 nor DAN-f1 function is required. Even when tested using two alternative split-Gal4 lines MB143B and MB145B. (F-G) The same result can also be seen after three cycle odor-high salt conditioning. A schematic overview of the conditioning protocol is shown at the left. In summary, in all four behavioral experiments the aversive associative performance indices of the experimental groups (N=15-20) were not decreased compared to the two (in all cases p > 0.05, N=15-25) or at least one genetic control group (in F, p < 0.05 comparing UAS control and experimental larvae). All behavioral data is shown as box-plots. Please note that the same data sets for the effector control were plotted in subfigures D and E (one cycle) and F and G (three cycle), as respective behavioral experiments have been conducted in parallel. Differences between groups are highlighted by horizontal lines between them. Performance indices different from random distribution are indicated below each box-plot. The precise sample size of each group is given below each box-plot. n.s. p > 0.05; * p < 0.05. The source data and results of all statistical tests are documented in the Supplemental Information file.

Split-Gal4 MB054B covers three dopaminergic neurons in the DL1 cluster.

(A-F) Split-Gal4 MB054B driver strain was crossed to UAS-mCD8::GFP effector strain. (A) A z projection of a larval central brain is shown using anti-GFP to label expression pattern of Split-Gal4 MB054B (cyan), anti-TH to label TH-positive cells (yellow) and anti-N-cadherin to visualize the neuropil (magenta). The DL1 cluster of the left hemisphere is highlighted. (B,C) Single channel visualization of all TH-positive cells (B) and GFP expressing cells (C) of the crossed Split-Gal4 MB054B driver line to effector UAS-mCD8::GFP within the larval central brain. (D) A detailed z projection of the DL1 cluster is shown using anti-GFP to label expression pattern of Split-Gal4 MB054B (cyan) and anti-TH to label TH-positive cells (yellow). Co-labeling of anti-TH and anti-GFP is highlighted by arrowheads. (E,F) Single channel visualization of all TH-positive cells (E) and GFP expressing cells (F) of the Split-Gal4 MB054B driver line crossed to effector UAS-mCD8::GFP within the DL1 cluster. Scale bars: (A-C) 50 µm, (D-F) 10 µm.

DAN-c1 was specifically labeled by SS02160 (Figure 2B), DAN-d1 by MB328B and MB143B (Figure 2C and Figure 2- figure supplement 1B), DAN-f1 by SS02180 and MB145B (Figure 2D and Figure 2- figure supplement 1C) and DAN-g1 by SS01716 (Figure 2E). In addition, we used the split-Gal4 lines MB065B (Figure 2F), SS01702 (Figure 2G) and MB054B (Figure H-K) that specifically labeled DAN-c1/DAN-f1, DAN-c1/MBIN-e1 or DAN-f1/DAN-g1, respectively. For our evaluation we crossed each split-Gal4 line with the reporter strain UAS-mCD8::GFP;mb247-LexA,lexAop-mRFP (Burke et al 2012) that allowed us to monitor the expression pattern (green) in the background of a MB specific reference staining (magenta). Our results confirm the reported specificity even despite the use of different reporter lines and lacking antibody staining (Eschbach et al 2020), as we analyzed only endogenous GFP expression using a more cautious and faster whole mount staining protocol (see Material and Methods). In addition, we also verified that the DANs labeled via MB054B are indeed dopaminergic as the GFP expression overlaps with an anti-TH antibody staining (Figure 2-figure supplement 2). Having established anatomical validity, we subsequently employed these lines to explore the physiological response of each DAN within the DL1 cluster to sugar and salt stimuli. For the purpose of analyzing DAN-d1 and DAN-f1, we limited our investigation to the split-Gal4 lines MB328B and SS02180 as these lines displayed more robust expression levels.

Physiological single-cell analysis of DL1 DANs

In order to investigate how high salt teaching signals are represented at the level of individual DL1 DANs, we conducted calcium imaging experiments using a microfluidic chip on intact and immobilized larvae, as previously described (Si et al 2019). We expressed GCaMP6m under the control of four different split-Gal4 lines specific to DL1 DANs (SS02160 (DAN-c1), MB328B (DAN-d1), SS02180 (DAN-f1), and SS01716 (DAN-g1)), as well as the pPAM specific R58E02-Gal4, and exposed the larvae to solutions containing 100 mM salt, 1 M salt, or 500 mM fructose. Our results showed that only DAN-c1 responded to 100 mM salt (Figure 3A). While DAN-c1, DAN-d1, and DAN-g1 responded to 1 M salt (Figure 3A, B, and D). Conversely, pPAM DANs showed a reduced signal in response to 1 M salt (Figure 3E). Both DAN-c1 and pPAM DANs showed a calcium increase in response to 500 mM fructose (Figure 3A and E). Although DAN-f1 did not respond to either salt concentration, a reduction in the calcium signal was seen for 500 mM fructose (Figure 3C). Overall, these results provide support for the functional division of the larval dopaminergic system at the physiological level, where most DL1 neurons are activated by high salt concentrations, with one exception inhibited by fructose. In contrast, pPAM cluster neurons display the opposite response, being inhibited by high salt concentrations and strongly activated by fructose.

Calcium responses of DANs to gustatory stimulation.

Four different split-Gal4 lines and the R58E02 driver line were crossed with UAS-GCaMP6m to express a calcium reporter in DANs. The responses of each DAN towards 100 mM NaCl, 1 M NaCl, and 500 mM fructose was tested in intact larvae using a microfluidic chip-based setup. (A) Calcium responses in DAN-c1 (N= 5) were induced by gustatory stimulation with 100 mM NaCl (orange), 1 M NaCl (red) and 500 mM fructose (green, for all p < 0.05). (B) DAN-d1 (N=5) calcium responses were only seen after 1 M NaCl stimulation (red, p < 0.05), but not after 100 mM NaCl (orange, p > 0.05) and 500 mM fructose stimulation (green, p > 0.05). (C) Stimulation with 100 mM (orange, p > 0.05) and 1 M NaCl (red, p > 0.05) did not induce calcium responses in DAN-f1 (N=6). However, stimulation with 500 mM fructose reduced the obtained calcium signal (green, p < 0.05). (D) 1 M NaCl (red, p < 0.05), but not 100 mM NaCl (orange, p > 0.05) and 500 mM fructose (green, p > 0.05), induced a calcium response in DAN-g1 (N=7). (E) pPAM DANs calcium responses (N=5) were only seen after 500 mM fructose stimulation (green, p < 0.05). Stimulation with low (orange) and high salt concentrations (red) did not increase calcium signals, however both reduced pPAM activity (p < 0.05). Each graph shows the mean calcium signal plotted as the relative response strength ΔF/F and the related standard error of the mean on the y axis. The time in seconds is given below each graph on the x axis. The grey box indicates the duration of the stimulus application. The sample size of each group (N=5-7) is given above each row. n.s. p > 0.05; * p < 0.05. The source data and results of all statistical tests are documented in the Supplemental Information file.

Behavioral single-cell analysis of DL1 DANs

Next, we analyzed the role of single DL1 DANs in aversive olfactory learning (Figure 2- figure supplement 1D-G, Figure 4- figure supplement 1-4). We employed split-Gal4 lines SS02160 (DAN-c1), MB328B (DAN-d1), SS02180 (DAN-f1), and SS01716 (DAN-g1) to express the apoptosis proteins Hid and Reaper, resulting in the ablation of the respective individual DL1 DANs. The ablated animals were subjected to the aversive odor-high salt memory assay, using one (Figure 4- figure supplement 1A-D) and three training cycles (Figure 4- figure supplement 1E-H). We utilized two different training protocols to differentiate between the effects on short-term memory (STM) and anesthesia-resistant memory (ARM), as previous studies have demonstrated that larvae exhibit both after one training cycle, whereas only ARM is present after three training cycles (Widmann et al 2016). However, we found that the ablation of individual DL1 DANs had no effect on aversive olfactory memories in any of the eight experiments conducted (Figure 4- figure supplement 1A-H). For DAN-d1 and DAN-f1 we verified these results by using the additional split-Gal4 lines MB143B and MB145B (Figure 2- figure supplement 1D-G). In addition, ablation of DAN-d1, DAN-f1 and DAN-g1 individually did not have any impact on odor-quinine and odor-fructose memories established after one cycle training (Figure 4-figure supplement 2 and 3). Moreover, MB input neurons that are not dopaminergic (OAN-a1, a2, MBIN-b1, b2, and OAN-e1) were also found to have no relevant effect (Figure 4-figure supplement 4). These results suggest that although DL1 DANs respond to high salt concentrations (Figure 3), they are not individually necessary in ablation experiments for mediating the aversive teaching signal. It appears that the aversive teaching signal is distributed among multiple cells, indicating the presence of some redundancy. However, it is possible that there are developmental compensatory effects for DL1 cluster DANs, allowing the remaining three neurons (or other cells) to take over the function of the ablated one. This hypothesis was analyzed later in optogenetic inhibition experiments.

Ablation of DAN-f1 and DAN-g1 together impairs aversive olfactory memory.

In all panels associative performance indices are shown for tests immediately after classical conditioning. In the upper panels (A-C) larvae are trained once by pairing an olfactory stimulus with high salt punishment, whereas in the lower panels (D-F) three training cycles were applied. Schematic overviews for both conditioning protocols are shown on the left. The three different DL1 DAN specific split-Gal4 driver MB054B, MB065B, and SS01702 that each label two neurons were crossed to the effector UAS-hid,rpr to induce apoptosis (A-F). (A) With MB054B used as driver strain to ablate the DL1 DAN combination DAN-f1/DAN-g1, the aversive associative performance index of the experimental group was decreased compared to both controls (p < 0.05, N=15 for each group). (B-F) In all other experiments ablation of different DL1 DAN combinations did not reveal a phenotype. (in all experiments at least one or even both control groups are compared to the experimental group p > 0.05, for each group N=15-25). Please note, that this also includes MB054B crossed with UAS-hid,rpr tested after three training trials (D). (G) To verify the memory phenotype of MB054B crossed with UAS-hid,rpr tested after one trail conditioning, we repeated the experiment using the odor pair hexyl acetate (HA) and benzaldehyde (BA). Again, experimental larvae tested after one trial learning showed a robust decrease in odor-high salt memory when compared to both genetic control groups (p < 0.05, for each group N=15). The memory phenotype was not seen after three training trials (J, p > 0.05, for each group N=15). (H, K) With MB054B used as driver strain to ablate the DL1 DAN combination DAN-f1/DAN-g1, aversive odor-quinine memory was not impaired after one or three cycle conditioning (p > 0.05 when comparing experimental and control groups, for each group N=15). (I, L) Similarly, appetitive odor-fructose learning after one and three cycle conditioning was not impaired when ablating DAN-f1/DAN-g1 (p > 0.05 when comparing experimental and control groups, for each group N=15). All behavioral data is shown as box-plots. Please note that the same data set for the effector control was plotted in subfigures E and F as respective behavioral experiments have been conducted in parallel. Differences between groups are highlighted by horizontal lines between them. Performance indices different from random distribution are indicated below each box-plot. The precise sample size of each group is given below each box-plot. n.s. p > 0.05; * p < 0.05. The source data and results of all statistical tests are documented in the Supplemental Information file.

Ablation of individual DANs does not impair aversive olfactory memory.

In all panels aversive associative performance indices are shown for tests immediately after odor-high salt classical conditioning. In the upper panels (A-D) larvae are trained once, whereas in the lower panels (E-H) three training cycles were applied. Schematic overviews for both conditioning protocols are shown on the left. The four different DL1 DAN specific split-Gal4 driver strains SS02160, MB328B, SS02180, and SS01716 were crossed to the effector UAS-hid,rpr to induce apoptosis (A-H). Cell-specific ablation of distinct DL1 DANs did not impair odor-high salt memories after either one or three cycle training. In no case were associative performance indices upon DL1 DAN ablation lower than in both genetic controls (A-H, in all experiments at least one or even both control groups are compared to the experimental group p > 0.05, for each group N= 15-20 for one cycle training, and N= 15-25 for three cycle training). All behavioral data is shown as box-plots. As the respective behavioral experiments in subfigures A-D and E-H were conducted in parallel, the same data sets for the corresponding effector control were plotted. Differences between groups are highlighted by horizontal lines between them. Performance indices different from random distribution are indicated below each box-plot. The exact sample size of each group is given below each box-plot. n.s. p > 0.05; * p < 0.05. The source data and results of all statistical tests are documented in Supplemental Information file.

Ablation of individual DL DANs does not impair aversive olfactory memory reinforced by bitter quinine.

In all panels aversive associative performance indices are shown for tests immediately after classical odor-quinine conditioning using one training trial. A schematic overview of the conditioning protocol is shown to the left of (A). (A-C) The three different DL1 DAN split-Gal4 driver MB328B, SS02180, and SS01716 that specifically label DAN-d1, DAN-f1, and DAN-g1, respectively, are crossed with the effector UAS-hid,rpr to induce apoptosis. In all three experiments the aversive associative performance indices of the experimental groups (N= 15) were not decreased compared to the two control groups (in all cases p > 0.05, for each group N=15). All behavioral data is shown as box-plots. Please note that the same respective data set for the effector control was plotted in all subfigures, as these behavioral experiments were conducted in parallel. Differences between groups are highlighted by horizontal lines between them. Performance indices different from random distribution are indicated below each box-plot. The exact sample size of each group is given below each box-plot. n.s. p > 0.05; * p < 0.05. The source data and results of all statistical tests are documented in Supplemental Information file.

Ablation of individual DL DANs does not impair appetitive olfactory memory.

In all panels appetitive associative performance indices are shown for tests immediately after classical odor-fructose conditioning using one training trial. A schematic overview of the conditioning protocol is shown to the left of (A). (A-C) The three different DL1 DAN split-Gal4 driver MB328B, SS02180, and SS01716 that specifically label DAN-d1, DAN-f1, and DAN-g1, respectively, are crossed with the effector UAS-hid,rpr to induce apoptosis. In all three experiments the appetitive associative performance indices of the experimental groups (N=15) were not decreased compared to the two related control groups (in all cases p > 0.05, for each group N=15). All behavioral data is shown as box-plots. Differences between groups are highlighted by horizontal lines between them. Performance indices different from random distribution are indicated below each box-plot. The precise sample size of each group is given below each box-plot. n.s. p > 0.05; * p < 0.05. The source data and results of all statistical tests are documented in the Supplemental Information file.

Functional analysis of mushroom body input neurons that are not dopaminergic.

In panels (A-C) larvae are trained once by pairing an olfactory stimulus with high salt punishment, whereas in panel (D-F) three training cycles were applied. Schematic overviews for both conditioning protocols are shown to the left of (A) and (D). The three different mushroom body input split-Gal4 driver SS24765, SS21716, and SS01958 that specifically label OAN-a1,a2, MBIN-b1,b2, and OAN-e1 are crossed with the effector UAS-hid,rpr to induce apoptosis. Ablation of any of the three types of input neurons did not impair odor-high salt memory regardless of whether larvae were trained once or three times (A-C and D-F). In all cases the experimental groups (N=15) performed on the same level as both (A-C and F, in all cases p > 0.05, N=15-20) or at least one genetic control (D, E, Gal4 control p > 0.05 and UAS-hid,rpr control p < 0.05, N=15-25). All behavioral data is shown as box-plots. As the respective behavioral experiments in subfigures A-C for one cycle and D and F for three cycles were conducted in parallel, the same data sets for the corresponding effector control were plotted. Differences between groups are highlighted by horizontal lines between them. Performance indices different from random distribution are indicated below each box-plot. The precise sample size of each group is given below each box-plot. n.s. p > 0.05; * p < 0.05. (G-I) Individual split-Gal4 driver lines were crossed with the reporter strain UAS-mCD8::GFP;mb247-lexA,lexAop-mRFP. Third instar larval brains were dissected, fixed and mounted to visualize the fluorescent reporter signal labeling the MB (mb247-lexA,lexAop-mRFP shown in magenta) and the respective neuronal pattern (GFP shown in green). SS24765 (OAN-a1,a2), SS21716 (MBIN-b1,b2), and SS01958 (OAN-e1) each specifically label a specific input neuron pair (G, H) or a single input neuron (I). Scale bars: 20 µm. The source data and results of all statistical tests are documented in the Supplemental Information file.

Sensory acuity tests.

In panels (A-E) naïve larval preferences towards different chemosensory stimuli are shown. Olfactory preference of MB054B split Gal4 crossed to UAS-hid,rpr to ablate DAN-f1/DAN-g1 are shown in (A-D). For all the tested odor stimuli, amyl acetate diluted 1:250 in paraffin oil, pure benzaldehyde, hexyl acetate diluted 1:100 in paraffin oil, and benzaldehyde diluted 1:100 in paraffin oil, no difference between the experimental group (N=15) and the two genetic control groups was seen (for all cases p > 0.05, for each group N=15). (E) Gustatory high salt preference of MB054B split-Gal4 crossed to UAS-hid,rpr to ablate DAN-f1/DAN-g1 (N=15) was not reduced but rather increased in their avoidance of 1.5 M high salt concentration compared to both genetic controls (both p < 0.05, for each group N=15). All behavioral data is shown as box-plots. Differences between groups are highlighted by horizontal lines between them. Performance indices different from random distribution are indicated below each box-plot. The precise sample size of each group is given below each box-plot. n.s. p > 0.05; * p < 0.05. The source data and results of all statistical tests are documented in the Supplemental Information file.

Anatomical validation of effector UAS-hid,rpr functionality to ablate neurons of MB054B Split-Gal4 in the DL1 cluster.

To confirm the functionality of the effector UAS-hid,rpr, anatomical examination of the single effector line UAS-hid,rpr (A-C’’’), the Split-Gal4 driver line MB054B (D-F’’’), as well as the cross of both (G-I’’’) were examined. (A, D, G) A z projection of a larval central brain of the respective strain or cross is shown using anti-TH to label TH positive cells (yellow) and anti-N-cadherin to visualize the neuropil (magenta). (B, E, H) Single channel visualization of all TH-positive cells (black) of the respective larval brains shown in (A, D, G). The DL1 cluster is highlighted by violet dashed lines. (C) Detailed z projection of the left hemispheric DL1 cluster of effector line UAS-hid,rpr is shown as single channel (anti-TH, black). Cell bodies of the DL1 cluster neurons are highlighted by violet asterisk. (C’-C’’’) Partial projection of the z-stack shown in (C) to distinguish single cell bodies. Anatomical analysis revealed seven TH-positive cell bodies in the DL1 cluster of UAS-hid,rpr, highlighted by violet arrowheads. (F) Detailed z projection of the left hemispheric DL1 cluster of Split-Gal4 line MB054B is shown as single channel (anti-TH, black). Cell bodies of the DL1 cluster neurons are highlighted by violet asterisk. (F’-F’’’) Partial projection of the z-stack shown in (F) to distinguish single cell bodies. Anatomical analysis revealed seven TH-positive cell bodies in the DL1 cluster of MB054B, highlighted by violet arrowheads. (I) Detailed z projection of the left hemispheric DL1 cluster of the crossed Split-Gal4 driver strain MB054B to UAS-hid,rpr is shown as single channel (anti-TH, black). Cell bodies of the DL1 cluster neurons are highlighted by violet asterisk. (I’-I’’’) Partial projection of the z-stack shown in (I) to distinguish single cell bodies. Crossing Split-Gal4 MB054B to effector line UAS-hid, rpr led to ablation of the respective MB054B cell bodies (see also Figure 2H-K), leaving only four TH-positive cell bodies (highlighted by violet arrowheads) in the DL1 cluster. Scale bars: 50 µm (A,B, D,E, G,H) and 20 µm (C-C’’’, F-F’’’, I-I’’’).

The combination of DAN-f1 and DAN-g1 is required to establish aversive odor-high saltmemories

In order to investigate a possible redundancy among DANs, we proceeded by using split-Gal4 lines MB054B (DAN-f1/DAN-g1), MB065B (DAN-c1/DAN-f1), and SS01702 (DAN-c1/MBIN-e1) to express apoptosis proteins Hid and Reaper. This approach selectively ablated either two DL1 DANs or one DL1 DAN and another MB input neuron with unknown neurotransmitter identity. It should be noted that no additional split-Gal4 lines are presently available for further DL1 DAN combinations or the complete DL1 cluster. We then subjected the larvae to the aversive odor-high salt memory assay, using one (Figure 4A-C) and three (Figure 4D-F) training cycles. Only larvae with ablated DAN-f1/DAN-g1 exhibited significantly reduced aversive olfactory memory after one cycle training (Figure 4A). No memory reduction was observed when ablating DAN-c1/DAN-f1 or DAN-c1/MBIN-e1 (Figure 4B and C). Ablation of all three sets of neurons, including MB054B, had no effect on aversive olfactory memory after three training cycles (Figure 4D-F). To ascertain the independence of the DAN-f1/DAN-g1 neuron-mediated impairment of aversive olfactory memory after one cycle training from the specific odors used, hexyl acetate and benzaldehyde were employed instead of amyl acetate and benzaldehyde. The results of these experiments confirmed the previous findings, as ablation of DAN-f1/DAN-g1 neurons significantly reduced aversive odor-high salt memory after one cycle training (Figure 4G), but not after three cycles (Figure 4J). Innate olfactory and gustatory behavior was not impaired by DAN-f1/DAN-g1 ablation (Figure 4-figure supplement 5).

Subsequently, we investigated the selectivity of DAN-f1/DAN-g1 neurons concerning the nature of the aversive teaching stimulus. To this end, we utilized quinine as an aversive teaching signal, a stimulus that has been demonstrated to be learnable by the larva, as evidenced by previous studies (Apostolopoulou et al 2014, El-Keredy et al 2012). Removal of the DAN-f1/DAN-g1 neurons using the split-Gal4 line MB054B had no effect on aversive odor-quinine memory independent of the number of training cycles (Figure 4H and K). Furthermore, ablation of DAN-f1/DAN-g1 did not alter appetitive odor-fructose memories after one or three training cycles (Figure 4I and L). Therefore, the DL1 DAN-f1/DAN-g1 combination seems to be specifically required for aversive odor-high salt memory. This is further supported by anatomical data that verified the effectiveness of Hid and Reaper using the MB054B split-Gal4 line (Figure 4- figure supplement 6).

Individual DL1 DANs can instruct an aversive memory recalled on a salt plate

In order to determine if the activation of individual DL1 DANs signals aspects of the natural high salt punishment, we conducted a one training trial learning experiment using ChR2XXL to artificially activate DAN-c1, DAN-d1, DAN-f1, DAN-g1, and the combination of DAN-f1/DAN-g1. We used split-Gal4 lines SS02160, MB328B, SS02180, SS01716, and MB054B, respectively (Figure 5). Simultaneously activating DAN-f1/DAN-g1 via ChR2XXL during an odor presentation establishes an aversive olfactory memory (Figure 5E and F). These results are comparable to those observed for TH-Gal4 (as shown in Figure 1F and G). Additionally, this memory is observed after one (Figure 5E) and three training cycles (Figure 5F). Furthermore, single-cell analysis indicates that individual activation of DAN-f1 and DAN-g1, but not of DAN-d1 and DAN-c1 (please note that experimental animals are different from zero and significantly different to one control), can induce an aversive odor memory (Figure 5A-D). In summary, these results may suggest that DAN-f1 and DAN-g1 encode aspects of the natural aversive high salt teaching signal under the conditions that we tested.

Optogenetic DL1 DAN activity can substitute for salt punishment.

In all panels associative performance indices are shown for tests immediately after classical conditioning. In panels (A-E) larvae are trained once by pairing an olfactory stimulus with artificial blue light activation, whereas in panel (F) three training cycles were applied. Schematic overviews for both conditioning protocols are shown to the left of (A) and (F). (A-D) To test whether optogenetic activation of the individual DL DANs DAN-c1, DAN-d1, DAN-f1, and DAN-g1 is sufficient to substitute for a punishment, we used the split-Gal4 lines SS02160, MB328B, SS02180, and SS01716 in combination with UAS-ChR2XXL. (E, F) For simultaneous optogenetic activation of DAN-f1/DAN-g1 we used MB054B. Larvae of the experimental genotypes for DAN-c1, DAN-f1, DAN-g1 and the DAN-f1/DAN-g1 combination (for all p < 0.05, for each group N=15), but not for DAN-d1 (N=15) and all genetic controls (for all p > 0.05, for each group N=15), showed an aversive associative memory. The results imply that in the tested conditions, the punishment signal can be mediated by the artificial activation of all individual DL1 DANs, with the exception of DAN-d1. All behavioral data is shown as box-plots. Please note that the same data set for the effector control was plotted in subfigures B-E as respective behavioral experiments have been conducted in parallel. Differences between groups are highlighted by horizontal lines between them. Performance indices different from random distribution are indicated below each box-plot. The sample size of each group (N=15) is given below each box-plot. n.s. p > 0.05; * p < 0.05. The source data and results of all statistical tests are documented in the Supplemental Information file.

Individual DL1 DANs are acutely necessary to establish aversive odor-high salt memories

Next, we examined the extent to which the individual DL1 DANs are acutely required for the transmission of a functional aversive teaching signal. We have already shown that the combination of DAN-f1/DAN-g1 is required for aversive odor-high salt memory (Figure 4A and G). However, in these experiments DANs were removed during the entire development, which may have led to compensatory mechanisms or side effects.

To overcome this limitation, we left the individual DL1 DANs intact during embryonic and larval development and used GtACR2 to selectively inhibit DAN activity only during training, thus isolating the teaching signals from other DAN functions. Optogenetic inhibition of DAN-f1/DAN-g1 resulted in a reduced aversive odor-high salt memory. This was the case after one training cycle (Figure 6E) and three training cycles (Figure 6F). Thus, the learning deficit in this experiment was more pronounced than after ablation of these two DL1 DANs throughout development (Figure 4D). Optogenetic inhibition of individual DL1 DANs showed no effect on aversive odor-high salt memories (Figure 6A-C), except for DAN-g1 that led to a small but significant reduction (Figure 6D).

Optogenetic inhibition of DL1 DAN activity impairs aversive olfactory memory.

In all panels associative performance indices are shown for tests immediately after classical odor high salt conditioning. In panels (A-E) larvae are trained once by pairing an olfactory stimulus with an aversive high salt stimulation, whereas in panel (F) three training cycles were applied. Schematic overviews for both conditioning protocols are shown to the left of (A) and (F). (A-D) To test whether optogenetic inhibition of the individual DL DANs DAN-c1, DAN-d1, DAN-f1, and DAN-g1 during training impairs punishment signaling, we used the split-Gal4 lines SS02160, MB328B, SS02180, and SS01716 in combination with UAS-GtACR2 and blue light stimulation during the entire training phase. (E, F) For simultaneous optogenetic inhibition of DAN-f1/DAN-g1 we used MB054B. (A-C) Larvae with inhibited DAN-c1, DAN-d1, or DAN-f1 function during training showed no impairment of odor-high salt memory comparable to controls that were kept on standard food without supplemented all-trans-retinal (ATR, 0.5 mM) and received the same protocol (for all p > 0.05, N=15 for each group). (D-F) In contrast inhibition of DAN-g1 alone, or the combination of DAN-f1/DAN-g1 after single trial and three trial conditioning impaired odor-high salt memory compared to controls (for all p < 0.05, for each group N=15). This shows that DAN-g1 function is of central importance for signaling a salt punishment teaching signal. All behavioral data is shown as box-plots. Differences between groups are highlighted by horizontal lines between them. Performance indices different from random distribution are indicated below each box-plot. The precise sample size of each group is given below each box-plot. n.s. p > 0.05; * p < 0.05. The source data and results of all statistical tests are documented in the Supplemental Information file.

A synaptic reconstruction of all chemosensory inputs for each individual DL1 and pPAM DAN

Certainly, a comprehensive analysis from the sensory neurons responsible for salt perception, via the interneurons involved in salt information processing, up to the dopaminergic DL1 teaching neurons would be highly desirable to understand why and how the different DANs respond to salt. Regrettably, such an analysis is currently not feasible due to our limited understanding of salt perception and which sensory cells encode this crucial information (Liu et al 2003). Therefore, this question cannot be answered by the larval connectome. However, it is possible to analyze how qualitative differences between individual DL1 DANs (Figure 3-6) are represented in the connectome, since not all DANs code equally for the punishment teaching signal. Therefore, we compared the input wiring diagrams of the individual DANs of the pPAM and DL1 clusters by mapping the synaptic connections from sensory neurons - to interneurons - to individual DANs, using the full larval brain EM volume (Figure 7, Figure 7- figure supplement 1) (Eschbach et al 2020, Miroschnikow et al 2018, Winding et al 2023).

Interneurons and hub analysis of sensory to DAN pathways.

(A, B) Schematic graph representation of individual DL1 (A) and pPAM (B) DANs. The outer ring at the bottom of each scheme represents the sensory composition of neurons targeting sensory-to-DANs interneurons (orange circle). The type of sensory information is encoded by the respective color (ACa: anterior central sensory compartment, AVa: ventral anterior sensory compartment, ACal: lateral anterior central sensory compartment, ACp: posterior anterior central sensory compartment, ACpl: posterior-lateral anterior central sensory compartment, VM: ventromedical sensory compartment, TD CO2: tracheal dendritic neurons responding to CO2, ORNs: olfactory receptor neurons). DAN input neurons are shown that get no direct sensory input (grey circles). Individual DANs are shown in the middle of the scheme as black circles. They are connected to MB KC (purple circles), which in turn connect to mushroom body output neuron (MBONs; dark red circles). Arrows indicate the direction of the synaptic connection and its strength (coded by arrow thickness). Numbers in circles indicate number of neurons. The percentages indicate the proportion relative to the total input that the cell receives from the specific neuronal partners. For example, DAN-f1 receives 17% of its input from six different sensory-to-DAN interneurons (yellow circle), 17% from 19 other, non-sensory interneurons, 41% from 64 MB KCs, and 2% from two MBONs. (C) Dot plot showing the importance of interneurons acting as sensory to DAN hub. Dot size was calculated using the fraction of total input an interneuron receives from sensory neurons multiplied by the fraction of total input this interneuron gives to an DAN. Colored backgrounds of dots are highlighted in orange for the connections with a hub size of 0.001 or above. (D, E) Schematic of graph representation. The outer ring represents the sensory composition of neurons targeting upstream neurons of DANs. The type of sensory information is encoded by the respective color. Synaptic threshold for upstream interneurons of DANs = 3 and of upstream sensory neurons = 1. Line thickness to interneurons and targets represents the percentage of synaptic input. White or orange circles connected to the outer ring represent the interneuron layer. The inner ring represents individual target neurons, DL1 and pPAM DANs. The identity of each DAN and interneuron is given by the label in its related circle. (F) EM reconstruction of DL1 and pPAM DANs (grey) highlighting their presynaptic sites in red (for DL1 DANs) and green (for pPAM DANs). At the top a horizontal view of the brain is shown. At the bottom a frontal view of the brain is shown.

A sensory neuron to DAN map based on EM connectivity.

(A) Adjacency matrix of all neurons used in this study (no synaptic threshold) showing sensory compartments (magenta) to interneuron (yellow) to DL1 (red) and pPAM DANs (green). All synapses between shown neuron types were included. Top: Nine different sensory origins synapse onto upstream neurons (interneurons) of the DANs. Interneuron ID and neuron name is listed on the left and right of the matrix. From top to bottom the connectivity of 35 interneurons are shown that establish the complete single hop circuit between sensory neurons and DANs. Please note that most of the interneurons form pairs (15 paired and 5 unpaired interneurons). (B) Schematic of all sensory neurons that innervate the DL1 cluster (left) and the pPAM cluster (right). The top shows a frontal view, at the bottom a view from above along the longitudinal axis is given. The sensory neurons upstream of DANs are color coded based sensory origin, which is explained at the bottom of the figure. (C) Comparison of the similarity of DL1 and pPAM sensory input. Most of the sensory input is either specific for the DL1 cluster (57 neurons, magenta) or the pPAM cluster (61 neuron, green). Only a small number of sensory neurons (7 neurons, black) instructs both clusters. (D) Frontal and sagittal views are shown from top to bottom that highlight the distinct sets of interneurons connecting DL1 DANs (4 neurons, red) or pPAM DANs (16 neurons, green) with the sensory neurons.

We found that DAN-c1, DAN-d1, DAN-f1, and DAN-g1 (black circle) clearly differ in their downstream connections. They do however receive input from 1, 6, 6 or 11 sensory-to-DAN interneurons (orange circle), respectively (Figure 7A), which in turn means that these DL1 DANs are connected to 3, 42, 35, and 9 sensory neurons (numbers in colored circuits below). Please note that we restricted our analysis on direct sensory-to-DANs interneurons (single hop analysis) and did not expand the analysis onto two hop or three hop connections to keep the evaluation restricted to the shortest paths from sensory neurons to DANs. There is no direct input from sensory neurons onto DL1 DANs. Sensory input thus appears to be increasingly instructing DAN-c1, DAN-d1, DAN-f1, and DAN-g1. The analysis also shows that all four DL1 DANs receive information from all gustatory senses via these interneurons (coded by the yellow color for sensory composition). In addition, DAN-d1, as well as DAN-f1, receive information from olfactory and thermosensory neurons (Figure 7D). In general, less processed sensory input seems to be more important for DAN-f1 and DAN-g1 than for DAN-c1 and DAN-d1, as they get 17% and 11% of their entire input from these sensory-to-DAN interneurons in contrast to 1% and 5%, respectively. The four DL1 DANs show a conserved connection motif, as they get additional input from other upstream neurons (US) that are not directly linked to sensory neurons (grey circles; between 17% and 37%), as well as from MB KCs (purple circles; between 36% and 49%) and little MBON feedback (dark-red circle; between 1% and 2%). Examination of the different input types of DL1 DANs presented here, shows that DAN-f1 and DAN-g1 receive the highest proportion of sensory input relative to other DANs. Even by looking at the sensory-to-DAN interneurons individually, one can again see the familiar pattern of differences and partial overlap (Fig. 7D). DAN-d1, DAN-f1 und DAN-g1 all get input from interneuron pair 1 in contrast to DAN-c1. However again, DAN-f1 and DAN-g1 are most strongly connected to this neuronal pair (indicated by the thickness of the lines). DAN-f1 and DAN-g1 are therefore also the most similar in terms of their gustatory input (due to their connection via interneuron pair 1), despite all the clearly recognizable differences.

For the pPAM DANs we found the same general connectivity principles as for DL1 DANs regarding input from MBONs, KCs, and non-sensory upstream neurons (Fig 7B). Please note that the DAN-h1 neuron is not present at the early L1 EM brain volume and thus cannot be included in this evaluation (Figure 7B and E). In contrast, however, it is striking that pPAM DANs seem to get input from more interneurons connected to the sensory system (up to 16 for DAN-j1) signaling additional internal somatosensory and CO2 information. A closer inspection of the sensory-to-DAN interneurons showed that 27 of the 35 neurons are specifically innervating only DL1 or pPAM DANs (Figure 7C). Only 8 out of 35 interneurons connect to both clusters. Based on a calculated hub score sensory-to-DAN interneurons pair 1 (FB2IN-12) and pair 5 (FFN-21) were likely most significant for instructing DL1 DANs with sensory information. Whereas pairs 10 (FFN-17) and 11 (FFN-12) were – based on connectivity – most important in instructing DL1 and pPAM DANs. For the three pPAM DANs sensory-to-DAN interneuron pairs 16, 17, 18, 19 and 20 resulted in the highest hub scores (FFN-37, FFN-4, FFN-5, FFN-18 and FB-4).

The difference in cellular connectivity of DL1 and pPAM DANs is also reflected in different local input brain hubs: axo-dendritic synapses of sensory-to-DAN interneurons with DL1 DANs formed a characteristic hook-shape in the superior protocerebrum; in contrast, pPAM DANs get synaptic input from sensory-to-DAN interneurons in the anterior inferior protocerebrum (Figure 7F).

Discussion

A functional assignment for individual DANs of the DL1 cluster in their ability to encode a salt punishment teaching signal

The discovery of DL1 neurons as aversive teaching signal mediators in larval Drosophila (Eschbach et al 2020, Schroll et al 2006, Selcho et al 2009, Weiglein et al 2021) complements previous findings that demonstrate the sufficiency of four DANs from the pPAM cluster as an appetitive reinforcement signal in these animals (Rohwedder et al 2016). This dichotomy demonstrates an organizational principle shared by adult Drosophila and larvae (Aso et al 2012, Aso & Rubin 2016, Aso et al 2010, Burke et al 2012, Claridge-Chang et al 2009, Das et al 2014, Huetteroth et al 2015, Liu et al 2012, Schleyer et al 2020, Schroll et al 2006, Selcho et al 2009, Weiglein et al 2021, Yamagata et al 2015), despite a significant reduction in cell numbers. The principle appears to apply to mammals as well (Groessl et al 2018, Lammel et al 2012, Menegas et al 2018, Schultz 2015). But how do individual DL1 neurons contribute to the overall teaching signal? Is the signal evenly distributed across multiple cells, supporting the mass action hypothesis, or do individual DANs specifically encode for the salt stimulus (Rohwedder et al 2016)? Our results show that the four DL1 DANs perform this function in a mixture of both forms. There is no unique labeled line because both ablation of each individual DAN (Figure 2- figure supplement 1D-G, Figure 4- figure supplement 1) and the acute optogenetic inhibition (Figure 6) do not abolish the odor-high salt memory. The teaching signal is encoded in a redundant manner within the DL1 cluster, yet its distribution among the four DANs is non-uniform and relies heavily on the functionalities of DAN-g1 and DAN-f1 (Table 1, Figure 4-6). Each DAN possesses a distinct identity, which is defined by its specific, albeit partially overlapping, neuronal input circuitry and physiological response to salt (Table 1, Figure 3 and 7).

DAN-c1 responds to both positive and negative gustatory stimuli (Figure 3A). However, the activity of the cell is not sufficient or necessary for aversive memory (Figure 4- figure supplement 1A and E, Figure 5A and Figure 6A). These results suggest that DAN-c1 activity does not encode a taste-specific teaching signal, which is supported by EM analysis, as this particular cell is only weakly connected to three sensory neurons via a single sensor-to-DAN interneuron (Figure 7A and D). However, this interpretation might depend on the type and duration of DAN stimulation, as it was recently shown that - as in our case - optogenetic activation does not induce aversive memory, but thermogenetic 30 minute stimulation in a non-reciprocal one trail paradigm leads to aversive memory (Qi et al 2024). In addition, it could also be that the function of DAN-c1 is more to integrate state-dependent network changes, similar to the role of some DANs in innate and state-dependent preference behavior of adults (Boto et al 2019, Cazale-Debat et al 2024, Cohn et al 2015, Lewis et al 2015, Mohammad et al 2024, Siju et al 2020). DAN-d1 activity appears to have a negligible impact on the formation of odor-high salt memories in the paradigms we tested. Even though the neuron responds to high salt concentrations (Figure 3B) and receives input from 42 sensory neurons via six sensor-to-DAN interneurons (Figure 7A and D), its activity did not induce an aversive memory (Figure 5B) and was not found to be essential for odor-high salt memory (Figure 2- figure supplement 1D and E, Figure 4- figure supplement 1B and F, Figure 5B). Nonetheless, it is possible that DAN-d1 could instruct other salt-dependent memory types, established after non-associative training, second-order conditioning, or reversal learning (Kacsoh et al 2015, Konig et al 2019, Mancini et al 2019, Paranjpe et al 2012, Tabone & de Belle 2011, Yamada et al 2023). Other learning experiments in larvae (Eschbach et al 2020, Weiglein et al 2021) demonstrate that DAN-d1 activation can indeed guide the formation of aversive odor memory. However, these studies employ three training repetitions rather than a single repetition, as in our approach. Additionally, they utilize absolute conditioning with a single odor, in contrast to our protocol of differential conditioning involving two odors. Furthermore, their testing does not incorporate salt as an aversive gustatory stimulus. The observed differences in results suggest that DAN-d1 activation may only establish aversive memory if the stimulation is applied multiple times and is more distinctly presented during training. In combination, DAN-f1 activity and DAN-g1 seem to have a crucial function in high salt punishment. DAN-f1 is inhibited by sugar (Figure 3C) and obtains input from 35 sensory neurons through six sensor-to-DAN interneurons (Figure 7A and D). Despite ablation and acute suppression of activity having no impact on the retention of a memory related to a high salt stimulus (Figure 2- figure supplement 1C, E and G, Figure 4- figure supplement 1C and G, Figure 6C), DAN-f1 activity can instruct the formation of an aversive memory (Figure 5C). However, there is the possibility that the impact of the ablation experiments could be more pronounced than what is observed here, given the potential for developmental compensation. This has been anatomically demonstrated in the adult dopamine mutants by serotonergic cells (Niens et al 2017). The identified function for DAN-f1 activity is consistent with previous research, although their experimental designs differ from our own in the test situation (Eschbach et al 2020, Weiglein et al 2021). Consequently, it is plausible that DAN-f1 activity encodes for a teaching signal that carries additional information beyond high salt. It is evident that DAN-g1 exerts the most potent effect. Following exposure to high salt, this cell becomes active (Figure 3D), is crucial for the formation of an odor-high salt memory (Figure 6D) and can even establish an aversive memory that can be recalled on a high salt containing test plate (Figure 5D). Additionally, it receives extensive sensory input from nine neurons through 11 distinct interneurons (Figure 7A and D). Therefore, DAN-g1 is the only cell of the DL1 cluster that satisfies all prerequisites to independently encode the high salt teaching signal. In summary, the DL1 cluster appears to be arranged in a reasonable functional configuration. The instructive signal is encoded by a mere four cells, with the requisite of partial redundancy to preclude complete loss of function in the event of single-cell failure. Nevertheless, the cells manifest variation, thereby covering different and cumulative characteristics of stimulation.

However, these cellular-level interpretations are inevitably oversimplified. The MB054B split line occasionally exhibits a weak indication of a third cell (Figure 2J), introducing a degree of variability in the experiments. Similarly, DAN-f1 demonstrates no direct response to salt (Figure 3C), suggesting it is not essential for salt learning, yet it triggers a potent aversive memory (Figure 5C). Nevertheless, it is worth mentioning that the physiological and EM analyses were conducted using first instar larvae, whereas the light microscopy experiments and behavioral tests were performed with third instar larvae. Hence, it is possible that developmental-specific alterations may have occurred, which were not accounted for in this study. Regrettably, our understanding of how salt is perceived in larvae remains limited (Liu et al 2003). Assuming peripheral organ perception as primary, the left and right interneuron 1 (also called FB2IN-12), connected to DAN-d1, DAN-f1, and DAN-g1, is presumed to be centrally important as it encodes for the main peripheral gustatory information (encoded by yellow color, Figure 7C, first and second line). However, these DANs display no response to low salt concentrations and varying responses to high salt concentrations and fructose. In contrast, DAN-c1 is the only cell that responds to high and low concentrations of salt (Figure 3A) but is not connected to interneuron pair 1 or other sensory-to DAN interneurons (except interneuron 20L; Figure 7C). Further experiments are therefore necessary to refine our initial cellular-level description and comprehend the functional spectrum of individual DL1 DANs more comprehensively.

Furthermore, other factors must also be taken into account. Blocking almost all TH-positive cells does not result in the complete disappearance of the salt memory. Instead, it is merely reduced by half (Figure 1H and I). There could be several reasons for this outcome, or a combination thereof. For example, it is plausible that the UAS-GtACR2 effector does not completely suppress the activity of dopaminergic neurons. Additionally, the memory may encompass different types (Widmann et al 2016), not all of which are associated with dopamine function. It is also noteworthy that TH-Gal4 does not encompass all dopaminergic neurons - even a neuron from the DL1 cluster is missing (Selcho et al 2009). Considering that we are using high salt concentrations in this experiment, it is also conceivable that no taste-driven memories are formed solely due to the systemic effects of salt (e.g., increased osmotic pressure). These effects naturally complicate the interpretation of the data at the single-cell level.

The functioning of the DL1 cluster DANs as a multimodal node for punishment

Larvae, of course, perceive not only high salt concentrations as punishing, but also other sensory stimuli such as bitter substances (quinines and caffeine), electric shock, temperature, mechanosensory input (vibration), and light (Aceves-Pina & Quinn 1979, Apostolopoulou et al 2016, Apostolopoulou et al 2014, Eschbach et al 2011, Gerber & Hendel 2006, Khurana et al 2012, Pauls et al 2010, von Essen et al 2011). However, the extent to which the DL1 cluster is involved in coding for most of these aversive teaching signals remains unclear. Currently, comparisons at the level of DANs can only be made for sensory inputs that are dependent on high salt or mechanosensation. Interestingly, similar to high salt concentrations, somatosensory information can also result in the establishment of aversive olfactory memory in larvae (Eschbach et al 2020). The transmission of these somatosensory teaching signals is orchestrated by mechanosensory neurons from the chordotonal organs, class IV multidendritic nociceptive neurons distributed along the body, and multisensory basin neurons (Hwang et al 2007, Jovanic et al 2016, Ohyama et al 2015, Tracey et al 2003). The connectome revealed that these mechanosensory neurons also link to the DL1 cluster through three to four interneurons and elicit the activation of DAN-d1, DAN-f1, and DAN-g1 (Eschbach et al 2020). Upon comparing the interneurons that relay high salt and mechanosensory information to the DL1 cluster, six pairs of interneurons are of critical significance, respectively. These pairs include FFN-20, FB2N-12, FB2N-19, FFN-23, FFN-29, and FB2N-15 for mechanosensory information (Eschbach et al 2020), and FB2N-12, FB2N-18, FB2IN-11, FB2IN-6, FFN-21, and FFN-24 for high salt concentrations (Figure 7; Figure 7- figure supplement 1A, rightmost column). Interestingly, only one pair of interneurons - FB2N12 - is shared between the two types of information and the other five types are specific to the sensory modality. The wiring patterns of DAN input neurons exhibit a notable consistency, with a limited number of neurons dedicated to a specific sensory modality, and single neurons shared across different modalities. This results in a unique combination of cellular codes that are partly redundant and partly specific for the four DL1 DANs. While certain DANs may hold greater significance for a particular information, such as DAN-d1 for somatosensory information, the DAN-g1 cell appears to play a general central role for encoding an aversive teaching signal. DAN-g1 responds to all aversive stimuli tested thus far and its activity has proven informative for all aversive memories analyzed to date. In order to gain a deeper comprehension of the processing of diverse sensory modalities in the larval brain, it is crucial to broaden the analytical framework presented in our work to encompass visual stimuli, various bitter substances, and temperature stimuli.

However, the overall depiction presented above of the information flow and processing of DL1 DANs is an oversimplification. Indeed, DL1 and pPAM DANs are among the most complex and highly recurrent neurons in the brain (Winding et al 2023). This exceptional level of recurrent connectivity enables DANs to provide high-dimensional feedback, which enables them to encode a diverse range of features and engage in parallel computations. Such computations can guide a working memory, contain feedback from neurons that integrate both learned and innate values, and receive long-range feedback from descending neurons that encode motor commands; all in addition to polysynaptic feedforward inputs from the entire set of sensory modalities.

A cellular comparison of DAN function across Drosophila metamorphosis

A recent study conducted by Truman and colleagues explored the integration, or lack thereof, of individual input and output neurons from the larval MB into the adult network following metamorphosis (Truman et al 2023). The study demonstrated that while all four DANs of the pPAM cluster perished, three out of the four cells from the DL1 cluster persisted as part of the adult MB circuitry. Only DAN-f1 altered its innervation in the adult brain, relocating from the larval MB to the adult superior medial protocerebrum. In contrast, DAN-c1, DAN-d1, and DAN-g1 maintained their function as MB DANs in the adult system, identified as PPL1-γ1pedc, PPL1-γ2α’1, and PPL1-γ1, respectively. Due to the lack of appropriate genetic tools, data on the function of PPL1-γ1 is scarce (Aso et al 2014). Therefore, a functional comparison between larval and adult DANs is only possible for the DAN-c1/PPL1-γ1pedc and DAN-d1/PPL1-γ2α’1 combination. Embryonic-born DAN-c1/PPL1-γ1pedc, DAN-d1/PPL1-γ2α’1, and DAN-g1/PPL1-γ1 exhibit sustained activity and respond to various sensory stimuli, including electric shock, temperature, or bitter taste, that can instruct short lasting aversive memories (Aso et al 2012, Aso & Rubin 2016, Aso et al 2010, Claridge-Chang et al 2009, Das et al 2014, Galili et al 2014, Huang et al 2024, Kirkhart & Scott 2015, Tomchik 2013, Villar et al 2022, Vrontou et al 2021). Despite the massive reorganization of the mushroom body, the basic function of DAN-c1/PPL1-γ1pedc and DAN-d1/PPL1-γ2α’1, mediating punishment teaching signals that instruct short-term memories, remains the same. Evidence indicates that the underlying molecular signaling pathways maintain their identity and kinetics throughout development, with larval and adult interstimulus interval curves for DAN-d1/PPL1-γ2α’1 being almost identical (Aso & Rubin 2016, Huang et al 2024, Weiglein et al 2021). During adulthood, PPL1-DANs are responsible for numerous other functions, including innate olfactory responses, suppression of appetite memory in response to nutritional status, pre-exposure learning, memory reconsolidation, and forgetting (Berry et al 2015, Berry et al 2018, Felsenberg et al 2017, Jacob et al 2021, Krashes et al 2009, McCurdy et al 2021, Placais et al 2012, Siju et al 2020, Tian et al 2016, Vrontou et al 2021). Additionally, they are involved in various behaviors like locomotion and sleep (Berry et al 2015).

An open question is whether larval DL1 DANs encode these same functions to a similar extent. Our study establishes a fundamental framework that can be leveraged to investigate this issue. Considering the numerical simplicity of the larval nervous system, the available genetic tools, and the complete connectome, it may be possible to attain an understanding of the larval dopaminergic system at the level of single cells and single synapses. It is plausible that nature has a limited number of effective neural circuit solutions for complex cognitive and behavioral problems, such as the remarkable ability of brains to learn from a small number of examples. Therefore, comprehending the larval dopaminergic system may provide insight into the shared circuit motifs that exist across the animal kingdom.

Materials and methods

Fly Strains

Flies were raised and maintained on Drosophila standard food at 25°C, 60-80% relative humidity and a 14/7-hour light/dark cycle. For anatomical analysis, UAS-mCD8::GFP (Bloomington stock center no. 32194) (Selcho et al 2009), UAS-mCD8::GFP;mb247-LexA,lexAop-mRFP/TM3,Sb (Burke et al 2012) and UAS-mCD8::GFP;nSyb-LexA,lexAop-mRFP/TM6B (Burke et al 2012, Riabinina et al 2015) were used to analyze the morphology of DANs. For behavioral experiments, UAS-hid,rpr (Abbott & Lengyel 1991, White et al 1996), UAS-GtACR2 (Mohammad et al 2017) (Bloomington stock center no. 92984) and UAS-ChR2XXL (Dawydow et al 2014) (Bloomington stock center no. 58374) were used to ablate or optogenetically silence or activate neurons by blue light (470 nm). UAS-GCaMP6m (Chen et al 2013) (Bloomington stock center no. 42748) flies were used for Ca2+-imaging experiments. The Gal4 strains TH-Gal4 (Friggi-Grelin et al 2003) and R58E02-Gal4 (Rohwedder et al 2016) (Bloomington stock center no. 41347) were used to manipulate different sets of DANs. Split-Gal4 lines MB054B, MB065B, MB328B, MB143B, MB145B, SS02160, SS02180, SS01716, SS01958, SS21716, SS24765, SS01702 (Eschbach et al 2020, Pfeiffer et al 2010, Saumweber et al 2018) were used for anatomical, physiological and behavioral experiments. w1118 flies (Bloomington stock center no. 3605) were used to obtain heterozygous driver and effector control groups. See supplementary table 1 for more information.

Anatomical Analysis

Immunostaining

To confirm expression patterns of R58E02-Gal4, TH-Gal4 and MB054B split-Gal4, flies were crossed to UAS-mCD8::GFP. To assess the functionality of the ablation by effector UAS-hid,rpr, split-Gal4 MB054B was crossed with the latter. Furthermore, MB054B>+ and +>UAS-hid,rpr were anatomically examined for validation. All flies were raised at 25°C for seven days before dissection. Third instar larvae (wandering stage) were put on ice and dissected in PBS (Phosphate Buffered Saline; Sigma Aldrich, cat. no. P4417). Brains were fixed in 4% formaldehyde solution (in PBS, Thermo Scientific, cat. no. 047392.9M) for 20 minutes at room temperature. After eight rinses in PBT (PBS with 3% Triton X-100, Sigma Aldrich, cat. no. X100), brains were blocked with 5% normal goat serum (NGS, Sigma Aldrich, cat. no. G9023) in PBT for 1 hour at room temperature and incubated for 48 hours with primary antibodies at 4°C. Before application of the secondary antibodies for at least 24 hours at 4°C, brains were washed seven times with PBT. After that, larval brains were again washed eight times with PBT and mounted on poly-L-lysin-coated coverslips (Janelia FlyLight DPX mounting receipt: https://www.janelia.org/project-team/flylight/protocols), dehydrated through a series of increasing concentrations of ethanol and cleared three times for 5 minutes in xylene (Sigma Aldrich, cat. no. 247642). At the end, larval brains were mounted in DPX mounting medium (dibutyl phthalate in xylene, Sigma Aldrich, cat. no. 06522) and left to rest in the dark for at least 24 hours before imaging. Primary antibodies were: rat anti-N-Cadherin (1:50; Hybridoma, cat. no. DN-Ex #8-s), rabbit anti-GFP (1:1000; Life Technologies, cat. no. A6455), mouse 4F3 anti-DLG (1:200; Hybridoma, cat. no. 4F3 anti-discs large) and mouse anti-TH (1:50, Immunostar, cat. no. 22941). Secondary antibodies were: goat anti-rat Alexa Fluor 647 (1:500 or 1:250, for anti-N-Cadherin, Life Technologies, cat. no. A21247), goat anti-rabbit Alexa Fluor 488 (1:500, for anti-GFP, Life Technologies, cat. no. A11008), goat anti-mouse Alexa Fluor 568 (1:500, for anti-DLG, Life Technologies, cat. no. A10037) and goat anti-mouse Cy3 (1:200, for anti-TH, life technologies, cat. no. A10521).

Native fluorescence

For native fluorescence, we used a previously described protocol for adult flies (Pitman et al 2011) and adapted it to larvae. All split-Gal4 lines were crossed to UAS-mCD8::GFP;mb247-LexA,lexAop-mRFP/TM3,Sb, while split-Gal4 line MB054B was additionally crossed to UAS-mCD8::GFP;nSyb-LexA,lexAop-mRFP/TM6B and raised at 25°C for seven days. Third instar larvae (wandering stage) were put on ice and dissected in PBS. Brains were fixed under vacuum in 4% formaldehyde solution (in PBS) for 25 minutes at room temperature. After five washes in PBT (PBS with 3% Triton X-100), larval brains were rinsed twice directly and once for 10 minutes under vacuum in PBS at room temperature. Samples were mounted on poly-L-lysin-coated coverslips and embedded in Vectashield (Vector Laboratories, cat. no. H-1000-10). Slides were stored in the dark and cold for at least 24 hours before imaging.

Confocal microscopy

Confocal microscopy was conducted on a Zeiss LSM800 confocal laser scanning microscope with ZEN 2.3 software. All images were projected and adjusted with ImageJ (Fiji is just ImageJ, Version 1.53c, Java 1.8.0_172 (64-bit)). Final Figures were arranged and labeled in Affinity Publisher 2.1.1.

Functional Imaging and microfluidics

For larval imaging experiments, adult flies were transferred to larvae collection cages (Genesee Scientific) containing grape juice agar plates and 180 mg of fresh yeast paste per cage. Flies were allowed to lay eggs on the agar plate for 1–2 days before the plate was removed for collection of larvae. Calcium imaging experiments were performed in L1 larvae (2 days AEL). We used a previously described method for microfluidic delivery of chemicals in aqueous form with simultaneous imaging of calcium activity in intact larvae (Si et al 2019). All experiments used an 8-channel microfluidic chip equipped with a vacuum port to stabilize the animal’s head. The same tastants were used as in the behavioral experiments (NaCl: VWR Chemicals, cat. no. 27810.364; D-Fructose: Sigma Aldrich cat. no. 47740). Stimuli consisted of 5-10 s pulses interleaved with 15 s water washout periods. An L1 larva was washed with deionized water and loaded into the microfluidic device using a 1 mL syringe filled with Triton X-100 (0.1% [v/v]) solution. The animal was pushed to the end of the loading channel with its dorsal side facing the objective. GCaMP6m signal was recorded using an inverted Nikon Ti-E spinning disk confocal microscope and a 60X water immersion objective (NA 1.2). A CCD microscope camera (Andor iXon EMCCD) captured frames at 30 Hz. Recordings from at least 5–7 larvae were collected for each condition. For each larva the average response over 5s around the peak value during stimulus presentation was compared to chance level analyzed via normal t-test. Responses during stimulation were normalized based on the baseline response 10s before the stimulation. P-values for each experiment are given in the respective figure panel. Each graph shows the mean calcium signal plotted as the relative response strength ΔF/F and the related standard error of the mean on the y axis. The time in seconds is given below each graph on the x axis. The grey box indicates the duration of the stimulus application. The sample size of each group (N=5-7) is given above each row. n.s. p > 0.05; * p < 0.05.

Behavioral Experiments

Associative olfactory learning

Standard experiments were done as described before (Apostolopoulou et al 2013, Gerber et al 2013, Hendel et al 2005, Michels et al 2017, Widmann et al 2018). Learning experiments were conducted on assay plates (85 mm diameter, Sarstedt, cat. no. 82.1472) filled with a thin layer of either 2.5% (w/v) pure agarose solution (Sigma Aldrich, cat. no. A9539) or 2.5% (w/v) agarose plus either 1.5 M sodium chloride solution (VWR Chemicals, cat. no. 27810.364), 2 M D-Fructose solution (Sigma Aldrich cat. no. 47740) or 10 mM quinine solution (quinine hemisulfate; Sigma Aldrich cat. no. Q1250). Before closing and labeling the lids, solutions were let to cool down at room temperature to avoid condensation. Plates were stored at 18°C and used within five days. As olfactory stimuli, we used two different odor combinations: either amyl acetate (AM, Sigma Aldrich, cat. no. 46022) diluted 1:250 in paraffin oil (Sigma Aldrich, cat. no. 76235) and undiluted benzaldehyde (BA, Sigma Aldrich, cat. no. 12010) or benzaldehyde and hexyl acetate (HA, Sigma Aldrich, cat. no. 108154) both diluted 1:100 in paraffin oil. For both odor combinations, 10 µL odor were loaded into custom-made Teflon containers (4.5 mm diameter) with perforated lids (Scherer et al 2003).

A minimum of 30 early third instar larvae were collected and exposed to a first odor (amyl acetate or hexyl acetate) while crawling on pure agarose for five minutes, followed by five minutes exposition to a second odor (benzaldehyde) on agarose medium with either 1.5 M sodium chloride, 10 mM quinine as negative reinforcer or 2 M D-Fructose as positive reinforcer (Odor1/Odor2+). For three cycle training, two more repetitions of training trials were performed. A second group of larvae received reciprocal training (Odor1+/Odor2). Please note that occurrence of odor and reinforcer during training have been randomized to exclude sequential or positional effects. After one or three training cycles, larvae were transferred onto test plates containing either pure agarose (reward learning) or agarose plus salt or quinine (punishment learning) on which odor 1 and odor 2 were represented on opposite sides. After five minutes, larvae were counted on each side of the test plate (#Odor1, #Odor2) or in a 10 mm neutral zone in the middle of the plate (#Neutral). A Preference index (PREF) for each group of larvae was calculated as follows: To measure specifically the effect of associative learning, we then calculated the associative performance index (PI) as follows: Negative Performance Indices represent aversive associated learning, whereas positive values represent appetitive learning. Division by 2 ensures scores are bound within (−1; 1).

Odor preference test

Pure agarose (2.5% w/v) plates were cast as previously described (Apostolopoulou et al 2013, Gerber et al 2013, Michels et al 2017, Scherer et al 2003, Widmann et al 2018). Furthermore, the odors were prepared according to the dilutions used in the learning experiments. An odor container (either AM 1:250, BA undiluted, HA 1:100, or BA 1:100) was placed on one side of the agarose plate. An empty container was placed on the opposite side to exclude visual or other side effects. Approximately 30 third instar feeding stage larvae were collected and placed on the plate for 5 min. After this time, larvae were counted, subdivided into larvae on the odor side (#Odor), larvae on the empty container side (#Empty) and larvae in a 10 mm neutral zone in the middle of the plate, as well as larvae that were on the lid of the Petri dish (#Neutral). Preferences were calculated as follows: Positive values indicate a preference for the respective odor, while negative values indicate odor avoidance. The preference tests were performed in different orientations (odor container on plate up/down/left/right) to exclude directional bias.

Gustatory preference test

High salt-dependent choice behavior experiments were performed using standard methods (Hendel et al 2005, Huser et al 2017, Huser et al 2012, Niewalda et al 2008, Selcho et al 2009, Widmann et al 2016). A 2.5% (w/v) agarose solution was boiled as previously described and a thick layer was poured into Petri dishes. After cooling, half of the agarose in the Petri dish was removed and filled by 2.5% (w/v) agarose solution with 1.5 M sodium chloride. For the choice experiment, about 30 third instar feeding stage larvae were put in the middle of the Petri dish and let them crawl for five minutes. Larvae were then counted as being located on the sodium chloride side (#NaCl), pure agarose side (#Agarose), or a neutral area of about 10 mm width in the middle of the plate (#Neutral), as well as larvae that were on the lid of the Petri dish (#Neutral). The gustatory preference indices for sodium chloride were calculated as follows: Negatives values for preference indices indicate an aversion to sodium chloride. The gustatory preference tests were performed in different orientations (NaCl solution on plate up/down/left/right) to exclude directional bias.

Optogenetic substitution experiment

To substitute an actual salt punishment by remotely activating DANs, we used UAS-ChR2XXL. Effector lines were crossed to w1118 to obtain appropriate genetic controls. Larvae were maintained in vials with Drosophila standard food on 25°C and wrapped in aluminum foil to ensure development in constant darkness. A group of 30 feeding-stage third-instar larvae were placed onto plates containing 2.5% agarose and exposed to either AM or BA. During the presentation of the first odor, the larvae were exposed to blue light (470 nm, 220 lux) for 5 minutes. The second odor BA (undiluted) was presented in darkness for five minutes on a pure agarose plate. As described for odor-high salt learning, training was performed reciprocally and the sequence of training trials was alternated across repetitions of the experiment. Furthermore, a control group, identical to the experimental group, underwent no blue light stimulation during the learning paradigm. Data were then scored as above. Please note, tests were done in the presence of 1.5 M sodium chloride to test for the specific quality of the memory. The activation of blue light itself does not cause any disruptive behavioral side effects, as it does not induce an appetitive or aversive memory on its own.

Optogenetic inhibition of neuronal activity

To acutely block synaptic output, we used UAS-GtACR2. Flies were maintained on standard food supplemented with 0.5 mM all-trans-retinal (ATR, Sigma Aldrich, cat. no. R2500) at 25°C as described before (Meloni et al 2020). Vials were wrapped in aluminum foil to ensure larval development in constant darkness. Groups of about 30 third instar feeding stage larvae received a reciprocal two-odor training as described before, exposed to blue light (470 nm, 1100 lux) during the entire training phase. The sequence of training trials was alternated across repetitions of the experiment. The test was performed for 5 minutes in darkness on an agarose plate mixed with 1.5 M sodium chloride. Control experiments were performed with the same genotype but with standard food lacking 0.5 mM all-trans-retinal. Please note that the blue light exposure itself does not cause any disruptive behavioral side effects, as it does not impair odor-high salt learning of control animals. Data were then calculated and scored as mentioned above.

Evaluation of DAN input wiring diagrams

Pie charts of sensory profiles were calculated using the percentage of total synaptic input of interneurons as fraction (thereby ignoring other inputs to show distribution of sensory origins). Percentages then give the percentage of total sensory synaptic input to interneurons. The calculation of the hub score was done in the following way: Fraction of total synaptic input from all sensory neurons to defined interneurons (see IDs) was multiplied by the total fraction of input of the DANs from this interneuron. For anatomical reconstructions and visualizations included here, we made use of earlier published data (Eichler et al 2017, Eschbach et al 2020, Miroschnikow et al 2018, Winding et al 2023).

Statistical Analysis

Behavioral results were analyzed in GraphPad Prism 8.4.3. To test for normal distribution, Shapiro-Wilk test was applied. Groups that did not violate the assumption of normal distribution, were analyzed with an unpaired t-test (comparison of two groups) or a one-way ANOVA followed by a Tukey’s post hoc test (comparisons between groups larger than two). For nonparametric statistics, Mann-Whitney test (comparison between two groups) or Kruskal-Wallis followed by Dunn’s multiple comparisons tests were applied. Furthermore, one sample t-test for parametric or Wilcoxon signed-rank test for nonparametric data sets were performed to compare means or medians against chance level. Results are visualized in box-plots, indicating the median as middle line, 25%/75% quantiles as box boundaries and minimum/maximum performance indices as whiskers. Each data point is represented as black dot and sample sizes are noted within graphics. Asterisks and ‘‘n.s.’’ indicate p < 0.05 and p > 0.05, respectively. The source data and results of all statistical tests are documented in the Supplemental Information file.

eLife Assessment

This comprehensive study presents important findings that delineate how specific dopaminergic neurons (DANs) instruct aversive learning in Drosophila larvae exposed to high salt through an integration of behavioral experiments, imaging, and connectomic analysis. The work reveals how a numerically minimal circuit achieves remarkable functional complexity, with redundancies and synergies within the DL1 cluster that challenge our understanding of how few neurons generate learning behaviors. By establishing a framework for sensory-driven learning pathways, the study makes a compelling and substantial contribution to understanding associative conditioning while demonstrating conservation of learning mechanisms across Drosophila developmental stages.

Reviewer #1 (Public review):

Summary:

In this paper Weber et al. investigate the role of 4 dopaminergic neurons of the Drosophila larva in mediating the association between an aversive high-salt stimulus and a neutral odor. The 4 DANs belong to the DL1 cluster and innervate non-overlapping compartments of the mushroom body, distinct from those involved in appetitive associative learning. Using specific driver lines for individual neurons, the authors show that activation of the DAN-g1 is sufficient to mimic an aversive memory and it is also necessary to form a high-salt memory of full strength, although optogenetic silencing of this neuron has only a partial phenotype. The authors use calcium imaging to show that the DAN-g1 is not the only DAN responding to salt. DAN-c1 and d1 also respond to salt, but they seem to play no role for the associative memory. DAN-f1, which does not respond to salt, is able to lead to the formation of a memory (if optogenetically activated), but it is not necessary for the salt-odor memory formation in normal conditions. However, when silenced together with DAN-g1, it enhances the memory deficit of DAN-g1. Overall, this work brings evidence of a complex interaction between DL1 DANs in both the encoding of salt signals and their teaching role in associative learning, with none of them being individually necessary and sufficient for both functions.

Strengths:

Overall, the manuscript contributes interesting results that are useful to understand the organization and function of the dopaminergic system. The behavioral role of the specific DANs is accessed using specific driver lines which allow to test their function individually and in pairs. Moreover, the authors perform calcium imaging to test whether DANs are activated by salt, a prerequisite for inducing a negative association to it. Proper genetic controls are carried across the manuscript.

Weaknesses:

The authors use two different approaches to silence dopaminergic neurons: optogenetics and induction of apoptosis. The results are not always consistent, but the authors discuss these differences appropriately. In general, the optogenetic approach is more appropriate as developmental compensations are not of major interest for the question investigated.

The physiological data would suggest the role of a certain subset of DANs in salt-odor association, but a different partially overlapping set is necessary in behavioral assays (with a partial phenotype). No manipulation completely abolishes the salt-odor association, leaving important open questions on the identity of the neural circuits involved in this behavior.

The EM data analysis reveals a non-trivial organization of sensory inputs into DANs, but it is difficult to extrapolate a link to the functional data presented in the paper.

Reviewer #2 (Public review):

Summary:

In this work the authors show that dopaminergic neurons (DANs) from the DL1 cluster in Drosophila larvae are required for the formation of aversive memories. DL1 DANs complement pPAM cluster neurons which are required for the formation of attractive memories. This shows the compartmentalized network organization of how an insect learning center (the mushroom body) encodes memory by integrating olfactory stimuli with aversive or attractive teaching signals. Interestingly, the authors found that the 4 main dopaminergic DL1 neurons act partially redundant, and that single cell ablation did not result in aversive memory defects. However, ablation or silencing of a specific DL1 subset (DAN-f1,g1) resulted in reduced salt aversion learning, which was specific to salt but no other aversive teaching stimuli tested. Importantly, activation of these DANs using an optogenetic approach was also sufficient to induce aversive learning in the presence of high salt. Together with the functional imaging of salt and fructose responses of the individual DANs and the implemented connectome analysis of sensory (and other) inputs to DL1/pPAM DANs this represents a very comprehensive study linking the structural, functional and behavioral role of DL1 DANs. This provides fundamental insight into the function of a simple yet efficiently organized learning center which displays highly conserved features of integrating teaching signals with other sensory cues via dopaminergic signaling.

Strengths:

This is a very careful, precise and meticulous study identifying the main larval DANs involved in aversive learning using high salt as a teaching signal. This is highly interesting because it allows to define the cellular substrates and pathways of aversive learning down to the single cell level in a system without much redundancy. It therefore sets the basis to conduct even more sophisticated experiments and together with the neat connectome analysis opens the possibility to unravel different sensory processing pathways within the DL1 cluster and integration with the higher order circuit elements (Kenyon cells and MBONs). The authors' claims are well substantiated by the data and balanced, putting their data in the appropriate context. The authors also implemented neat pathway analyses using the larval connectome data to its full advantage, thus providing network pathways that contribute towards explaining the obtained results.

Weaknesses:

Previous comments were fully addressed by the authors.

Reviewer #3 (Public review):

The study of Weber et al. provides a thorough investigation of the roles of four individual dopamine neurons for aversive associative learning in the Drosophila larva. They focus on the neurons of the DL-1 cluster which already have been shown to signal aversive teaching signals. But the authors go beyond the previous publications and test whether each of these dopamine neurons responds to salt or sugar, is necessary for learning about salt, bitter, or sugar, and is sufficient to induce a memory when optogenetically activated. In addition, previously published connectomic data is used to analyze the synaptic input to each of these dopamine neurons. The authors conclude that the aversive teaching signal induced by salt is distributed across the four DL-1 dopamine neurons, with two of them, DAN-f1 and DAN-g1, being particularly important. Overall, the experiments are well designed and performed, support the authors' conclusions, and deepen our understanding of the dopaminergic punishment system.

Strengths:

(1) This study provides, at least to my knowledge, the first in vivo imaging of larval dopamine neurons in response to tastants. Although the selection of tastants is limited, the results close an important gap in our understanding of the function of these neurons.

(2) The authors performed a large number of experiments to probe for the necessity of each individual dopamine neuron, as well as combinations of neurons, for associative learning. This includes two different training regimen (1 or 3 trials), three different tastants (salt, quinine and fructose) and two different effectors, one ablating the neuron, the other one acutely silencing it. This thorough work is highly commendable, and the results prove that it was worth it. The authors find that only one neuron, DAN-g1, is partially necessary for salt learning when acutely silenced, whereas a combination of two neurons, DAN-f1 and DAN-g1, are necessary for salt learning when either being ablated or silenced.

(3) In addition, the authors probe whether any of the DL-1 neurons is sufficient for inducing an aversive memory. They found this to be the case for two of the neurons, largely confirming previous results obtained by a different learning paradigm, parameters and effector.

(4) This study also takes into account connectomic data to analyze the sensory input that each of the dopamine neurons receives. This analysis provides a welcome addition to previous studies and helps to gain a more complete understanding. The authors find large differences in inputs that each neuron receives, and little overlap in input that the dopamine neurons of the "aversive" DL-1 cluster and the "appetitive" pPAM cluster seem to receive.

(5) Finally, the authors try to link all the gathered information in order to describe an updated working model of how aversive teaching signals are carried by dopamine neurons to the larva's memory center. This includes important comparisons both between two different aversive stimuli (salt and nociception) and between the larval and adult stages.

Author response:

The following is the authors’ response to the previous reviews

Public Reviews:

Reviewer #1 (Public review):

Summary:

In this paper Weber et al. investigate the role of 4 dopaminergic neurons of the Drosophila larva in mediating the association between an aversive high-salt stimulus and a neutral odor. The 4 DANs belong to the DL1 cluster and innervate non-overlapping compartments of the mushroom body, distinct from those involved in appetitive associative learning. Using specific driver lines for individual neurons, the authors show that activation of the DAN-g1 is sufficient to mimic an aversive memory and it is also necessary to form a high-salt memory of full strength, although optogenetic silencing of this neuron has only a partial phenotype. The authors use calcium imaging to show that the DAN-g1 is not the only DAN responding to salt. DAN-c1 and d1 also respond to salt, but they seem to play no role for the associative memory. DAN-f1, which does not respond to salt, is able to lead to the formation of a memory (if optogenetically activated), but it is not necessary for the salt-odor memory formation in normal conditions. However, when silenced together with DAN-g1, it enhances the memory deficit of DAN-g1. Overall, this work brings evidence of a complex interaction between DL1 DANs in both the encoding of salt signals and their teaching role in associative learning, with none of them being individually necessary and sufficient for both functions.

Strengths:

Overall, the manuscript contributes interesting results that are useful to understand the organization and function of the dopaminergic system. The behavioral role of the specific DANs is accessed using specific driver lines which allow to test their function individually and in pairs. Moreover, the authors perform calcium imaging to test whether DANs are activated by salt, a prerequisite for inducing a negative association to it. Proper genetic controls are carried across the manuscript.

Weaknesses:

The authors use two different approaches to silence dopaminergic neurons: optogenetics and induction of apoptosis. The results are not always consistent, but the authors discuss these differences appropriately. In general, the optogenetic approach is more appropriate as developmental compensations are not of major interest for the question investigated.

The physiological data would suggest the role of a certain subset of DANs in salt-odor association, but a different partially overlapping set is necessary in behavioral assays (with a partial phenotype). No manipulation completely abolishes the salt-odor association, leaving important open questions on the identity of the neural circuits involved in this behavior.

The EM data analysis reveals a non-trivial organization of sensory inputs into DANs, but it is difficult to extrapolate a link to the functional data presented in the paper.

We would like to once again thank Reviewer 1 for the positive assessment of our work and for the valuable suggestions provided on the first revision of the manuscript. In this second revision, we have addressed the linguistic issues and most of the minor comments as recommended. We now hope that the current version of our manuscript meets the reviewer’s expectations both in terms of language and content.

Reviewer #2 (Public review):

Summary:

In this work the authors show that dopaminergic neurons (DANs) from the DL1 cluster in Drosophila larvae are required for the formation of aversive memories. DL1 DANs complement pPAM cluster neurons which are required for the formation of attractive memories. This shows the compartmentalized network organization of how an insect learning center (the mushroom body) encodes memory by integrating olfactory stimuli with aversive or attractive teaching signals. Interestingly, the authors found that the 4 main dopaminergic DL1 neurons act partially redundant, and that single cell ablation did not result in aversive memory defects. However, ablation or silencing of a specific DL1 subset (DAN-f1,g1) resulted in reduced salt aversion learning, which was specific to salt but no other aversive teaching stimuli tested. Importantly, activation of these DANs using an optogenetic approach was also sufficient to induce aversive learning in the presence of high salt. Together with the functional imaging of salt and fructose responses of the individual DANs and the implemented connectome analysis of sensory (and other) inputs to DL1/pPAM DANs this represents a very comprehensive study linking the structural, functional and behavioral role of DL1 DANs. This provides fundamental insight into the function of a simple yet efficiently organized learning center which displays highly conserved features of integrating teaching signals with other sensory cues via dopaminergic signaling.

Strengths:

This is a very careful, precise and meticulous study identifying the main larval DANs involved in aversive learning using high salt as a teaching signal. This is highly interesting because it allows to define the cellular substrates and pathways of aversive learning down to the single cell level in a system without much redundancy. It therefore sets the basis to conduct even more sophisticated experiments and together with the neat connectome analysis opens the possibility to unravel different sensory processing pathways within the DL1 cluster and integration with the higher order circuit elements (Kenyon cells and MBONs). The authors' claims are well substantiated by the data and balanced, putting their data in the appropriate context. The authors also implemented neat pathway analyses using the larval connectome data to its full advantage, thus providing network pathways that contribute towards explaining the obtained results.

Weaknesses:

Previous comments were fully addressed by the authors.

We sincerely thank Reviewer 2 for the positive evaluation of our work. We are glad that our responses in the first revision addressed the previous concerns and appreciate the reviewer’s constructive feedback once again.

Reviewer #3 (Public review):

Summary:

The study of Weber et al. provides a thorough investigation of the roles of four individual dopamine neurons for aversive associative learning in the Drosophila larva. They focus on the neurons of the DL-1 cluster which already have been shown to signal aversive teaching signals. But the authors go beyond the previous publications and test whether each of these dopamine neurons responds to salt or sugar, is necessary for learning about salt, bitter, or sugar, and is sufficient to induce a memory when optogenetically activated. In addition, previously published connectomic data is used to analyze the synaptic input to each of these dopamine neurons. The authors conclude that the aversive teaching signal induced by salt is distributed across the four DL-1 dopamine neurons, with two of them, DAN-f1 and DAN-g1, being particularly important. Overall, the experiments are well designed and performed, support the authors' conclusions, and deepen our understanding of the dopaminergic punishment system.

Strengths:

(1) This study provides, at least to my knowledge, the first in vivo imaging of larval dopamine neurons in response to tastants. Although the selection of tastants is limited, the results close an important gap in our understanding of the function of these neurons.

(2) The authors performed a large number of experiments to probe for the necessity of each individual dopamine neuron, as well as combinations of neurons, for associative learning. This includes two different training regimen (1 or 3 trials), three different tastants (salt, quinine and fructose) and two different effectors, one ablating the neuron, the other one acutely silencing it. This thorough work is highly commendable, and the results prove that it was worth it. The authors find that only one neuron, DAN-g1, is partially necessary for salt learning when acutely silenced, whereas a combination of two neurons, DAN-f1 and DAN-g1, are necessary for salt learning when either being ablated or silenced.

(3) In addition, the authors probe whether any of the DL-1 neurons is sufficient for inducing an aversive memory. They found this to be the case for two of the neurons, largely confirming previous results obtained by a different learning paradigm, parameters and effector.

(4) This study also takes into account connectomic data to analyze the sensory input that each of the dopamine neurons receives. This analysis provides a welcome addition to previous studies and helps to gain a more complete understanding. The authors find large differences in inputs that each neuron receives, and little overlap in input that the dopamine neurons of the "aversive" DL-1 cluster and the "appetitive" pPAM cluster seem to receive.

(5) Finally, the authors try to link all the gathered information in order to describe an updated working model of how aversive teaching signals are carried by dopamine neurons to the larva's memory center. This includes important comparisons both between two different aversive stimuli (salt and nociception) and between the larval and adult stages.

We would also like to thank Reviewer 3 for the positive assessment of our work. Many of the constructive comments provided were incorporated into the first revision, contributing significantly to the improved clarity and overall quality of the manuscript.

Recommendations for the authors:

Reviewer #1 (Recommendations for the authors):

Here are some minor comments (and some semantics that could be addressed to improve the manuscript)

Title: is the title correct given that c1 and d1 do not really signal punishment?

We think the title is correct and would like to keep it as it is.

L72 striatum misspelled

We have corrected the error.

L74 constitute instead of provide?

We made the suggested modification in the text.

L129: "But can these four individual DANs also process other sensory modalities?" other then what? What was used before?

We have made the required change, which now allows us to contrast somatosensory and chemosensory information.

L172: (Please refer to the discussion regarding the partial reduction of the memory); would be more natural to explain shortly here, or add a sentence before this parenthesis that point to the effect

We made the requested change in the manuscript and added a short sentence before the parenthesis.

L182: "DL1 neurons convey a dopaminergic aversive teaching signal" you cannot make this statement from just TH-GAL4!

We agree - that's why we have completely revised the sentence and now further restricted it and also refer to further larval and adult published data

L264: "possible redundancy among" I don't think you are testing a redundancy here, it is more likely a developmental compensation.

We made the requested change in the sentence and added a potential developmental compensation as an interpretation of our results.

L296: "to determine if the activation of individual DL1 DANs signals aspects of the natural high salt punishment," - how can the optogenetic activation tell something about aspects of the natural salt punishment? I understand the fact that salt is present, but still I find it inaccurate

Our approach is based on the framework established by Bertram Gerber and colleagues over the past two decades in larval Drosophila research. According to this logic, memory recall is dependent on the specific properties of the test context, particularly the type and concentration of the stimulus presented on the test plate. Aversive memory retrieval occurs only when the test conditions closely match those of the training stimulus. Consequently, the larva's behavior on the test plate serves as an indicator of the memory content being recalled. We therefore adhere to this established methodology (Gerber & Hendel, 2006; Schleyer et al., 2011; Schleyer et al., 2015).

L307 "DAN-f1 and DAN-g1 encode aspects of the natural aversive high salt teaching" you cannot conclude that given that f1 does not even respond to salt. I understand the logic of the salt during test, but I think it is still a stretched interpretation

We agree and thus have deleted the sentence.

L310 "Individual DL1 DANs are acutely necessary" this is too general, it seems that only one is

We have changed the title and now clearly state that this is only one DAN of the DL1 cluster.

Reviewer #2 (Recommendations for the authors):

In Fig.6 the text flow could be optimized as the authors first mention Fig. 6E,F before they follow up with Fig. 6A-D.

Thanks for bringing this up – we changed it in the revised version of the manuscript. Now 6A-D is mentioned first.

In Fig.6 the finding that optogenetic inactivation but not ablation of DAN-g1 slightly but significantly reduces aversive salt learning suggests that there is an individual contribution of this DAN in this paradigm. The authors emphasize redundancy of DL1 DANs although the effect size seems comparable between DAN-g1 and DAN-f1,g1 silencing.

In response to this concern and the one of reviewer 2, we have revised the section title and removed the final sentence of the section before to avoid placing emphasis on the potential redundancy of DL1 DANs within this results section.

Reviewer #3 (Recommendations for the authors):

The authors replied to each issue I raised, and revised their manuscript accordingly. In particular, regarding my major concern (the sufficiency of the neurons for salt-"specific" memories), I think the authors found a good solution.

I have no further comments.

We sincerely thank the reviewer for the positive feedback on our revision. We are pleased that the revised manuscript meets the expectations and appreciate the time and effort invested in the review process.

Version 3

Abstract

Abstract

Dopaminergic neurons (DANs) carry out multiple tasks in the brain, including the transmission of information related to rewards and punishments across various animal species. They are responsible for evaluating sensory input, storing resulting associations as memory, and continuously updating them based on their relevance and reliability. Accurate comprehension of the dopaminergic system’s operation necessitates an understanding of the specific functions mediated by individual DANs. To this end, our research employs Drosophila larvae, which possess approximately 12,000 neurons in their brains, of which only around 1% (approximately 120) are DANs.

The presynaptic projections to the mushroom body (MB) - a brain region pivotal for associative olfactory learning in insects - are limited to only eight larval dopaminergic neurons. These DANs are further subdivided into two clusters: the primary protocerebral anterior medial cluster (pPAM) comprises four cells, and the dorsolateral 1 cluster (DL1) comprises the remaining four cells. Our findings confirm previous research that demonstrates that the pPAM DANs innervating the MB’s medial lobe encode for a gustatory sugar reward signal. Furthermore, we have identified four DANs in the DL1 cluster - DAN-c1, DAN-d1, DAN-f1, and DAN-g1 - each of which innervates distinct compartments of the MB peduncle, lateral appendix, and vertical lobe. Optogenetic activation of DAN-f1 and DAN-g1 alone suffices to substitute for punishment. Furthermore, optogenetic inhibition, calcium imaging results and electron microscopy-based reconstruction of all sensory input circuits to the four DL1 DANs demonstrate that each DAN encodes a different aspect of punishment, with DAN-g1 being of central importance for the salt dependent teaching signal.

To summarize, our investigation has revealed the existence of a cellular division of labor among larval DANs concerning the transmission of dopaminergic reward (pPAM cluster) and punishment signals (DL1 cluster). Individual DANs in each cluster encode for distinct but partially overlapping aspects of the teaching signal. The striking resemblance in the organizing principle of larval DANs with that of its adult counterpart and the mammalian basal ganglion suggests that there may be a limited number of efficient neural circuit solutions available to address more complex cognitive challenges in nature.

Introduction

To cope with a constantly changing environment, animals benefit from learning behavioral rules that need constant adaptation and updating. To test this ability, researchers often use Pavlovian or classical conditioning experiments (Pavlov 1927), which examine an organism’s capacity to form associations between sensory cues, called conditioned stimuli (CS), and rewards or punishments, referred to as aversive and appetitive unconditioned stimuli (US) or teaching signals (Gerber et al 2009, Glanzman 1995, Heisenberg 2003, Menzel 2022, Waddell 2013, Weber et al 2023b, Widmann et al 2018). Dopaminergic neurons (DANs) and other modulatory neurons mediate teaching signals that allow animals to classify a given CS into positive or negative valence based on past experience (Cognigni et al 2018, Menzel 2001, Schultz 2015, Thum & Gerber 2019, Watabe-Uchida et al 2017). For example, when a CS is paired with simultaneous activation of a specific set of DANs in the PAM cluster, larval or adult Drosophila subsequently perceive the odor as attractive (Burke et al 2012, Eschbach et al 2020, Liu et al 2012, Rohwedder et al 2016, Schroll et al 2006) – a similar effect was seen after optogenetic activation of DANs located in the ventral tegmental area of the mouse brain to induce conditioned place preference (Tsai et al 2009). In contrast, activation of larval DL1 DANs (Eschbach et al 2020, Schroll et al 2006, Weiglein et al 2021), adult PPL1 DANs (Aso et al 2012, Aso & Rubin 2016, Aso et al 2010, Claridge-Chang et al 2009), or DANs in the posterior tail of the striatum of mice were shown to reinforce avoidance learning (Menegas et al 2018). These examples illustrate that the dopaminergic system constitutes a highly conserved and fundamental neural principle of the brain, which has been retained throughout the process of evolution and executes comparable functions across diverse animal species, including even humans (Frick et al 2022).

Our research focuses on the Drosophila larva, an organism endowed with a relatively simple central nervous system, composed of approximately 12,000 neurons (Dumstrei et al 2003, Winding et al 2023). Classical conditioning in larvae can be analyzed by standardized behavioral assays (Almeida-Carvalho et al 2017, Apostolopoulou et al 2013, Scherer et al 2003, Weber et al 2023a), in conjunction with a wealth of genetic methods for targeting single molecules and neurons (Dietzl et al 2007, Li et al 2014, Perkins et al 2015, Port et al 2020). In addition, the connectome of the brain was recently reconstructed, which enables the extraction and analysis of related wiring motifs, as well as entire learning and memory circuits (Winding et al 2023). Based on these advances, it was possible to establish a more fundamental cellular and synaptic understanding of the larval memory circuit, the mushroom body (MB) network. The MB is a higher-order parallel fiber system present in numerous invertebrate brains, including hemimetabolous as well as holometabolous insects and their larval stages (Strausfeld et al 1998, Strausfeld et al 2020). Analogous to the vertebrate cerebrum, the MB in insects is responsible for forming and maintaining associations (Heisenberg 2003, Menzel 2001, Menzel 2022, Thum & Gerber 2019, Tomer et al 2010, Waddell 2013). In the first developmental stage (L1), the larval MB comprises approximately 110 intrinsic Kenyon cells (KC) per hemisphere, which sparsely encode for conditioned stimulus (CS) information, and is dominated by olfactory input (Eichler et al 2017, Masuda-Nakagawa et al 2009).

The transmission of olfactory information from the peripheral sensory organ is achieved via only 21 olfactory receptor neurons (ORNs) per body side, which in turn send signals across 35 uni- and multiglomerular projection neurons, ultimately reaching the MB KCs (Berck et al 2016, Fishilevich et al 2005, Kreher et al 2005, Masuda-Nakagawa et al 2009, Ramaekers et al 2005). KCs receive modulatory input from 17 neurons, consisting of eight dopaminergic neurons, four octopaminergic neurons, and five neurons with unknown signaling molecule identity (Eichler et al 2017, Rohwedder et al 2016, Saumweber et al 2018, Selcho et al 2009). The DANs are divided into two clusters based on the location of their soma and related neuronal lineage, namely the primary protocerebral anterior medial cluster (pPAM) and the dorsolateral 1 cluster (DL1). Each DAN innervates a specific and non-overlapping site in the MB, which, together with one to three output neurons, defines a distinct compartment (Eichler et al 2017). The larval MB comprises a total of 11 compartments, eight of which are defined by DAN input (Eichler et al 2017, Saumweber et al 2018). The four DANs of the pPAM cluster (DAN-h1, DAN-i1, DAN-j1, and DAN-k1) innervate the four related compartments (h, i, j, and k) of the medial lobe and provide teaching signals for sugar rewards (Eichler et al 2017, Rohwedder et al 2016, Schleyer et al 2020). The four DANs of the DL1 cluster (DAN-c1, DAN-d1, DAN-f1, and DAN-g1) innervate the lower peduncle (c), the lateral appendix (d), the intermediate (f), and the lower vertical lobe (g) (Eichler et al 2017, Saumweber et al 2018). Previous studies have shown that the DL1 cluster is capable of mediating aversive teaching signals (Eschbach et al 2020, Schroll et al 2006, Selcho et al 2009, Weiglein et al 2021). Moreover, larvae can use a variety of natural stimuli as aversive teaching signals such as heat, electric shock, vibration, bitter substances (like quinine or caffeine), and high salt concentrations (low salt concentrations, however, are rewarding) (Aceves-Pina & Quinn 1979, Apostolopoulou et al 2016, Apostolopoulou et al 2014, Eschbach et al 2011, Gerber & Hendel 2006, Khurana et al 2009, Khurana et al 2012, Niewalda et al 2008, Pauls et al 2010, von Essen et al 2011). However, it is currently unknown if and how these natural stimuli are represented by the DL1 cluster.

We only know that optogenetic stimulation of nociceptive sensory neurons and their downstream basin neurons has been observed to activate three out of the four DL1 DANs, namely DAN-d1, DAN-f1, and DAN-g1 (Eschbach et al 2020). Furthermore, the simultaneous presentation of odor with optogenetic activation of these three DANs has been found to elicit aversion in larvae (Eschbach et al 2020, Weiglein et al 2021). These findings suggest that the activation of DL1 DANs may induce a punishing effect during training, based on somatosensory perception. But can these four individual DANs also process other aversive sensory modalities such as chemosensory information? Utilizing electron microscopy-based reconstruction, all first and second order input neurons of the MB, consisting of a complete set of 102 neuron pairs, were examined (Eschbach et al 2020, Winding et al 2023). The analysis identified numerous afferent sensory inputs via 20 pairs of feedforward neurons (FFNs), which can now be integrated with the published wiring diagram of all chemosensory input neurons (Miroschnikow et al 2018). This enables a more comprehensive understanding of the distinct modes of action of the four DL1 DANs in relation to the processing of different sensory modalities.

Consequently, we have investigated the role of the four DL1 DANs in the larval brain regarding their involvement in encoding a salt-dependent aversive teaching signal at the single-cell level. Utilizing behavioral assays, fluorescence microscopy, calcium imaging experiments, and connectome analyses, we have established a more comprehensive description of the coding space of the DANs at the single-cell level. Our investigation unveils a cellular specialization among the four DL1 DANs, where individual DANs code for different but partially overlapping aspects of the aversive teaching signal

Results

The larval dopaminergic system is functionally bipartite to convey positive and negative teaching signals

There are eight DANs in the larval brain that innervate the mushroom body that can be anatomically subdivided into four pPAM cluster and four DL1 cluster neurons (Figure 1A-C and Figure 1- figure supplement 1 A-C). Three of the four pPAM cluster DANs can be labeled with the driver line R58E02-Gal4 (Figure 1- figure supplement 1A-E) (Rohwedder et al 2016). Pairing an odor presentation with simultaneous optogenetic blue light activation of R58E02 positive DANs using a mutant microbial-type rhodopsin ChR2XXL, which depolarizes neurons artificially (Dawydow et al 2014), establishes an appetitive olfactory memory (Figure 1- figure supplement 1F and G) (Rohwedder et al 2016). Please be aware that the activation of blue light itself does not cause any disruptive behavioral side effects, as it does not independently induce memory formation. In line with this finding, it was shown that ablation or inhibition of R58E02 positive DANs strongly reduces appetitive olfactory memory (Rohwedder et al 2016) but has no functional significance for aversive olfactory memory (Figure 1- figure supplement 1H and I) (Rohwedder et al 2016). Therefore, it can be concluded that pPAM cluster DANs are not involved in the encoding of aversive teaching signals. How are aversive teaching signals encoded?

The larval dopaminergic system is subdivided in two functionally distinct clusters.

(A) The larval dopaminergic neurons (DANs) can be anatomically subdivided into the primary protocerebral anterior medial (pPAM) and dorsolateral 1 (DL1) cluster based on their cell body position. (B) The DL1 cluster (cell bodies in green and purple) consists of four DANs providing input to the c, d, g, and f compartment of the vertical lobe, peduncle and lateral appendix of the MB (in grey). (C) Schematic representation of the larval MB using single letter acronyms based on (Saumweber et al 2018) to indicate compartment innervation by MB input and output neurons. DL1 DANs specifically innervate the compartments c, d, f and g in the vertical lobe, lateral appendix and peduncle, whereas pPAM DANs innervate the shaft (h) and medial lobe (I, j, k) of the MB. (D,E) Four DL1 DANs (DAN-c1, DAN-d1, DAN-g1, and DAN-f1) are included in the expression pattern of the TH-GAL4 driver line. However, expression of a UAS-mCD8::GFP reporter via TH-GAL4 labels many more neurons (in green, anti-GFP) throughout the entire CNS (in red and blue, anti-N-cadherin and anti-discs large), in total about 100 neurons. (F) To test whether optogenetic activation of the DL1 DANs is sufficient to substitute for a punishment, we used the TH-GAL4 driver in combination with UAS-ChRXXL. Experimental and control larvae were trained by simultaneously presenting and odor and blue light and thus artificial activation of DL1 DANs, whereas a second odor was presented in darkness. Only larvae of the experimental genotype (p > 0.05, N=15), but not of the two genetic controls (both p < 0.05, for each group N=15), retrieve an aversive associative olfactory memory. The same result was seen after one or three training trials (F and G, respectively). (H, I) To test for the acute function of the DL1 DANs in aversive associative olfactory memory, we expressed GtACR2 via the TH-GAL4 driver. Acute optogenetic inhibition of synaptic output from DL1 and other DANs reduced odor high salt memory. Experimental larvae raised on supplemented food (0.5 mM all-trans-retinal, ATR) and trained in blue light performed on a lower level than control animals kept on standard food (p < 0.05, for each group N=15). A memory impairment was seen after one and three training trials (G and H, respectively). All behavioral data is shown as box-plots. Differences between groups are highlighted by horizontal lines between them. Performance indices different from random distribution are indicated below each box-plot. The precise sample size is given below each box-plot. n.s. p > 0.05; * p < 0.05. Scale bars: in C 50 µm and in D 25 µm. The source data and results of all statistical tests are documented in the Supplemental Information file.

The driver line TH-Gal4 can be used to label the four DANs of the DL1 cluster (Figure 1A-E), as well as most other DANs, except for a small subset of DANs, such as pPAM DANs and neurons in the subesophageal zone (Figure 1D and E) (Selcho et al 2009). By coupling odor presentation with simultaneous optogenetic activation of TH-Gal4 positive DANs through ChR2XXL, an aversive olfactory memory can be established (Schroll et al 2006), which was observed after both one (Figure 1F) and three (Figure 1G) odor-DAN activation pairings. Consistent with these findings, aversive associative memory formed after one or three training cycles of odor high salt pairings were significantly reduced when DANs activity was inhibited optogenetically via GtACR2 (Figure 1H and I), a light-gated channel from the algae Guillardia theta (Mohammad et al 2017). The observed partial reduction may be attributable to biological, technical, or systemic factors, or a combination of these (a more detailed description can be found in the corresponding part of the discussion).In these experiments salt has to be added during the test to initiate the recall of the aversive memory (indicated as red test plates, Figure 1F-I) (Gerber & Hendel 2006, Schleyer et al 2015, Schleyer et al 2011). Previous studies have shown that the larva only recalls aversive odor-high salt memory when tested in the presence of the teaching signal. The prevailing explanation suggests that the larva engages in memory-based search during the test, but only if the memory content can be used in a beneficial way (Gerber & Hendel 2006, Schleyer et al 2015, Schleyer et al 2011).

Based on our findings and previously published data on both larval and adult stages, we propose a functional division of labor among DANs projecting to the MB. Specifically, pPAM neurons are likely to convey a dopaminergic appetitive teaching signal, whereas DL1 neurons may transmit a dopaminergic aversive teaching signal. Nevertheless, the findings related to DL1 DANs necessitate additional validation as the TH-Gal4 driver line used in the experiment labels approximately 120 cells, thus limiting the ability to draw cell-specific conclusions.

Anatomical single-cell analysis of DL1 DANs

In an earlier investigation, we conducted a screening of several thousand Gal4 lines in order to construct a high-resolution atlas of the mushroom body at the cellular level via light microscopy (Saumweber et al 2018). A subsequent reconstruction of the wiring diagram of the entire MB at the electron microscopic level encompassed not only individual neurons but all synaptic connections (Eichler et al 2017). Utilizing this anatomical work, we identified a total of 102 Gal4 lines featuring MB extrinsic neuron expression patterns. These lines were subjected to intersectional strategies in order to restrict expression patterns to single neurons (Eschbach et al 2020, Saumweber et al 2018). For the present study, we employed eight split-Gal4 lines chosen from this set, as they contained only one or two DANs of the DL1 cluster (Figure 2).

Anatomical single cell analysis of DL1 DAN specific split-GAL4 driver lines.

(A) The larval MB is organized into 11 compartments: CX calyx; IP and LP intermediate and lower peduncle; LA lateral appendix; UVL, IVL, and LVL upper, intermediate, and lower vertical lobe; SHA, UT, IT, LT shaft as well as upper, intermediate, and lower toe of the medial lobe. Single-letter synonyms of compartment names are given as “a–k”. These letters are used to indicate compartment innervation by the MB input and output neurons (Saumweber et al 2018). DL1 cluster DANs are DAN-c1, DAN-d1, DAN-g1, and DAN-f1 that innervate the respective four different compartments of the MB. (B-J) Individual split-Gal4 driver lines were crossed with the reporter strain UAS-mCD8::GFP;mb247-lexA,lexAop-mRFP. Third instar larval brains were dissected, fixed and mounted to visualize the fluorescent reporter signal labeling the MB (mb247-lexA,lexAop-mRFP shown in magenta) and the respective DAN pattern (GFP shown in green). (B-E) SS02160 (DAN-c1), MB328B (DAN-d1), SS02180 (DAN-f1), SS01716 (DAN-g1) each specifically label a single DL1 DAN (cell bodies are highlighted by white arrowheads). (F–K) Two neurons can be seen in MB065B, SS01702, and MB054B split-Gal4 that express in DAN-c1/DAN-f1, DAN-c1/MBIN-e1, and DAN-f1/DAN-g1. Please note that MB065B shows strong expression DAN-f1 but weaker staining in DAN-c1. (H–K) MB054B showed reliable strong expression in DAN-f1 and DAN-g1. In some brains a third weak cell body was visible right next to the other two DANs (I; GFP channel inverted and shown in black; J; cell body highlighted with gray arrowhead). Due to the low expression level we were not able to identify this cell given that only the g and f compartment of the MB were innervated (H). (K) Analysis of the entire brain via native fluorescence expression of GFP (green) and n-Syb (magenta) did not reveal additional cells for MB054B split-Gal4. Scale bars: (B-J) 20 µm, (K) 50µm.

DAN-c1 was specifically labeled by SS02160 (Figure 2B), DAN-d1 by MB328B and MB143B (Figure 2C and Figure 2- figure supplement 1B), DAN-f1 by SS02180 and MB145B (Figure 2D and Figure 2- figure supplement 1C) and DAN-g1 by SS01716 (Figure 2E). In addition, we used the split-Gal4 lines MB065B (Figure 2F), SS01702 (Figure 2G) and MB054B (Figure H-K) that specifically labeled DAN-c1/DAN-f1, DAN-c1/MBIN-e1 or DAN-f1/DAN-g1, respectively. For our evaluation we crossed each split-Gal4 line with the reporter strain UAS-mCD8::GFP;mb247-LexA,lexAop-mRFP (Burke et al 2012) that allowed us to monitor the expression pattern (green) in the background of a MB specific reference staining (magenta). Our results confirm the reported specificity even despite the use of different reporter lines and lacking antibody staining (Eschbach et al 2020), as we analyzed only endogenous GFP expression using a more cautious and faster whole mount staining protocol (see Material and Methods). In addition, we also verified that the DANs labeled via MB054B are indeed dopaminergic as the GFP expression overlaps with an anti-TH antibody staining (Figure 2- figure supplement 2). Having established anatomical validity, we subsequently employed these lines to explore the physiological response of each DAN within the DL1 cluster to sugar and salt stimuli. For the purpose of analyzing DAN-d1 and DAN-f1, we limited our investigation to the split-Gal4 lines MB328B and SS02180 as these lines displayed more robust expression levels.

Physiological single-cell analysis of DL1 DANs

In order to investigate how high salt teaching signals are represented at the level of individual DL1 DANs, we conducted calcium imaging experiments using a microfluidic chip on intact and immobilized larvae, as previously described (Si et al 2019). We expressed GCaMP6m under the control of four different split-Gal4 lines specific to DL1 DANs (SS02160 (DAN-c1), MB328B (DAN-d1), SS02180 (DAN-f1), and SS01716 (DAN-g1)), as well as the pPAM specific R58E02-Gal4, and exposed the larvae to solutions containing 100 mM salt, 1 M salt, or 500 mM fructose. Our results showed that only DAN-c1 responded to 100 mM salt (Figure 3A). While DAN-c1, DAN-d1, and DAN-g1 responded to 1 M salt (Figure 3A, B, and D). Conversely, pPAM DANs showed a reduced signal in response to 1 M salt (Figure 3E). Both DAN-c1 and pPAM DANs showed a calcium increase in response to 500 mM fructose (Figure 3A and E). Although DAN-f1 did not respond to either salt concentration, a reduction in the calcium signal was seen for 500 mM fructose (Figure 3C). Overall, these results provide support for the functional division of the larval dopaminergic system at the physiological level, where most DL1 neurons are activated by high salt concentrations, with one exception inhibited by fructose. In contrast, pPAM cluster neurons display the opposite response, being inhibited by high salt concentrations and strongly activated by fructose.

Calcium responses of DANs to gustatory stimulation.

Four different split-Gal4 lines and the R58E02 driver line were crossed with UAS-GCaMP6m to express a calcium reporter in DANs. The responses of each DAN towards 100 mM NaCl, 1 M NaCl, and 500 mM fructose was tested in intact larvae using a microfluidic chip-based setup. (A) Calcium responses in DAN-c1 (N= 5) were induced by gustatory stimulation with 100 mM NaCl (orange), 1 M NaCl (red) and 500 mM fructose (green, for all p < 0.05). (B) DAN-d1 (N=5) calcium responses were only seen after 1 M NaCl stimulation (red, p < 0.05), but not after 100 mM NaCl (orange, p > 0.05) and 500 mM fructose stimulation (green, p > 0.05). (C) Stimulation with 100 mM (orange, p > 0.05) and 1 M NaCl (red, p > 0.05) did not induce calcium responses in DAN-f1 (N=6). However, stimulation with 500 mM fructose reduced the obtained calcium signal (green, p < 0.05). (D) 1 M NaCl (red, p < 0.05), but not 100 mM NaCl (orange, p > 0.05) and 500 mM fructose (green, p > 0.05), induced a calcium response in DAN-g1 (N=7). (E) pPAM DANs calcium responses (N=5) were only seen after 500 mM fructose stimulation (green, p < 0.05). Stimulation with low (orange) and high salt concentrations (red) did not increase calcium signals, however both reduced pPAM activity (p < 0.05). Each graph shows the mean calcium signal plotted as the relative response strength ΔF/F and the related standard error of the mean on the y axis. The time in seconds is given below each graph on the x axis. The grey box indicates the duration of the stimulus application. The sample size of each group (N=5-7) is given above each row. n.s. p > 0.05; * p < 0.05. The source data and results of all statistical tests are documented in the Supplemental Information file.

Behavioral single-cell analysis of DL1 DANs

Next, we analyzed the role of single DL1 DANs in aversive olfactory learning (Figure 2- figure supplement 1D-G, Figure 4- figure supplement 1-4). We employed split-Gal4 lines SS02160 (DAN-c1), MB328B (DAN-d1), SS02180 (DAN-f1), and SS01716 (DAN-g1) to express the apoptosis proteins Hid and Reaper, resulting in the ablation of the respective individual DL1 DANs. The ablated animals were subjected to the aversive odor-high salt memory assay, using one (Figure 4- figure supplement 1A-D) and three training cycles (Figure 4- figure supplement 1E-H). We utilized two different training protocols to differentiate between the effects on short-term memory (STM) and anesthesia-resistant memory (ARM), as previous studies have demonstrated that larvae exhibit both after one training cycle, whereas only ARM is present after three training cycles (Widmann et al 2016). However, we found that the ablation of individual DL1 DANs had no effect on aversive olfactory memories in any of the eight experiments conducted (Figure 4- figure supplement 1A-H). For DAN-d1 and DAN-f1 we verified these results by using the additional split-Gal4 lines MB143B and MB145B (Figure 2- figure supplement 1D-G). In addition, ablation of DAN-d1, DAN-f1 and DAN-g1 individually did not have any impact on odor-quinine and odor-fructose memories established after one cycle training (Figure 4- figure supplement 2 and 3). Moreover, MB input neurons that are not dopaminergic (OAN-a1, a2, MBIN-b1, b2, and OAN-e1) were also found to have no relevant effect (Figure 4- figure supplement 4). These results suggest that although DL1 DANs respond to high salt concentrations (Figure 3), they are not individually necessary in ablation experiments for mediating the aversive teaching signal. The data suggest that the aversive teaching signal may be distributed across multiple neurons, indicating possible functional redundancy or, alternatively, developmental compensation that could attenuate the effects of DL1 cell ablation. However, it is possible that there are developmental compensatory effects for DL1 cluster DANs, allowing the remaining three neurons (or other cells) to take over the function of the ablated one. This hypothesis was analyzed later in optogenetic inhibition experiments.

Ablation of DAN-f1 and DAN-g1 together impairs aversive olfactory memory.

In all panels associative performance indices are shown for tests immediately after classical conditioning. In the upper panels (A-C) larvae are trained once by pairing an olfactory stimulus with high salt punishment, whereas in the lower panels (D-F) three training cycles were applied. Schematic overviews for both conditioning protocols are shown on the left. The three different DL1 DAN specific split-Gal4 driver MB054B, MB065B, and SS01702 that each label two neurons were crossed to the effector UAS-hid,rpr to induce apoptosis (A-F). (A) With MB054B used as driver strain to ablate the DL1 DAN combination DAN-f1/DAN-g1, the aversive associative performance index of the experimental group was decreased compared to both controls (p < 0.05, N=15 for each group). (B-F) In all other experiments ablation of different DL1 DAN combinations did not reveal a phenotype. (in all experiments at least one or even both control groups are compared to the experimental group p > 0.05, for each group N=15-25). Please note, that this also includes MB054B crossed with UAS-hid,rpr tested after three training trials (D). (G) To verify the memory phenotype of MB054B crossed with UAS-hid,rpr tested after one trail conditioning, we repeated the experiment using the odor pair hexyl acetate (HA) and benzaldehyde (BA). Again, experimental larvae tested after one trial learning showed a robust decrease in odor-high salt memory when compared to both genetic control groups (p < 0.05, for each group N=15). The memory phenotype was not seen after three training trials (J, p > 0.05, for each group N=15). (H, K) With MB054B used as driver strain to ablate the DL1 DAN combination DAN-f1/DAN-g1, aversive odor-quinine memory was not impaired after one or three cycle conditioning (p > 0.05 when comparing experimental and control groups, for each group N=15). (I, L) Similarly, appetitive odor-fructose learning after one and three cycle conditioning was not impaired when ablating DAN-f1/DAN-g1 (p > 0.05 when comparing experimental and control groups, for each group N=15). All behavioral data is shown as box-plots. Please note that the same data set for the effector control was plotted in subfigures E and F as respective behavioral experiments have been conducted in parallel. Differences between groups are highlighted by horizontal lines between them. Performance indices different from random distribution are indicated below each box-plot. The precise sample size of each group is given below each box-plot. n.s. p > 0.05; * p < 0.05. The source data and results of all statistical tests are documented in the Supplemental Information file.

The combination of DAN-f1 and DAN-g1 is required to establish aversive odor-high salt memories

In order to investigate a possible redundancy among DANs, we proceeded by using split- Gal4 lines MB054B (DAN-f1/DAN-g1), MB065B (DAN-c1/DAN-f1), and SS01702 (DAN-c1/MBIN-e1) to express apoptosis proteins Hid and Reaper. This approach selectively ablated either two DL1 DANs or one DL1 DAN and another MB input neuron with unknown neurotransmitter identity. It should be noted that no additional split-Gal4 lines are presently available for further DL1 DAN combinations or the complete DL1 cluster. We then subjected the larvae to the aversive odor-high salt memory assay, using one (Figure 4A-C) and three (Figure 4D-F) training cycles. Only larvae with ablated DAN-f1/DAN-g1 exhibited significantly reduced aversive olfactory memory after one cycle training (Figure 4A). No memory reduction was observed when ablating DAN-c1/DAN-f1 or DAN-c1/MBIN-e1 (Figure 4B and C). Ablation of all three sets of neurons, including MB054B, had no effect on aversive olfactory memory after three training cycles (Figure 4D-F). To ascertain the independence of the DAN-f1/DAN-g1 neuron-mediated impairment of aversive olfactory memory after one cycle training from the specific odors used, hexyl acetate and benzaldehyde were employed instead of amyl acetate and benzaldehyde. The results of these experiments confirmed the previous findings, as ablation of DAN-f1/DAN-g1 neurons significantly reduced aversive odor-high salt memory after one cycle training (Figure 4G), but not after three cycles (Figure 4J). Innate olfactory and gustatory behavior was not impaired by DAN-f1/DAN-g1 ablation (Figure 4- figure supplement 5).

Subsequently, we investigated the selectivity of DAN-f1/DAN-g1 neurons concerning the nature of the aversive teaching stimulus. To this end, we utilized quinine as an aversive teaching signal, a stimulus that has been demonstrated to be learnable by the larva, as evidenced by previous studies (Apostolopoulou et al 2014, El-Keredy et al 2012). Removal of the DAN-f1/DAN-g1 neurons using the split-Gal4 line MB054B had no effect on aversive odor-quinine memory independent of the number of training cycles (Figure 4H and K). Furthermore, ablation of DAN-f1/DAN-g1 did not alter appetitive odor-fructose memories after one or three training cycles (Figure 4I and L). Therefore, the DL1 DAN-f1/DAN-g1 combination seems to be specifically required for aversive odor-high salt memory. This is further supported by anatomical data that verified the effectiveness of Hid and Reaper using the MB054B split-Gal4 line (Figure 4- figure supplement 6).

Individual DL1 DANs can instruct an aversive memory recalled on a salt plate

In order to determine if the activation of individual DL1 DANs signals aspects of the natural high salt punishment, we conducted a one training trial learning experiment using ChR2XXL to artificially activate DAN-c1, DAN-d1, DAN-f1, DAN-g1, and the combination of DAN-f1/DAN-g1. We used split-Gal4 lines SS02160, MB328B, SS02180, SS01716, and MB054B, respectively (Figure 5). Simultaneously activating DAN-f1/DAN-g1 via ChR2XXL during an odor presentation establishes an aversive olfactory memory (Figure 5E and F). These results are comparable to those observed for TH-Gal4 (as shown in Figure 1F and G). Additionally, this memory is observed after one (Figure 5E) and three training cycles (Figure 5F). Furthermore, single-cell analysis indicates that individual activation of DAN-f1 and DAN-g1, but not of DAN-d1 and DAN-c1 (please note that experimental animals are different from zero and significantly different to one control), can induce an aversive odor memory (Figure 5A-D).

Optogenetic DL1 DAN activity can substitute for salt punishment.

In all panels associative performance indices are shown for tests immediately after classical conditioning. In panels (A-E) larvae are trained once by pairing an olfactory stimulus with artificial blue light activation, whereas in panel (F) three training cycles were applied. Schematic overviews for both conditioning protocols are shown to the left of (A) and (F). (A-D) To test whether optogenetic activation of the individual DL DANs DAN-c1, DAN-d1, DAN-f1, and DAN-g1 is sufficient to substitute for a punishment, we used the split-Gal4 lines SS02160, MB328B, SS02180, and SS01716 in combination with UAS-ChR2XXL. (E, F) For simultaneous optogenetic activation of DAN-f1/DAN-g1 we used MB054B. Larvae of the experimental genotypes for DAN-c1, DAN-f1, DAN-g1 and the DAN-f1/DAN-g1 combination (for all p < 0.05, for each group N=15), but not for DAN-d1 (N=15) and all genetic controls (for all p > 0.05, for each group N=15), showed an aversive associative memory. The results imply that in the tested conditions, the punishment signal can be mediated by the artificial activation of all individual DL1 DANs, with the exception of DAN-d1. All behavioral data is shown as box-plots. Please note that the same data set for the effector control was plotted in subfigures B-E as respective behavioral experiments have been conducted in parallel. Differences between groups are highlighted by horizontal lines between them. Performance indices different from random distribution are indicated below each box-plot. The sample size of each group (N=15) is given below each box-plot. n.s. p > 0.05; * p < 0.05. The source data and results of all statistical tests are documented in the Supplemental Information file.

An individual DL1 DAN is acutely necessary to establish an aversive odor-high salt memory

Next, we examined the extent to which the individual DL1 DANs are acutely required for the transmission of a functional aversive teaching signal. We have already shown that the combination of DAN-f1/DAN-g1 is required for aversive odor-high salt memory (Figure 4A and G). However, in these experiments DANs were removed during the entire development, which may have led to compensatory mechanisms or side effects.

To overcome this limitation, we left the individual DL1 DANs intact during embryonic and larval development and used GtACR2 to selectively inhibit DAN activity only during training, thus isolating the teaching signals from other DAN functions. Optogenetic inhibition of individual DL1 DANs showed no effect on aversive odor-high salt memories (Figure 6A-C), except for DAN-g1 that led to a small but significant reduction (Figure 6D). Optogenetic inhibition of DAN-f1/DAN-g1 resulted in a similar reduced aversive odor-high salt memory. This was the case after one training cycle (Figure 6E) and three training cycles (Figure 6F). Thus, the learning deficit in this experiment was more pronounced than after ablation of these two DL1 DANs throughout development (Figure 4D).

Optogenetic inhibition of DL1 DAN activity impairs aversive olfactory memory.

In all panels associative performance indices are shown for tests immediately after classical odor high salt conditioning. In panels (A-E) larvae are trained once by pairing an olfactory stimulus with an aversive high salt stimulation, whereas in panel (F) three training cycles were applied. Schematic overviews for both conditioning protocols are shown to the left of (A) and (F). (A-D) To test whether optogenetic inhibition of the individual DL DANs DAN-c1, DAN-d1, DAN-f1, and DAN-g1 during training impairs punishment signaling, we used the split-Gal4 lines SS02160, MB328B, SS02180, and SS01716 in combination with UAS-GtACR2 and blue light stimulation during the entire training phase. (E, F) For simultaneous optogenetic inhibition of DAN-f1/DAN-g1 we used MB054B. (A-C) Larvae with inhibited DAN-c1, DAN-d1, or DAN-f1 function during training showed no impairment of odor-high salt memory comparable to controls that were kept on standard food without supplemented all-trans-retinal (ATR, 0.5 mM) and received the same protocol (for all p > 0.05, N=15 for each group). (D-F) In contrast inhibition of DAN-g1 alone, or the combination of DAN-f1/DAN-g1 after single trial and three trial conditioning impaired odor-high salt memory compared to controls (for all p < 0.05, for each group N=15). This shows that DAN-g1 function is of central importance for signaling a salt punishment teaching signal. All behavioral data is shown as box-plots. Differences between groups are highlighted by horizontal lines between them. Performance indices different from random distribution are indicated below each box-plot. The precise sample size of each group is given below each box-plot. n.s. p > 0.05; * p < 0.05. The source data and results of all statistical tests are documented in the Supplemental Information file.

A synaptic reconstruction of all chemosensory inputs for each individual DL1 and pPAM DAN

Certainly, a comprehensive analysis from the sensory neurons responsible for salt perception, via the interneurons involved in salt information processing, up to the dopaminergic DL1 teaching neurons would be highly desirable to understand why and how the different DANs respond to salt. Regrettably, such an analysis is currently not feasible due to our limited understanding of salt perception and which sensory cells encode this crucial information (Liu et al 2003). Therefore, this question cannot be answered by the larval connectome. However, it is possible to analyze how qualitative differences between individual DL1 DANs (Figure 3-6) are represented in the connectome, since not all DANs code equally for the punishment teaching signal. Therefore, we compared the input wiring diagrams of the individual DANs of the pPAM and DL1 clusters by mapping the synaptic connections from sensory neurons - to interneurons - to individual DANs, using the full larval brain EM volume (Figure 7, Figure 7- figure supplement 1) (Eschbach et al 2020, Miroschnikow et al 2018, Winding et al 2023).

Interneurons and hub analysis of sensory to DAN pathways.

(A, B) Schematic graph representation of individual DL1 (A) and pPAM (B) DANs. The outer ring at the bottom of each scheme represents the sensory composition of neurons targeting sensory-to-DANs interneurons (orange circle). The type of sensory information is encoded by the respective color (ACa: anterior central sensory compartment, AVa: ventral anterior sensory compartment, ACal: lateral anterior central sensory compartment, ACp: posterior anterior central sensory compartment, ACpl: posterior-lateral anterior central sensory compartment, VM: ventromedical sensory compartment, TD CO2: tracheal dendritic neurons responding to CO2, ORNs: olfactory receptor neurons). DAN input neurons are shown that get no direct sensory input (grey circles). Individual DANs are shown in the middle of the scheme as black circles. They are connected to MB KC (purple circles), which in turn connect to mushroom body output neuron (MBONs; dark red circles). Arrows indicate the direction of the synaptic connection and its strength (coded by arrow thickness). Numbers in circles indicate number of neurons. The percentages indicate the proportion relative to the total input that the cell receives from the specific neuronal partners. For example, DAN-f1 receives 17% of its input from six different sensory-to-DAN interneurons (yellow circle), 17% from 19 other, non-sensory interneurons, 41% from 64 MB KCs, and 2% from two MBONs. (C) Dot plot showing the importance of interneurons acting as sensory to DAN hub. Dot size was calculated using the fraction of total input an interneuron receives from sensory neurons multiplied by the fraction of total input this interneuron gives to an DAN. Colored backgrounds of dots are highlighted in orange for the connections with a hub size of 0.001 or above. (D, E) Schematic of graph representation. The outer ring represents the sensory composition of neurons targeting upstream neurons of DANs. The type of sensory information is encoded by the respective color. Synaptic threshold for upstream interneurons of DANs = 3 and of upstream sensory neurons = 1. Line thickness to interneurons and targets represents the percentage of synaptic input. White or orange circles connected to the outer ring represent the interneuron layer. The inner ring represents individual target neurons, DL1 and pPAM DANs. The identity of each DAN and interneuron is given by the label in its related circle. (F) EM reconstruction of DL1 and pPAM DANs (grey) highlighting their presynaptic sites in red (for DL1 DANs) and green (for pPAM DANs). At the top a horizontal view of the brain is shown. At the bottom a frontal view of the brain is shown.

We found that DAN-c1, DAN-d1, DAN-f1, and DAN-g1 (black circle) clearly differ in their downstream connections. They do however receive input from 1, 6, 6 or 11 sensory-to-DAN interneurons (orange circle), respectively (Figure 7A), which in turn means that these DL1 DANs are connected to 3, 42, 35, and 9 sensory neurons (numbers in colored circuits below). Please note that we restricted our analysis on direct sensory-to-DANs interneurons (single hop analysis) and did not expand the analysis onto two hop or three hop connections to keep the evaluation restricted to the shortest paths from sensory neurons to DANs. There is no direct input from sensory neurons onto DL1 DANs. Sensory input thus appears to be increasingly instructing DAN-c1, DAN-d1, DAN-f1, and DAN-g1. The analysis also shows that all four DL1 DANs receive information from all gustatory senses via these interneurons (coded by the yellow color for sensory composition). In addition, DAN-d1, as well as DAN-f1, receive information from olfactory and thermosensory neurons (Figure 7D). In general, less processed sensory input seems to be more important for DAN-f1 and DAN-g1 than for DAN-c1 and DAN-d1, as they get 17% and 11% of their entire input from these sensory-to-DAN interneurons in contrast to 1% and 5%, respectively. The four DL1 DANs show a conserved connection motif, as they get additional input from other upstream neurons (US) that are not directly linked to sensory neurons (grey circles; between 17% and 37%), as well as from MB KCs (purple circles; between 36% and 49%) and little MBON feedback (dark-red circle; between 1% and 2%). Examination of the different input types of DL1 DANs presented here, shows that DAN-f1 and DAN-g1 receive the highest proportion of sensory input relative to other DANs. Even by looking at the sensory-to-DAN interneurons individually, one can again see the familiar pattern of differences and partial overlap (Fig. 7D). DAN-d1, DAN-f1 und DAN-g1 all get input from interneuron pair 1 in contrast to DAN-c1. However again, DAN-f1 and DAN-g1 are most strongly connected to this neuronal pair (indicated by the thickness of the lines). DAN-f1 and DAN-g1 are therefore also the most similar in terms of their gustatory input (due to their connection via interneuron pair 1), despite all the clearly recognizable differences.

For the pPAM DANs we found the same general connectivity principles as for DL1 DANs regarding input from MBONs, KCs, and non-sensory upstream neurons (Fig 7B). Please note that the DAN-h1 neuron is not present at the early L1 EM brain volume and thus cannot be included in this evaluation (Figure 7B and E). In contrast, however, it is striking that pPAM DANs seem to get input from more interneurons connected to the sensory system (up to 16 for DAN-j1) signaling additional internal somatosensory and CO2 information. A closer inspection of the sensory-to-DAN interneurons showed that 27 of the 35 neurons are specifically innervating only DL1 or pPAM DANs (Figure 7C). Only 8 out of 35 interneurons connect to both clusters. Based on a calculated hub score sensory-to-DAN interneurons pair 1 (FB2IN-12) and pair 5 (FFN-21) were likely most significant for instructing DL1 DANs with sensory information. Whereas pairs 10 (FFN-17) and 11 (FFN-12) were – based on connectivity – most important in instructing DL1 and pPAM DANs. For the three pPAM DANs sensory-to-DAN interneuron pairs 16, 17, 18, 19 and 20 resulted in the highest hub scores (FFN-37, FFN-4, FFN-5, FFN-18 and FB-4).

The difference in cellular connectivity of DL1 and pPAM DANs is also reflected in different local input brain hubs: axo-dendritic synapses of sensory-to-DAN interneurons with DL1 DANs formed a characteristic hook-shape in the superior protocerebrum; in contrast, pPAM DANs get synaptic input from sensory-to-DAN interneurons in the anterior inferior protocerebrum (Figure 7F).

Discussion

A functional assignment for individual DANs of the DL1 cluster in their ability to encode a salt punishment teaching signal

The discovery of DL1 neurons as aversive teaching signal mediators in larval Drosophila (Eschbach et al 2020, Schroll et al 2006, Selcho et al 2009, Weiglein et al 2021) complements previous findings that demonstrate the sufficiency of four DANs from the pPAM cluster as an appetitive reinforcement signal in these animals (Rohwedder et al 2016). This dichotomy demonstrates an organizational principle shared by adult Drosophila and larvae (Aso et al 2012, Aso & Rubin 2016, Aso et al 2010, Burke et al 2012, Claridge-Chang et al 2009, Das et al 2014, Huetteroth et al 2015, Liu et al 2012, Schleyer et al 2020, Schroll et al 2006, Selcho et al 2009, Weiglein et al 2021, Yamagata et al 2015), despite a significant reduction in cell numbers. The principle appears to apply to mammals as well (Groessl et al 2018, Lammel et al 2012, Menegas et al 2018, Schultz 2015). But how do individual DL1 neurons contribute to the overall teaching signal? Is the signal evenly distributed across multiple cells, supporting the mass action hypothesis, or do individual DANs specifically encode for the salt stimulus (Rohwedder et al 2016)? Our results show that the four DL1 DANs perform this function in a mixture of both forms. There is no unique labeled line because both ablation of each individual DAN (Figure 2- figure supplement 1D-G, Figure 4- figure supplement 1) and the acute optogenetic inhibition (Figure 6) do not abolish the odor-high salt memory. The teaching signal is encoded in a redundant manner within the DL1 cluster, yet its distribution among the four DANs is non-uniform and relies heavily on the functionalities of DAN-g1 and DAN-f1 (Table 1, Figure 4-6). Each DAN possesses a distinct identity, which is defined by its specific, albeit partially overlapping, neuronal input circuitry and physiological response to salt (Table 1, Figure 3 and 7).

DAN-c1 responds to both positive and negative gustatory stimuli (Figure 3A). However, the activity of the cell is not sufficient or necessary for aversive memory (Figure 4- figure supplement 1A and E, Figure 5A and Figure 6A). These results suggest that DAN-c1 activity does not encode a taste-specific teaching signal, which is supported by EM analysis, as this particular cell is only weakly connected to three sensory neurons via a single sensor-to-DAN interneuron (Figure 7A and D). However, this interpretation might depend on the type and duration of DAN stimulation, as it was recently shown that - as in our case - optogenetic activation does not induce aversive memory, but thermogenetic 30 minute stimulation in a non-reciprocal one trail paradigm leads to aversive memory (Qi et al 2024). In addition, it could also be that the function of DAN-c1 is more to integrate state-dependent network changes, similar to the role of some DANs in innate and state-dependent preference behavior of adults (Boto et al 2019, Cazale-Debat et al 2024, Cohn et al 2015, Lewis et al 2015, Mohammad et al 2024, Siju et al 2020). DAN-d1 activity appears to have a negligible impact on the formation of odor-high salt memories in the paradigms we tested. Even though the neuron responds to high salt concentrations (Figure 3B) and receives input from 42 sensory neurons via six sensor-to-DAN interneurons (Figure 7A and D), its activity did not induce an aversive memory (Figure 5B) and was not found to be essential for odor-high salt memory (Figure 2- figure supplement 1D and E, Figure 4- figure supplement 1B and F, Figure 5B). Nonetheless, it is possible that DAN-d1 could instruct other salt-dependent memory types, established after non-associative training, second-order conditioning, or reversal learning (Kacsoh et al 2015, Konig et al 2019, Mancini et al 2019, Paranjpe et al 2012, Tabone & de Belle 2011, Yamada et al 2023). Other learning experiments in larvae (Eschbach et al 2020, Weiglein et al 2021) demonstrate that DAN-d1 activation can indeed guide the formation of aversive odor memory. However, these studies employ three training repetitions rather than a single repetition, as in our approach. Additionally, they utilize absolute conditioning with a single odor, in contrast to our protocol of differential conditioning involving two odors. Furthermore, their testing does not incorporate salt as an aversive gustatory stimulus. The observed differences in results suggest that DAN-d1 activation may only establish aversive memory if the stimulation is applied multiple times and is more distinctly presented during training. . In combination, DAN-f1 activity and DAN-g1 seem to have a crucial function in high salt punishment. DAN-f1 is inhibited by sugar (Figure 3C) and obtains input from 35 sensory neurons through six sensor-to-DAN interneurons (Figure 7A and D). Despite ablation and acute suppression of activity having no impact on the retention of a memory related to a high salt stimulus (Figure 2- figure supplement 1C, E and G, Figure 4- figure supplement 1C and G, Figure 6C), DAN-f1 activity can instruct the formation of an aversive memory (Figure 5C). However, there is the possibility that the impact of the ablation experiments could be more pronounced than what is observed here, given the potential for developmental compensation. This has been anatomically demonstrated in the adult dopamine mutants by serotonergic cells (Niens et al 2017). The identified function for DAN-f1 activity is consistent with previous research, although their experimental designs differ from our own in the test situation (Eschbach et al 2020, Weiglein et al 2021). Consequently, it is plausible that DAN-f1 activity encodes for a teaching signal that carries additional information beyond high salt. It is evident that DAN-g1 exerts the most potent effect. Following exposure to high salt, this cell becomes active (Figure 3D), is crucial for the formation of an odor-high salt memory (Figure 6D) and can even establish an aversive memory that can be recalled on a high salt containing test plate (Figure 5D). Additionally, it receives extensive sensory input from nine neurons through 11 distinct interneurons (Figure 7A and D). Therefore, DAN-g1 is the only cell of the DL1 cluster that satisfies all prerequisites to independently encode the high salt teaching signal. In summary, the DL1 cluster appears to be arranged in a reasonable functional configuration. The instructive signal is encoded by a mere four cells, with the requisite of partial redundancy to preclude complete loss of function in the event of single-cell failure. Nevertheless, the cells manifest variation, thereby covering different and cumulative characteristics of stimulation.

However, these cellular-level interpretations are inevitably oversimplified. The MB054B split line occasionally exhibits a weak indication of a third cell (Figure 2J), introducing a degree of variability in the experiments. Similarly, DAN-f1 demonstrates no direct response to salt (Figure 3C), suggesting it is not essential for salt learning, yet it triggers a potent aversive memory (Figure 5C). Nevertheless, it is worth mentioning that the physiological and EM analyses were conducted using first instar larvae, whereas the light microscopy experiments and behavioral tests were performed with third instar larvae. Hence, it is possible that developmental-specific alterations may have occurred, which were not accounted for in this study. Regrettably, our understanding of how salt is perceived in larvae remains limited (Liu et al 2003). Assuming peripheral organ perception as primary, the left and right interneuron 1 (also called FB2IN-12), connected to DAN-d1, DAN-f1, and DAN-g1, is presumed to be centrally important as it encodes for the main peripheral gustatory information (encoded by yellow color, Figure 7C, first and second line). However, these DANs display no response to low salt concentrations and varying responses to high salt concentrations and fructose. In contrast, DAN-c1 is the only cell that responds to high and low concentrations of salt (Figure 3A) but is not connected to interneuron pair 1 or other sensory-to DAN interneurons (except interneuron 20L; Figure 7C). Further experiments are therefore necessary to refine our initial cellular-level description and comprehend the functional spectrum of individual DL1 DANs more comprehensively.

Furthermore, other factors must also be taken into account. Blocking almost all TH-positive cells does not result in the complete disappearance of the salt memory. Instead, it is merely reduced by half (Figure 1H and I). There could be several reasons for this outcome, or a combination thereof. For example, it is plausible that the UAS-GtACR2 effector does not completely suppress the activity of dopaminergic neurons. Additionally, the memory may encompass different types (Widmann et al 2016), not all of which are associated with dopamine function. It is also noteworthy that TH-Gal4 does not encompass all dopaminergic neurons - even a neuron from the DL1 cluster is missing (Selcho et al 2009). Considering that we are using high salt concentrations in this experiment, it is also conceivable that no taste-driven memories are formed solely due to the systemic effects of salt (e.g., increased osmotic pressure). These effects naturally complicate the interpretation of the data at the single-cell level.

The functioning of the DL1 cluster DANs as a multimodal node for punishment

Larvae, of course, perceive not only high salt concentrations as punishing, but also other sensory stimuli such as bitter substances (quinines and caffeine), electric shock, temperature, mechanosensory input (vibration), and light (Aceves-Pina & Quinn 1979, Apostolopoulou et al 2016, Apostolopoulou et al 2014, Eschbach et al 2011, Gerber & Hendel 2006, Khurana et al 2012, Pauls et al 2010, von Essen et al 2011). However, the extent to which the DL1 cluster is involved in coding for most of these aversive teaching signals remains unclear. Currently, comparisons at the level of DANs can only be made for sensory inputs that are dependent on high salt or mechanosensation. Interestingly, similar to high salt concentrations, somatosensory information can also result in the establishment of aversive olfactory memory in larvae (Eschbach et al 2020). The transmission of these somatosensory teaching signals is orchestrated by mechanosensory neurons from the chordotonal organs, class IV multidendritic nociceptive neurons distributed along the body, and multisensory basin neurons (Hwang et al 2007, Jovanic et al 2016, Ohyama et al 2015, Tracey et al 2003). The connectome revealed that these mechanosensory neurons also link to the DL1 cluster through three to four interneurons and elicit the activation of DAN-d1, DAN-f1, and DAN-g1 (Eschbach et al 2020). Upon comparing the interneurons that relay high salt and mechanosensory information to the DL1 cluster, six pairs of interneurons are of critical significance, respectively. These pairs include FFN-20, FB2N-12, FB2N-19, FFN-23, FFN-29, and FB2N-15 for mechanosensory information (Eschbach et al 2020), and FB2N-12, FB2N-18, FB2IN-11, FB2IN-6, FFN-21, and FFN-24 for high salt concentrations (Figure 7; Figure 7- figure supplement 1A, rightmost column). Interestingly, only one pair of interneurons - FB2N12 - is shared between the two types of information and the other five types are specific to the sensory modality. The wiring patterns of DAN input neurons exhibit a notable consistency, with a limited number of neurons dedicated to a specific sensory modality, and single neurons shared across different modalities. This results in a unique combination of cellular codes that are partly redundant and partly specific for the four DL1 DANs. While certain DANs may hold greater significance for a particular information, such as DAN-d1 for somatosensory information, the DAN-g1 cell appears to play a general central role for encoding an aversive teaching signal. DAN-g1 responds to all aversive stimuli tested thus far and its activity has proven informative for all aversive memories analyzed to date. In order to gain a deeper comprehension of the processing of diverse sensory modalities in the larval brain, it is crucial to broaden the analytical framework presented in our work to encompass visual stimuli, various bitter substances, and temperature stimuli.

However, the overall depiction presented above of the information flow and processing of DL1 DANs is an oversimplification. Indeed, DL1 and pPAM DANs are among the most complex and highly recurrent neurons in the brain (Winding et al 2023). This exceptional level of recurrent connectivity enables DANs to provide high-dimensional feedback, which enables them to encode a diverse range of features and engage in parallel computations. Such computations can guide a working memory, contain feedback from neurons that integrate both learned and innate values, and receive long-range feedback from descending neurons that encode motor commands; all in addition to polysynaptic feedforward inputs from the entire set of sensory modalities.

A cellular comparison of DAN function across Drosophila metamorphosis

A recent study conducted by Truman and colleagues explored the integration, or lack thereof, of individual input and output neurons from the larval MB into the adult network following metamorphosis (Truman et al 2023). The study demonstrated that while all four DANs of the pPAM cluster perished, three out of the four cells from the DL1 cluster persisted as part of the adult MB circuitry. Only DAN-f1 altered its innervation in the adult brain, relocating from the larval MB to the adult superior medial protocerebrum. In contrast, DAN-c1, DAN-d1, and DAN-g1 maintained their function as MB DANs in the adult system, identified as PPL1-γ1pedc, PPL1-γ2α’1, and PPL1-γ1, respectively. Due to the lack of appropriate genetic tools, data on the function of PPL1-γ1 is scarce (Aso et al 2014). Therefore, a functional comparison between larval and adult DANs is only possible for the DAN-c1/PPL1-γ1pedc and DAN-d1/PPL1-γ2α’1 combination. Embryonic-born DAN-c1/PPL1-γ1pedc, DAN-d1/PPL1-γ2α’1, and DAN-g1/PPL1-γ1 exhibit sustained activity and respond to various sensory stimuli, including electric shock, temperature, or bitter taste, that can instruct short lasting aversive memories (Aso et al 2012, Aso & Rubin 2016, Aso et al 2010, Claridge-Chang et al 2009, Das et al 2014, Galili et al 2014, Huang et al 2024, Kirkhart & Scott 2015, Tomchik 2013, Villar et al 2022, Vrontou et al 2021). Despite the massive reorganization of the mushroom body, the basic function of DAN-c1/PPL1-γ1pedc and DAN-d1/PPL1-γ2α’1, mediating punishment teaching signals that instruct short-term memories, remains the same. Evidence indicates that the underlying molecular signaling pathways maintain their identity and kinetics throughout development, with larval and adult interstimulus interval curves for DAN-d1/PPL1-γ2α’1 being almost identical (Aso & Rubin 2016, Huang et al 2024, Weiglein et al 2021). During adulthood, PPL1-DANs are responsible for numerous other functions, including innate olfactory responses, suppression of appetite memory in response to nutritional status, pre-exposure learning, memory reconsolidation, and forgetting (Berry et al 2015, Berry et al 2018, Felsenberg et al 2017, Jacob et al 2021, Krashes et al 2009, McCurdy et al 2021, Placais et al 2012, Siju et al 2020, Tian et al 2016, Vrontou et al 2021). Additionally, they are involved in various behaviors like locomotion and sleep (Berry et al 2015).

An open question is whether larval DL1 DANs encode these same functions to a similar extent. Our study establishes a fundamental framework that can be leveraged to investigate this issue. Considering the numerical simplicity of the larval nervous system, the available genetic tools, and the complete connectome, it may be possible to attain an understanding of the larval dopaminergic system at the level of single cells and single synapses. It is plausible that nature has a limited number of effective neural circuit solutions for complex cognitive and behavioral problems, such as the remarkable ability of brains to learn from a small number of examples. Therefore, comprehending the larval dopaminergic system may provide insight into the shared circuit motifs that exist across the animal kingdom.

Materials and Methods

Fly Strains

Flies were raised and maintained on Drosophila standard food at 25°C, 60-80% relative humidity and a 14/7-hour light/dark cycle. For anatomical analysis, UAS-mCD8::GFP (Bloomington stock center no. 32194) (Selcho et al 2009), UAS-mCD8::GFP;mb247-LexA,lexAop-mRFP/TM3,Sb (Burke et al 2012) and UAS-mCD8::GFP;nSyb-LexA,lexAop-mRFP/TM6B (Burke et al 2012, Riabinina et al 2015) were used to analyze the morphology of DANs. For behavioral experiments, UAS-hid,rpr (Abbott & Lengyel 1991, White et al 1996), UAS-GtACR2 (Mohammad et al 2017) (Bloomington stock center no. 92984) and UAS-ChR2XXL (Dawydow et al 2014) (Bloomington stock center no. 58374) were used to ablate or optogenetically silence or activate neurons by blue light (470 nm). UAS-GCaMP6m (Chen et al 2013) (Bloomington stock center no. 42748) flies were used for Ca2+-imaging experiments. The Gal4 strains TH-Gal4 (Friggi-Grelin et al 2003) and R58E02-Gal4 (Rohwedder et al 2016) (Bloomington stock center no. 41347) were used to manipulate different sets of DANs. Split-Gal4 lines MB054B, MB065B, MB328B, MB143B, MB145B, SS02160, SS02180, SS01716, SS01958, SS21716, SS24765, SS01702 (Eschbach et al 2020, Pfeiffer et al 2010, Saumweber et al 2018) were used for anatomical, physiological and behavioral experiments. w1118 flies (Bloomington stock center no. 3605) were used to obtain heterozygous driver and effector control groups. See supplementary table 1 for more information.

Anatomical Analysis

Immunostaining

To confirm expression patterns of R58E02-Gal4, TH-Gal4 and MB054B split-Gal4, flies were crossed to UAS-mCD8::GFP. To assess the functionality of the ablation by effector UAS-hid,rpr, split-Gal4 MB054B was crossed with the latter. Furthermore, MB054B>+ and +>UAS-hid,rpr were anatomically examined for validation. All flies were raised at 25°C for seven days before dissection. Third instar larvae (wandering stage) were put on ice and dissected in PBS (Phosphate Buffered Saline; Sigma Aldrich, cat. no. P4417). Brains were fixed in 4% formaldehyde solution (in PBS, Thermo Scientific, cat. no. 047392.9M) for 20 minutes at room temperature. After eight rinses in PBT (PBS with 3% Triton X-100, Sigma Aldrich, cat. no. X100), brains were blocked with 5% normal goat serum (NGS, Sigma Aldrich, cat. no. G9023) in PBT for 1 hour at room temperature and incubated for 48 hours with primary antibodies at 4°C. Before application of the secondary antibodies for at least 24 hours at 4°C, brains were washed seven times with PBT. After that, larval brains were again washed eight times with PBT and mounted on poly-L-lysin-coated coverslips (Janelia FlyLight DPX mounting receipt: https://www.janelia.org/project-team/flylight/protocols), dehydrated through a series of increasing concentrations of ethanol and cleared three times for 5 minutes in xylene (Sigma Aldrich, cat. no. 247642). At the end, larval brains were mounted in DPX mounting medium (dibutyl phthalate in xylene, Sigma Aldrich, cat. no. 06522) and left to rest in the dark for at least 24 hours before imaging. Primary antibodies were: rat anti-N-Cadherin (1:50; Hybridoma, cat. no. DN-Ex #8-s), rabbit anti-GFP (1:1000; Life Technologies, cat. no. A6455),mouse 4F3 anti-DLG (1:200; Hybridoma, cat. no. 4F3 anti-discs large) and mouse anti-TH (1:50, Immunostar, cat. no. 22941). Secondary antibodies were: goat anti-rat Alexa Fluor 647 (1:500 or 1:250, for anti-N-Cadherin, Life Technologies, cat. no. A21247), goat anti-rabbit Alexa Fluor 488 (1:500, for anti-GFP, Life Technologies, cat. no. A11008), goat anti-mouse Alexa Fluor 568 (1:500, for anti-DLG, Life Technologies, cat. no. A10037) and goat anti-mouse Cy3 (1:200, for anti-TH, life technologies, cat. no. A10521).

Native fluorescence

For native fluorescence, we used a previously described protocol for adult flies (Pitman et al 2011) and adapted it to larvae. All split-Gal4 lines were crossed to UAS-mCD8::GFP;mb247-LexA,lexAop-mRFP/TM3,Sb, while split-Gal4 line MB054B was additionally crossed to UAS-mCD8::GFP;nSyb-LexA,lexAop-mRFP/TM6B and raised at 25°C for seven days. Third instar larvae (wandering stage) were put on ice and dissected in PBS. Brains were fixed under vacuum in 4% formaldehyde solution (in PBS) for 25 minutes at room temperature. After five washes in PBT (PBS with 3% Triton X-100), larval brains were rinsed twice directly and once for 10 minutes under vacuum in PBS at room temperature. Samples were mounted on poly-L-lysin-coated coverslips and embedded in Vectashield (Vector Laboratories, cat. no. H-1000-10). Slides were stored in the dark and cold for at least 24 hours before imaging.

Confocal microscopy

Confocal microscopy was conducted on a Zeiss LSM800 confocal laser scanning microscope with ZEN 2.3 software. All images were projected and adjusted with ImageJ (Fiji is just ImageJ, Version 1.53c, Java 1.8.0_172 (64-bit)). Final Figures were arranged and labeled in Affinity Publisher 2.1.1.

Functional Imaging and microfluidics

For larval imaging experiments, adult flies were transferred to larvae collection cages (Genesee Scientific) containing grape juice agar plates and 180 mg of fresh yeast paste per cage. Flies were allowed to lay eggs on the agar plate for 1–2 days before the plate was removed for collection of larvae. Calcium imaging experiments were performed in L1 larvae (2 days AEL). We used a previously described method for microfluidic delivery of chemicals in aqueous form with simultaneous imaging of calcium activity in intact larvae (Si et al 2019). All experiments used an 8-channel microfluidic chip equipped with a vacuum port to stabilize the animal’s head. The same tastants were used as in the behavioral experiments (NaCl: VWR Chemicals, cat. no. 27810.364; D-Fructose: Sigma Aldrich cat. no. 47740). Stimuli consisted of 5-10 s pulses interleaved with 15 s water washout periods. An L1 larva was washed with deionized water and loaded into the microfluidic device using a 1 mL syringe filled with Triton X-100 (0.1% [v/v]) solution. The animal was pushed to the end of the loading channel with its dorsal side facing the objective. GCaMP6m signal was recorded using an inverted Nikon Ti-E spinning disk confocal microscope and a 60X water immersion objective (NA 1.2). A CCD microscope camera (Andor iXon EMCCD) captured frames at 30 Hz. Recordings from at least 5–7 larvae were collected for each condition. For each larva the average response over 5s around the peak value during stimulus presentation was compared to chance level analyzed via normal t-test. Responses during stimulation were normalized based on the baseline response 10s before the stimulation. P-values for each experiment are given in the respective figure panel. Each graph shows the mean calcium signal plotted as the relative response strength ΔF/F and the related standard error of the mean on the y axis. The time in seconds is given below each graph on the x axis. The grey box indicates the duration of the stimulus application. The sample size of each group (N=5-7) is given above each row. n.s. p > 0.05; * p < 0.05.

Behavioral Experiments

Associative olfactory learning

Standard experiments were done as described before (Apostolopoulou et al 2013, Gerber et al 2013, Hendel et al 2005, Michels et al 2017, Widmann et al 2018). Learning experiments were conducted on assay plates (85 mm diameter, Sarstedt, cat. no. 82.1472) filled with a thin layer of either 2.5% (w/v) pure agarose solution (Sigma Aldrich, cat. no. A9539) or 2.5% (w/v) agarose plus either 1.5 M sodium chloride solution (VWR Chemicals, cat. no. 27810.364), 2 M D-Fructose solution (Sigma Aldrich cat. no. 47740) or 10 mM quinine solution (quinine hemisulfate; Sigma Aldrich cat. no. Q1250). Before closing and labeling the lids, solutions were let to cool down at room temperature to avoid condensation. Plates were stored at 18°C and used within five days. As olfactory stimuli, we used two different odor combinations: either amyl acetate (AM, Sigma Aldrich, cat. no. 46022) diluted 1:250 in paraffin oil (Sigma Aldrich, cat. no. 76235) and undiluted benzaldehyde (BA, Sigma Aldrich, cat. no. 12010) or benzaldehyde and hexyl acetate (HA, Sigma Aldrich, cat. no. 108154) both diluted 1:100 in paraffin oil. For both odor combinations, 10 µL odor were loaded into custom-made Teflon containers (4.5 mm diameter) with perforated lids (Scherer et al 2003).

A minimum of 30 early third instar larvae were collected and exposed to a first odor (amyl acetate or hexyl acetate) while crawling on pure agarose for five minutes, followed by five minutes exposition to a second odor (benzaldehyde) on agarose medium with either 1.5 M sodium chloride, 10 mM quinine as negative reinforcer or 2 M D-Fructose as positive reinforcer (Odor1/Odor2+). For three cycle training, two more repetitions of training trials were performed. A second group of larvae received reciprocal training (Odor1+/Odor2). Please note that occurrence of odor and reinforcer during training have been randomized to exclude sequential or positional effects. After one or three training cycles, larvae were transferred onto test plates containing either pure agarose (reward learning) or agarose plus salt or quinine (punishment learning) on which odor 1 and odor 2 were represented on opposite sides. After five minutes, larvae were counted on each side of the test plate (#Odor1, #Odor2) or in a 10 mm neutral zone in the middle of the plate (#Neutral). A Preference index (PREF) for each group of larvae was calculated as follows: To measure specifically the effect of associative learning, we then calculated the associative performance index (PI) as follows: Negative Performance Indices represent aversive associated learning, whereas positive values represent appetitive learning. Division by 2 ensures scores are bound within (-1; 1).

Odor preference test

Pure agarose (2.5% w/v) plates were cast as previously described (Apostolopoulou et al 2013, Gerber et al 2013, Michels et al 2017, Scherer et al 2003, Widmann et al 2018). Furthermore, the odors were prepared according to the dilutions used in the learning experiments. An odor container (either AM 1:250, BA undiluted, HA 1:100, or BA 1:100) was placed on one side of the agarose plate. An empty container was placed on the opposite side to exclude visual or other side effects. Approximately 30 third instar feeding stage larvae were collected and placed on the plate for 5 min. After this time, larvae were counted, subdivided into larvae on the odor side (#Odor), larvae on the empty container side (#Empty) and larvae in a 10 mm neutral zone in the middle of the plate, as well as larvae that were on the lid of the Petri dish (#Neutral). Preferences were calculated as follows: Positive values indicate a preference for the respective odor, while negative values indicate odor avoidance. The preference tests were performed in different orientations (odor container on plate up/down/left/right) to exclude directional bias.

Gustatory preference test

High salt-dependent choice behavior experiments were performed using standard methods (Hendel et al 2005, Huser et al 2017, Huser et al 2012, Niewalda et al 2008, Selcho et al 2009, Widmann et al 2016). A 2.5% (w/v) agarose solution was boiled as previously described and a thick layer was poured into Petri dishes. After cooling, half of the agarose in the Petri dish was removed and filled by 2.5% (w/v) agarose solution with 1.5 M sodium chloride. For the choice experiment, about 30 third instar feeding stage larvae were put in the middle of the Petri dish and let them crawl for five minutes. Larvae were then counted as being located on the sodium chloride side (#NaCl), pure agarose side (#Agarose), or a neutral area of about 10 mm width in the middle of the plate (#Neutral), as well as larvae that were on the lid of the Petri dish (#Neutral). The gustatory preference indices for sodium chloride were calculated as follows: Negatives values for preference indices indicate an aversion to sodium chloride. The gustatory preference tests were performed in different orientations (NaCl solution on plate up/down/left/right) to exclude directional bias.

Optogenetic substitution experiment

To substitute an actual salt punishment by remotely activating DANs, we used UAS-ChR2XXL. Effector lines were crossed to w1118 to obtain appropriate genetic controls. Larvae were maintained in vials with Drosophila standard food on 25°C and wrapped in aluminum foil to ensure development in constant darkness. A group of 30 feeding-stage third-instar larvae were placed onto plates containing 2.5% agarose and exposed to either AM or BA. During the presentation of the first odor, the larvae were exposed to blue light (470 nm, 220 lux) for 5 minutes. The second odor BA (undiluted) was presented in darkness for five minutes on a pure agarose plate. As described for odor-high salt learning, training was performed reciprocally and the sequence of training trials was alternated across repetitions of the experiment. Furthermore, a control group, identical to the experimental group, underwent no blue light stimulation during the learning paradigm. Data were then scored as above. Please note, tests were done in the presence of 1.5 M sodium chloride to test for the specific quality of the memory. The activation of blue light itself does not cause any disruptive behavioral side effects, as it does not induce an appetitive or aversive memory on its own.

Optogenetic inhibition of neuronal activity

To acutely block synaptic output, we used UAS-GtACR2. Flies were maintained on standard food supplemented with 0.5 mM all-trans-retinal (ATR, Sigma Aldrich, cat. no. R2500) at 25°C as described before (Meloni et al 2020). Vials were wrapped in aluminum foil to ensure larval development in constant darkness. Groups of about 30 third instar feeding stage larvae received a reciprocal two-odor training as described before, exposed to blue light (470 nm, 1100 lux) during the entire training phase. The sequence of training trials was alternated across repetitions of the experiment. The test was performed for 5 minutes in darkness on an agarose plate mixed with 1.5 M sodium chloride. Control experiments were performed with the same genotype but with standard food lacking 0.5 mM all-trans-retinal. Please note that the blue light exposure itself does not cause any disruptive behavioral side effects, as it does not impair odor-high salt learning of control animals. Data were then calculated and scored as mentioned above.

Evaluation of DAN input wiring diagrams

Pie charts of sensory profiles were calculated using the percentage of total synaptic input of interneurons as fraction (thereby ignoring other inputs to show distribution of sensory origins). Percentages then give the percentage of total sensory synaptic input to interneurons. The calculation of the hub score was done in the following way: Fraction of total synaptic input from all sensory neurons to defined interneurons (see IDs) was multiplied by the total fraction of input of the DANs from this interneuron. For anatomical reconstructions and visualizations included here, we made use of earlier published data (Eichler et al 2017, Eschbach et al 2020, Miroschnikow et al 2018, Winding et al 2023).

Statistical Analysis

Behavioral results were analyzed in GraphPad Prism 8.4.3. To test for normal distribution, Shapiro-Wilk test was applied. Groups that did not violate the assumption of normal distribution, were analyzed with an unpaired t-test (comparison of two groups) or a one-way ANOVA followed by a Tukey’s post hoc test (comparisons between groups larger than two). For nonparametric statistics, Mann-Whitney test (comparison between two groups) or Kruskal-Wallis followed by Dunn’s multiple comparisons tests were applied. Furthermore, one sample t-test for parametric or Wilcoxon signed-rank test for nonparametric data sets were performed to compare means or medians against chance level. Results are visualized in box-plots, indicating the median as middle line, 25%/75% quantiles as box boundaries and minimum/maximum performance indices as whiskers. Each data point is represented as black dot and sample sizes are noted within graphics. Asterisks and ‘‘n.s.’’ indicate p < 0.05 and p > 0.05, respectively. The source data and results of all statistical tests are documented in the Supplemental Information file.